---
title: "Step 1: Occurrence Data Preparation"
subtitle: "*Dermacentor variabilis* Species Distribution Model"
author: "Alexander W. Gofton"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme:
      light: flatly
      dark: solar
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    results: hold
    cache.lazy: false
---

## Overview

This script prepares occurrence data for *Dermacentor variabilis* (American dog tick)
from a GBIF download (22,668 records). Key steps:

1. Load and filter raw GBIF occurrence records
2. Taxonomic filtering: restrict to eastern N. America (true *D. variabilis* sensu stricto,
   excluding western populations now classified as *D. similis* per Lado et al. 2021)
3. Coordinate cleaning using `CoordinateCleaner`
4. Spatial thinning at 10 km to reduce sampling bias
5. Map and summarise final occurrence dataset

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(CoordinateCleaner)
library(spThin)
```

## 1. Project paths

```{r}
#| label: paths

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/SDM/Dvariabilis_SDM"
data_dir <- file.path(base_dir, "data")
processed_dir <- file.path(base_dir, "processed_data")
figures_dir <- file.path(base_dir, "figures")

# Create output dirs if needed
dir.create(processed_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)
```

## 2. Load raw GBIF occurrence data

The GBIF download contains 22,668 records for *Dermacentor variabilis*.

```{r}
#| label: load-gbif

occ_file <- file.path(data_dir, "Dvar_occurences", "occurrence.txt")

# Read tab-delimited GBIF occurrence file — select only columns we need
occ_raw <- read_tsv(
  occ_file,
  col_select = c(
    gbifID, species, basisOfRecord,
    decimalLatitude, decimalLongitude,
    coordinateUncertaintyInMeters,
    countryCode, stateProvince,
    year, month, day
  ),
  show_col_types = FALSE
)

sprintf("Loaded %d raw GBIF records", nrow(occ_raw))
```

## 3. Initial filtering

Remove records that lack coordinates, have zero coordinates, or are from non-observation
basis of record types (e.g., fossils, living specimens in zoos).

```{r}
#| label: initial-filter

occ_filtered <- occ_raw %>%
  # Remove missing coordinates

  filter(!is.na(decimalLatitude), !is.na(decimalLongitude)) %>%
  # Remove zero coordinates (common GBIF artifact)
  filter(!(decimalLatitude == 0 & decimalLongitude == 0)) %>%
  # Keep only observation-based records
  filter(basisOfRecord %in% c(
    "HUMAN_OBSERVATION", "OBSERVATION",
    "MACHINE_OBSERVATION", "PRESERVED_SPECIMEN"
  ))

sprintf(
  "After initial filtering: %d -> %d records (removed %d)",
  nrow(occ_raw), nrow(occ_filtered), nrow(occ_raw) - nrow(occ_filtered)
)
```

## 4. Taxonomic/geographic filtering — restrict to eastern N. America

Lado et al. (2021) split western N. American populations of *D. variabilis* into
*D. similis*. We restrict to records **east of ~105 W longitude** to retain only
true *D. variabilis* sensu stricto. We also restrict to N. American country codes
(US, CA, MX) since that is the native range.

```{r}
#| label: taxonomic-filter

# Native range country codes
native_countries <- c("US", "CA", "MX")

occ_native <- occ_filtered %>%
  # Restrict to North America
  filter(countryCode %in% native_countries) %>%
  # Restrict to eastern populations (east of Rocky Mountains)
  filter(decimalLongitude >= -105)

sprintf(
  "After restricting to eastern N. America: %d -> %d records",
  nrow(occ_filtered), nrow(occ_native)
)

# Geographic summary
occ_native %>%
  count(countryCode, name = "n_records") %>%
  mutate(pct = round(100 * n_records / sum(n_records), 1)) %>%
  arrange(desc(n_records))
```

## 5. Coordinate cleaning with CoordinateCleaner

Flag and remove records with common geospatial errors: country centroids,
capital coordinates, institution locations, duplicates, etc.

```{r}
#| label: coord-clean

# Rename columns for CoordinateCleaner (expects specific names)
occ_for_cc <- occ_native %>%
  rename(
    decimallongitude = decimalLongitude,
    decimallatitude = decimalLatitude
  )

# Run coordinate cleaning tests
cc_flags <- clean_coordinates(
  occ_for_cc,
  lon = "decimallongitude",
  lat = "decimallatitude",
  countries = "countryCode",
  species = "species",
  tests = c(
    "capitals",      # Within 10 km of country capitals
    "centroids",     # Within 5 km of country/province centroids
    "duplicates",    # Exact coordinate duplicates
    "equal",         # Identical lat/lon (diagonal line artifacts)
    "gbif",          # GBIF headquarters
    "institutions",  # Known biodiversity institutions
    "seas",          # In the ocean
    "zeros"          # Zero coordinates
  ),
  verbose = TRUE
)

# Summary of flagged records
summary(cc_flags)

# Keep only clean records
occ_clean <- occ_for_cc[cc_flags$.summary, ] %>%
  rename(
    decimalLongitude = decimallongitude,
    decimalLatitude = decimallatitude
  )

sprintf(
  "After coordinate cleaning: %d -> %d records (flagged %d)",
  nrow(occ_native), nrow(occ_clean),
  nrow(occ_native) - nrow(occ_clean)
)
```

## 6. Remove exact coordinate duplicates

Even after CoordinateCleaner, ensure no exact duplicates remain.

```{r}
#| label: dedup

n_before <- nrow(occ_clean)
occ_dedup <- occ_clean %>%
  distinct(decimalLongitude, decimalLatitude, .keep_all = TRUE)

sprintf(
  "After removing exact duplicates: %d -> %d records",
  n_before, nrow(occ_dedup)
)
```

## 7. Spatial thinning at 10 km

Reduce sampling bias by ensuring a minimum distance of 10 km between records.
This is approximately 2x the ~5 km cell size of the 2.5 arcmin climate data.

```{r}
#| label: spatial-thin

# Prepare data for spThin
thin_input <- occ_dedup %>%
  select(species, decimalLongitude, decimalLatitude)

# Run spatial thinning (10 km minimum distance, 100 repetitions)
set.seed(42)
thin_result <- thin(
  loc.data = thin_input,
  lat.col = "decimalLatitude",
  long.col = "decimalLongitude",
  spec.col = "species",
  thin.par = 10, # 10 km minimum distance
  reps = 100,
  locs.thinned.list.return = TRUE,
  write.files = FALSE,
  write.log.file = FALSE
)

# Select the replicate with the most records retained
n_retained <- sapply(thin_result, nrow)
best_rep <- which.max(n_retained)
thin_coords <- thin_result[[best_rep]]

sprintf(
  "Spatial thinning at 10 km: %d -> %d records (best of %d replicates)",
  nrow(occ_dedup),
  nrow(thin_coords),
  length(thin_result)
)
```

## 8. Rejoin thinned coordinates with metadata

```{r}
#| label: rejoin

# Round coordinates for matching (spThin may introduce tiny float differences)
occ_thinned <- occ_dedup %>%
  mutate(
    lon_round = round(decimalLongitude, 5),
    lat_round = round(decimalLatitude, 5)
  ) %>%
  inner_join(
    thin_coords %>%
      rename(decimalLongitude = Longitude, decimalLatitude = Latitude) %>%
      mutate(
        lon_round = round(decimalLongitude, 5),
        lat_round = round(decimalLatitude, 5)
      ) %>%
      select(lon_round, lat_round),
    by = c("lon_round", "lat_round")
  ) %>%
  select(-lon_round, -lat_round)

sprintf("Final thinned dataset: %d records", nrow(occ_thinned))
```

## 9. Summary statistics

```{r}
#| label: summary-stats

cat("--- Final Occurrence Dataset Summary ---\n")
sprintf("Total records: %d", nrow(occ_thinned))
sprintf("Country codes: %s", paste(unique(occ_thinned$countryCode), collapse = ", "))
sprintf(
  "Longitude range: %.2f to %.2f",
  min(occ_thinned$decimalLongitude), max(occ_thinned$decimalLongitude)
)
sprintf(
  "Latitude range: %.2f to %.2f",
  min(occ_thinned$decimalLatitude), max(occ_thinned$decimalLatitude)
)
sprintf(
  "Year range: %s to %s",
  min(occ_thinned$year, na.rm = TRUE), max(occ_thinned$year, na.rm = TRUE)
)

# Basis of record breakdown
occ_thinned %>%
  count(basisOfRecord, name = "n") %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  arrange(desc(n))
```

## 10. Visualisation — occurrence map

```{r}
#| label: occ-map
#| fig-width: 10
#| fig-height: 7

# Get North America boundaries
world <- ne_countries(scale = "medium", returnclass = "sf")
na_bbox <- c(xmin = -105, xmax = -60, ymin = 20, ymax = 55)

# Convert to sf
occ_sf <- st_as_sf(
  occ_thinned,
  coords = c("decimalLongitude", "decimalLatitude"),
  crs = 4326
)

ggplot() +
  geom_sf(data = world, fill = "grey90", colour = "grey60", linewidth = 0.2) +
  geom_sf(data = occ_sf, colour = "#D55E00", size = 0.8, alpha = 0.6) +
  coord_sf(
    xlim = c(na_bbox["xmin"], na_bbox["xmax"]),
    ylim = c(na_bbox["ymin"], na_bbox["ymax"])
  ) +
  labs(
    title = expression(italic("Dermacentor variabilis") ~ "— Thinned GBIF Occurrences"),
    subtitle = sprintf(
      "n = %d records (eastern N. America, 10 km thinned)",
      nrow(occ_thinned)
    ),
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_line(colour = "grey95"),
    plot.title = element_text(size = 14, face = "bold")
  )

ggsave(
  file.path(figures_dir, "01_occurrence_map.png"),
  width = 10, height = 7, dpi = 300
)
```

## 11. Data cleaning pipeline summary

```{r}
#| label: pipeline-summary

pipeline <- tibble(
  Step = c(
    "Raw GBIF download",
    "After initial filtering (coords, basis of record)",
    "After restricting to eastern N. America",
    "After CoordinateCleaner",
    "After removing exact duplicates",
    "After spatial thinning (10 km)"
  ),
  Records = c(
    nrow(occ_raw),
    nrow(occ_filtered),
    nrow(occ_native),
    nrow(occ_clean),
    nrow(occ_dedup),
    nrow(occ_thinned)
  )
) %>%
  mutate(
    Removed = lag(Records, default = first(Records)) - Records,
    Pct_retained = round(100 * Records / first(Records), 1)
  )

pipeline
```

## 12. Save outputs

```{r}
#| label: save

# Save thinned occurrences as CSV and RDS
write_csv(
  occ_thinned,
  file.path(processed_dir, "occurrences_thinned.csv")
)
saveRDS(
  occ_thinned,
  file.path(processed_dir, "occurrences_thinned.rds")
)

# Save as sf object too
saveRDS(
  occ_sf,
  file.path(processed_dir, "occurrences_thinned_sf.rds")
)

sprintf("Saved %d thinned occurrence records to processed_data/", nrow(occ_thinned))
```
