---
title: "Step 2: Background Points from Ixodidae Records"
subtitle: "*Dermacentor variabilis* Species Distribution Model"
author: "Alexander W. Gofton"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme:
      light: flatly
      dark: solar
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    results: hold
    cache.lazy: false
---

## Overview

This script generates target-group background (pseudo-absence) points using
**all Ixodidae GBIF records that are NOT *D. variabilis***, clipped to the native
range accessible area (M) of *D. variabilis*.

Target-group backgrounds correct for spatial sampling bias — the same collector
effort patterns that produced *D. variabilis* records also apply to other tick species,
so using Ixodidae records as background controls for this bias.

Key steps:

1. Load thinned *D. variabilis* occurrences from Step 1
2. Load Ixodidae GBIF records (~547k records)
3. Define accessible area "M" as a 500 km buffered convex hull around occurrences
4. Filter to records identified at minimum to genus level (remove unidentified bulk records)
5. Clip Ixodidae records to M, excluding *D. variabilis*
6. Subsample to 10,000 background points

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
```

## 1. Project paths

```{r}
#| label: paths

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/SDM/Dvariabilis_SDM"
data_dir <- file.path(base_dir, "data")
processed_dir <- file.path(base_dir, "processed_data")
figures_dir <- file.path(base_dir, "figures")
```

## 2. Load thinned D. variabilis occurrences

```{r}
#| label: load-occurrences

occ <- readRDS(file.path(processed_dir, "occurrences_thinned.rds"))
sprintf("Loaded %d thinned D. variabilis occurrences", nrow(occ))

# Convert to sf
occ_sf <- st_as_sf(
  occ,
  coords = c("decimalLongitude", "decimalLatitude"),
  crs = 4326
)
```

## 3. Define accessible area M — 500 km buffered convex hull

The accessible area represents the geographic region *D. variabilis* has had the
opportunity to disperse to and sample from. We use a 500 km buffer around the
convex hull of occurrence points, following Alkishe et al. (2020)'s buffer approach.

```{r}
#| label: define-M

# Create convex hull around occurrences
occ_hull <- st_convex_hull(st_union(occ_sf))

# Buffer by 500 km (need to transform to a projected CRS for distance calculations)
# Use an equal-area projection centred on N. America
occ_hull_proj <- st_transform(occ_hull, crs = "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-85 +datum=WGS84")
M_proj <- st_buffer(occ_hull_proj, dist = 500000)  # 500 km in metres
M <- st_transform(M_proj, crs = 4326)

# Clip M to land (remove ocean areas)
world <- ne_countries(scale = "medium", returnclass = "sf") %>%
  st_union()
M_land <- st_intersection(M, world)

sprintf("Accessible area M created: 500 km buffered convex hull, clipped to land")

# Save M for later use
saveRDS(M_land, file.path(processed_dir, "accessible_area_M.rds"))
```

## 4. Load Ixodidae GBIF records

This dataset contains ~547k records of all Ixodidae ticks from GBIF.

```{r}
#| label: load-ixodidae

ixod_file <- file.path(data_dir, "Ixodidae_background_points", "0023290-260208012135463.csv")

ixod_raw <- read_tsv(
  ixod_file,
  col_select = c(
    gbifID, species, family, genus,
    decimalLatitude, decimalLongitude,
    countryCode, basisOfRecord, year
  ),
  show_col_types = FALSE
)

sprintf("Loaded %d Ixodidae records", nrow(ixod_raw))
```

```{r}
#| label: plot-ixodiae-records-global
#| fig-width: 12
#| fig-height: 6

world_sf <- ne_countries(scale = "medium", returnclass = "sf")

ixod_coords <- ixod_raw %>%
  filter(!is.na(decimalLongitude), !is.na(decimalLatitude))

ixod_map <- ggplot() +
  geom_sf(data = world_sf, fill = "grey90", colour = "grey70", linewidth = 0.2) +
  geom_point(data = ixod_coords,
             aes(x = decimalLongitude, y = decimalLatitude),
             colour = "#0072B2", size = 0.3, alpha = 0.2) +
  coord_sf(expand = FALSE) +
  labs(
    title = "Global Ixodidae records (GBIF)",
    subtitle = sprintf("%s records with coordinates", format(nrow(ixod_coords), big.mark = ",")),
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_line(colour = "grey80", linewidth = 0.2))

ixod_map
```

## 5. Filter Ixodidae records

Remove *D. variabilis* records, records without coordinates, and restrict to
observation-based record types.

**Critical:** We also remove records not identified to at least genus level.
Many GBIF Ixodidae records (~80%) lack genus-level identification — these are
typically bulk-digitised museum entries or low-quality citizen science records
that do not represent confirmed tick observations. Retaining them would flood
the background with spatially biased, taxonomically uninformative data that
does not correct for sampling effort as target-group backgrounds are designed
to do.

```{r}
#| label: filter-ixodidae

ixod_filtered <- ixod_raw %>%
  # Remove records without coordinates
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude)) %>%
  # Remove zero coordinates
  filter(!(decimalLatitude == 0 & decimalLongitude == 0)) %>%
  # CRITICAL: Only keep records identified to at least genus level

  # Unidentified "Ixodidae" records do not represent genuine tick collection
  # events and cannot correct for sampling bias
  filter(!is.na(genus), genus != "") %>%
  # Remove D. variabilis records (the target species)
  filter(species != "Dermacentor variabilis" | is.na(species)) %>%
  # Also remove D. similis (the recently split western species)
  filter(species != "Dermacentor similis" | is.na(species)) %>%
  # Keep observation-based records
  filter(basisOfRecord %in% c(
    "HUMAN_OBSERVATION", "OBSERVATION",
    "MACHINE_OBSERVATION", "PRESERVED_SPECIMEN"
  ))

sprintf(
  "After filtering Ixodidae: %d -> %d records (removed %d without genus ID)",
  nrow(ixod_raw), nrow(ixod_filtered),
  nrow(ixod_raw) - nrow(ixod_filtered)
)

# Top genera in filtered dataset
cat("\nTop genera in filtered dataset:\n")
ixod_filtered %>%
  count(genus, sort = TRUE) %>%
  head(15)
```

```{r}
ixod_filt_coords <- ixod_filtered %>%
  filter(!is.na(decimalLongitude), !is.na(decimalLatitude))

ixod_filt_map <- ggplot() +
  geom_sf(data = world_sf, fill = "grey90", colour = "grey70", linewidth = 0.2) +
  geom_point(data = ixod_filt_coords,
             aes(x = decimalLongitude, y = decimalLatitude),
             colour = "#0072B2", size = 0.3, alpha = 0.2) +
  coord_sf(expand = FALSE) +
  labs(
    title = "Global Ixodidae records (GBIF)",
    subtitle = sprintf("%s records with coordinates", format(nrow(ixod_coords), big.mark = ",")),
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_line(colour = "grey80", linewidth = 0.2))

ixod_filt_map
```

## 6. Clip Ixodidae records to accessible area M

```{r}
#| label: clip-to-M

# Convert to sf
ixod_sf <- st_as_sf(
  ixod_filtered,
  coords = c("decimalLongitude", "decimalLatitude"),
  crs = 4326
)

# Spatial intersection with M
ixod_in_M <- st_intersection(ixod_sf, M_land)

sprintf(
  "Ixodidae records within accessible area M: %d (from %d total)",
  nrow(ixod_in_M), nrow(ixod_filtered)
)

# Extract coordinates back
coords_M <- st_coordinates(ixod_in_M)
ixod_in_M_df <- ixod_in_M %>%
  st_drop_geometry() %>%
  mutate(
    decimalLongitude = coords_M[, 1],
    decimalLatitude = coords_M[, 2]
  )

# Summary by genus
ixod_in_M_df %>%
  count(genus, sort = TRUE) %>%
  head(10)
```

```{r}
ixod_filt_clip_coords <- ixod_in_M_df %>%
  filter(!is.na(decimalLongitude), !is.na(decimalLatitude))

ixod_filt_clip_map <- ggplot() +
  geom_sf(data = world_sf, fill = "grey90", colour = "grey70", linewidth = 0.2) +
  geom_point(data = ixod_filt_clip_coords,
             aes(x = decimalLongitude, y = decimalLatitude),
             colour = "#0072B2", size = 0.3, alpha = 0.2) +
  coord_sf(expand = FALSE) +
  labs(
    title = "Global Ixodidae records (GBIF)",
    subtitle = sprintf("%s records with coordinates", format(nrow(ixod_coords), big.mark = ",")),
    x = NULL, y = NULL
  ) +
    coord_sf(
    xlim = c(-120, -60),
    ylim = c(60, 10)
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_line(colour = "grey80", linewidth = 0.2))

ixod_filt_clip_map
```

## 7. Subsample background points

Sample up to 10,000 background points from the genus-identified Ixodidae records
within M. With the genus-level filter, the pool is smaller but represents genuine
tick collection events. 10,000 points is sufficient for both MaxEnt and BRT.

```{r}
#| label: subsample

n_target <- 10000

set.seed(42)

if (nrow(ixod_in_M_df) >= n_target) {
  bg_points <- ixod_in_M_df %>%
    slice_sample(n = n_target)
  sprintf("Subsampled %d background points from %d available", n_target, nrow(ixod_in_M_df))
} else {
  bg_points <- ixod_in_M_df
  sprintf(
    "Using all %d available Ixodidae records (fewer than %d target)",
    nrow(bg_points), n_target
  )
}

# Create clean background coordinates dataframe
bg_coords <- bg_points %>%
  select(decimalLongitude, decimalLatitude) %>%
  rename(lon = decimalLongitude, lat = decimalLatitude)

sprintf("Final background dataset: %d points", nrow(bg_coords))
```

## 8. Visualisation — occurrence and background points

```{r}
#| label: map-bg
#| fig-width: 10
#| fig-height: 7

# Convert M to plottable format
M_plot <- st_as_sf(M_land)
world_map <- ne_countries(scale = "medium", returnclass = "sf")

# Background points as sf
bg_sf <- st_as_sf(bg_coords, coords = c("lon", "lat"), crs = 4326)

# Plot extent — slightly larger than M
m_bbox <- st_bbox(M_land)

ggplot() +
  geom_sf(data = world_map, fill = "grey90", colour = "grey60", linewidth = 0.2) +
  geom_sf(data = M_plot, fill = NA, colour = "orange", linewidth = 0.8, linetype = "solid") +
  geom_sf(data = bg_sf, colour = "blue", size = 1.0, alpha = 0.5) +
  geom_sf(data = occ_sf, colour = "red", size = 1.0, alpha = 0.5) +
  coord_sf(
    xlim = c(m_bbox["xmin"] - 2, m_bbox["xmax"] + 2),
    ylim = c(m_bbox["ymin"] - 2, m_bbox["ymax"] + 2)
  ) +
  labs(
    title = expression(italic("D. variabilis") ~ "Occurrences and Ixodidae Background Points"),
    subtitle = sprintf(
      "Red: %d occurrences | Blue: %d Ixodidae background | Orange dashed: Accessible area M",
      nrow(occ), nrow(bg_coords)
    ),
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_line(colour = "grey95"),
    plot.title = element_text(size = 14, face = "bold")
  )

ggsave(
  file.path(figures_dir, "02_background_points_map.png"),
  width = 10, height = 7, dpi = 300
)
```

## Second approach to generating background points
Generate 10,000 random points within the 500 km buffered convex hull (M) around known occurrence points.

```{r}
#| label: random-bg
#| fig-width: 10
#| fig-height: 7

set.seed(42)
n_random <- 10000

# Sample random points within M_land
random_pts_sf <- st_sample(M_land, size = n_random, type = "random")
random_pts_sf <- st_as_sf(random_pts_sf)

# Extract coordinates
random_bg_coords <- data.frame(
  lon = st_coordinates(random_pts_sf)[, 1],
  lat = st_coordinates(random_pts_sf)[, 2]
)

sprintf("Generated %d random background points within accessible area M", nrow(random_bg_coords))

# Map
random_bg_sf <- st_as_sf(random_bg_coords, coords = c("lon", "lat"), crs = 4326)

ggplot() +
  geom_sf(data = world_map, fill = "grey90", colour = "grey60", linewidth = 0.2) +
  geom_sf(data = M_plot, fill = NA, colour = "orange", linewidth = 0.8) +
  geom_sf(data = random_bg_sf, colour = "steelblue", size = 0.3, alpha = 0.4) +
  geom_sf(data = occ_sf, colour = "red", size = 1.0, alpha = 0.7) +
  coord_sf(
    xlim = c(m_bbox["xmin"] - 2, m_bbox["xmax"] + 2),
    ylim = c(m_bbox["ymin"] - 2, m_bbox["ymax"] + 2)
  ) +
  labs(
    title = expression(italic("D. variabilis") ~ "Occurrences and Random Background Points"),
    subtitle = sprintf(
      "Red: %d occurrences | Blue: %d random background within M",
      nrow(occ), nrow(random_bg_coords)
    ),
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_line(colour = "grey95"))

ggsave(
  file.path(figures_dir, "02_random_background_points_map.png"),
  width = 10, height = 7, dpi = 300
)
```

## 9. Save outputs

```{r}
#| label: save

# Save background coordinates
write_csv(bg_coords, file.path(processed_dir, "background_coords.csv"))
saveRDS(bg_coords, file.path(processed_dir, "background_coords.rds"))

# Save full background data with metadata
saveRDS(bg_points, file.path(processed_dir, "background_points_full.rds"))

# Save accessible area M
st_write(
  st_as_sf(M_land),
  file.path(processed_dir, "accessible_area_M.gpkg"),
  delete_dsn = TRUE,
  quiet = TRUE
)

# Save random background coordinates
write_csv(random_bg_coords, file.path(processed_dir, "background_coords_random.csv"))
saveRDS(random_bg_coords, file.path(processed_dir, "background_coords_random.rds"))

sprintf("Saved %d background points and accessible area M to processed_data/", nrow(bg_coords))
```
