---
title: "Step 4b: Variable Selection Sensitivity Analysis"
subtitle: "*Dermacentor variabilis* Species Distribution Model"
author: "Alexander W. Gofton"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme:
      light: flatly
      dark: solar
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    results: hold
    cache.lazy: false
---

## Overview

The primary analysis forced retention of Bio1 (annual mean temperature), which
resulted in VIF values up to 47.25 and a 66.4% MaxEnt contribution that masked
the importance of other variables. Bio1 encodes cold-winter gradients in
eastern North America that are ecologically irrelevant to Australian projection.

This script defines **three alternative variable sets** for sensitivity testing:

| Set | Variables | Rationale |
|-----|-----------|-----------|
| A (current) | bio1, bio5, bio7, bio10, bio12, bio15, aridity_idx, PET_seasonality | Baseline with Bio1 forced |
| B (no Bio1) | bio5, bio7, bio10, bio12, bio15, aridity_idx, PET_seasonality | Remove dominant collinear variable |
| C (moisture-focused) | bio5, bio10, bio12, bio15, aridity_idx, PET_seasonality, vpd_annual | Replace Bio1 + Bio7 with VPD |

Set C aligns with the CLIMEX finding (Halliday & Sutherst 1990) that **low humidity
is the primary limiting factor** for *D. variabilis* in Australia.

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(usdm)
```

## 1. Load data

```{r}
#| label: load

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/SDM/Dvariabilis_SDM"
processed_dir <- file.path(base_dir, "processed_data")

# Full environmental stacks (all 21 candidate variables)
env_stack_M <- rast(file.path(processed_dir, "env_stack_M.tif"))
env_stack_aus <- rast(file.path(processed_dir, "env_stack_aus.tif"))

# Occurrence and background data (contain all 21 variables)
occ_env <- readRDS(file.path(processed_dir, "occurrences_with_env.rds"))
bg_env <- readRDS(file.path(processed_dir, "background_with_env.rds"))

# Current selected variables (Set A)
set_A <- readLines(file.path(processed_dir, "selected_variables.txt"))

cat("Available variables in full stack:", paste(names(env_stack_M), collapse = ", "), "\n")
cat("Current Set A:", paste(set_A, collapse = ", "), "\n")
```

## 2. Define alternative variable sets

```{r}
#| label: define-sets

# Set B: Remove Bio1 (dominant collinear variable)
set_B <- setdiff(set_A, "bio1")

# Set C: Moisture-focused — remove Bio1 and Bio7, add VPD
# Bio7 (temp annual range) encodes continentality/winter severity
# VPD (vapor pressure deficit) directly measures desiccation risk
set_C <- c("bio5", "bio10", "bio12", "bio15",
           "aridity_idx", "PET_seasonality", "vpd_annual")

# Verify all Set C variables exist in the data
missing_C <- setdiff(set_C, names(env_stack_M))
if (length(missing_C) > 0) {
  cat("WARNING: Variables missing from env stack:", paste(missing_C, collapse = ", "), "\n")
  cat("Available variables:", paste(names(env_stack_M), collapse = ", "), "\n")
  set_C <- intersect(set_C, names(env_stack_M))
  cat("Adjusted Set C:", paste(set_C, collapse = ", "), "\n")
} else {
  cat("All Set C variables available\n")
}

cat("\n--- Variable Sets ---\n")
cat("Set A (current):", paste(set_A, collapse = ", "), sprintf("(%d vars)\n", length(set_A)))
cat("Set B (no Bio1):", paste(set_B, collapse = ", "), sprintf("(%d vars)\n", length(set_B)))
cat("Set C (moisture):", paste(set_C, collapse = ", "), sprintf("(%d vars)\n", length(set_C)))
```

## 3. VIF analysis for each set

```{r}
#| label: vif

# Function to compute VIF for a variable set
compute_vif <- function(data, vars, set_name) {
  cat(sprintf("\n--- VIF for %s ---\n", set_name))
  env_cols <- data %>%
    dplyr::select(all_of(vars)) %>%
    as.data.frame() %>%
    na.omit()

  vif_vals <- vif(env_cols)
  print(vif_vals)

  max_vif <- max(vif_vals$VIF)
  cat(sprintf("Maximum VIF: %.2f %s\n",
              max_vif,
              ifelse(max_vif > 10, "*** EXCEEDS THRESHOLD ***", "(OK)")))

  return(vif_vals)
}

vif_A <- compute_vif(occ_env, set_A, "Set A (current)")
vif_B <- compute_vif(occ_env, set_B, "Set B (no Bio1)")
vif_C <- compute_vif(occ_env, set_C, "Set C (moisture)")
```

## 4. Pairwise correlations for each set

```{r}
#| label: correlations

check_correlations <- function(data, vars, set_name) {
  env_cols <- data %>%
    dplyr::select(all_of(vars)) %>%
    as.data.frame() %>%
    na.omit()

  cor_mat <- cor(env_cols)
  high_cor <- which(abs(cor_mat) > 0.7 & upper.tri(cor_mat), arr.ind = TRUE)

  cat(sprintf("\n--- High correlations (|r| > 0.7) for %s ---\n", set_name))
  if (nrow(high_cor) > 0) {
    for (i in seq_len(nrow(high_cor))) {
      v1 <- colnames(cor_mat)[high_cor[i, 1]]
      v2 <- colnames(cor_mat)[high_cor[i, 2]]
      r <- cor_mat[high_cor[i, 1], high_cor[i, 2]]
      cat(sprintf("  %s ~ %s: r = %.3f\n", v1, v2, r))
    }
  } else {
    cat("  None — all pairwise |r| < 0.7\n")
  }
}

check_correlations(occ_env, set_A, "Set A")
check_correlations(occ_env, set_B, "Set B")
check_correlations(occ_env, set_C, "Set C")
```

## 5. Prepare environmental stacks for each set

```{r}
#| label: prepare-stacks

# Create subsetted stacks for Set B and Set C
# (Set A already exists as env_selected_M.tif / env_selected_aus.tif)

# Set B
env_B_M <- env_stack_M[[set_B]]
env_B_aus <- env_stack_aus[[set_B]]
writeRaster(env_B_M, file.path(processed_dir, "env_setB_M.tif"), overwrite = TRUE)
writeRaster(env_B_aus, file.path(processed_dir, "env_setB_aus.tif"), overwrite = TRUE)
cat("Set B stacks saved\n")

# Set C
env_C_M <- env_stack_M[[set_C]]
env_C_aus <- env_stack_aus[[set_C]]
writeRaster(env_C_M, file.path(processed_dir, "env_setC_M.tif"), overwrite = TRUE)
writeRaster(env_C_aus, file.path(processed_dir, "env_setC_aus.tif"), overwrite = TRUE)
cat("Set C stacks saved\n")

# Save variable lists
writeLines(set_B, file.path(processed_dir, "selected_variables_setB.txt"))
writeLines(set_C, file.path(processed_dir, "selected_variables_setC.txt"))
cat("Variable lists saved\n")
```

## 6. Summary comparison

```{r}
#| label: summary

cat("======================================================\n")
cat("  VARIABLE SET COMPARISON\n")
cat("======================================================\n\n")

# VIF comparison table
vif_compare <- tibble(
  Variable = set_A,
  VIF_SetA = vif_A$VIF[match(set_A, vif_A$Variables)]
)

vif_compare <- vif_compare %>%
  left_join(tibble(Variable = set_B,
                   VIF_SetB = vif_B$VIF[match(set_B, vif_B$Variables)]),
            by = "Variable") %>%
  left_join(tibble(Variable = set_C,
                   VIF_SetC = vif_C$VIF[match(set_C, vif_C$Variables)]),
            by = "Variable")

cat("VIF comparison across variable sets:\n")
print(vif_compare, n = Inf)

cat("\nKey improvements in Set B and C:\n")
cat("  - Bio1 removed: eliminates VIF > 40 and 66.4% contribution artefact\n")
cat("  - Set C adds VPD: directly measures desiccation risk (primary CLIMEX limiting factor)\n")
cat("  - Set C removes Bio7: eliminates continentality/winter severity encoding\n")
cat("\nExpected outcomes:\n")
cat("  - MaxEnt-BRT prediction correlation should increase from 0.384 to >0.6\n")
cat("  - Australian suitable area estimates should converge between models\n")
cat("  - MESS novelty should decrease, especially in eastern coastal regions\n")
```
