---
title: "Step 5b: Boosted Regression Trees (BRT) Model"
subtitle: "*Dermacentor variabilis* Species Distribution Model"
author: "Alexander W. Gofton"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme:
      light: flatly
      dark: solar
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    results: hold
    cache.lazy: false
---

## Overview

This script fits a **Boosted Regression Trees (BRT)** model as a secondary model
for comparison with MaxEnt. BRT captures non-linear relationships and interactions
that MaxEnt may miss, and has been shown to outperform linear models for tick
distribution modeling (90.6% vs 80.6% accuracy; PMC10054924).

We use the same occurrence data, background points, and environmental variables
as the MaxEnt model (Step 5) for direct comparability.

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(dismo)
library(gbm)
```

## 1. Load data

```{r}
#| label: load-data

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/SDM/Dvariabilis_SDM"
processed_dir <- file.path(base_dir, "processed_data")
figures_dir <- file.path(base_dir, "figures")

# Occurrence data with environmental values
occ_env <- readRDS(file.path(processed_dir, "occurrences_with_env.rds"))
bg_env <- readRDS(file.path(processed_dir, "background_with_env.rds"))

# Selected variables
selected_vars <- readLines(file.path(processed_dir, "selected_variables.txt"))

sprintf("Occurrences: %d | Background: %d | Variables: %d",
        nrow(occ_env), nrow(bg_env), length(selected_vars))
```

## 2. Prepare presence/absence data for BRT

BRT requires a binary response: 1 for presence, 0 for pseudo-absence (background).

We also create observation weights to address the prevalence imbalance between
presence and background points. Without weighting, BRT optimises overall deviance
and learns "mostly absent" as the baseline, producing diffuse moderate-suitability
predictions and artificially low thresholds.

```{r}
#| label: prep-data

# Presence data (extract selected variables)
pres_data <- occ_env %>%
  dplyr::select(all_of(selected_vars)) %>%
  mutate(presence = 1)

# Background (pseudo-absence) data
bg_data <- bg_env %>%
  dplyr::select(all_of(selected_vars)) %>%
  mutate(presence = 0)

# Combine
brt_data <- bind_rows(pres_data, bg_data)

# Create weights to equalise class influence
# Presence weight = n_bg / n_pres, background weight = 1
n_pres <- sum(brt_data$presence == 1)
n_bg <- sum(brt_data$presence == 0)
brt_data$weights <- ifelse(brt_data$presence == 1, n_bg / n_pres, 1)

sprintf("BRT training data: %d rows (%d presence, %d background)",
        nrow(brt_data), n_pres, n_bg)
sprintf("Prevalence: %.1f%% — presence weight: %.2f, background weight: 1.00",
        100 * n_pres / nrow(brt_data), n_bg / n_pres)
```

## 3. Fit BRT model using dismo::gbm.step

We use `gbm.step()` which automatically selects the optimal number of trees
via cross-validation.

```{r}
#| label: fit-brt
#| cache: true

set.seed(42)

cat("Fitting BRT model — this may take several minutes...\n")

# Identify predictor column indices (exclude presence and weights columns)
pred_cols <- which(names(brt_data) %in% selected_vars)

brt_model <- gbm.step(
  data = as.data.frame(brt_data),
  gbm.x = pred_cols,
  gbm.y = which(names(brt_data) == "presence"),
  family = "bernoulli",
  tree.complexity = 5,     # Interaction depth
  learning.rate = 0.005,   # Learning rate
  bag.fraction = 0.75,     # Stochastic fraction
  n.folds = 10,            # CV folds
  site.weights = brt_data$weights,  # Equalise class influence
  verbose = TRUE
)

# If learning rate is too high (< 1000 trees), reduce and retry
if (is.null(brt_model) || brt_model$n.trees < 1000) {
  cat("Reducing learning rate to 0.001 for more trees...\n")
  brt_model <- gbm.step(
    data = as.data.frame(brt_data),
    gbm.x = pred_cols,
    gbm.y = which(names(brt_data) == "presence"),
    family = "bernoulli",
    tree.complexity = 5,
    learning.rate = 0.001,
    bag.fraction = 0.75,
    n.folds = 10,
    site.weights = brt_data$weights,
    verbose = TRUE
  )
}

sprintf("BRT model fitted: %d trees, CV deviance = %.4f (+/- %.4f)",
        brt_model$n.trees, brt_model$cv.statistics$deviance.mean,
        brt_model$cv.statistics$deviance.se)
```

## 4. Variable importance

```{r}
#| label: var-importance

# Get relative influence
brt_importance <- summary(brt_model, plotit = FALSE) %>%
  as_tibble() %>%
  rename(variable = var, relative_influence = rel.inf) %>%
  arrange(desc(relative_influence))

cat("BRT Variable Importance:\n")
print(brt_importance, n = Inf)

# Plot variable importance
ggplot(brt_importance, aes(x = reorder(variable, relative_influence),
                           y = relative_influence)) +
  geom_col(fill = "#009E73") +
  coord_flip() +
  labs(
    title = expression(italic("D. variabilis") ~ "— BRT Variable Importance"),
    x = "Variable",
    y = "Relative Influence (%)"
  ) +
  theme_minimal(base_size = 12)

ggsave(
  file.path(figures_dir, "05b_brt_variable_importance.png"),
  width = 8, height = 6, dpi = 300
)
```

## 5. Partial dependence plots (response curves)

```{r}
#| label: partial-dependence
#| fig-width: 12
#| fig-height: 8

# Generate partial dependence plots for all variables
n_vars <- length(selected_vars)
ncol_plot <- 3
nrow_plot <- ceiling(n_vars / ncol_plot)

png(file.path(figures_dir, "05b_brt_partial_dependence.png"),
    width = 12, height = 2.5 * nrow_plot, units = "in", res = 300)
par(mfrow = c(nrow_plot, ncol_plot), mar = c(4, 4, 2, 1))

for (i in seq_along(selected_vars)) {
  gbm.plot(brt_model, variable.no = i, plot.layout = c(1, 1),
           write.title = FALSE, y.label = "Fitted function",
           show.contrib = TRUE)
  title(main = selected_vars[i], cex.main = 1)
}
dev.off()

# Also display inline
par(mfrow = c(nrow_plot, ncol_plot), mar = c(4, 4, 2, 1))
for (i in seq_along(selected_vars)) {
  gbm.plot(brt_model, variable.no = i, plot.layout = c(1, 1),
           write.title = FALSE, y.label = "Fitted function",
           show.contrib = TRUE)
  title(main = selected_vars[i], cex.main = 1)
}
```

## 6. Interaction effects

BRT can detect pairwise interactions between variables.

```{r}
#| label: interactions

# Find and display pairwise interactions
interactions <- gbm.interactions(brt_model)

cat("Top pairwise interactions:\n")
interactions$rank.list %>%
  head(10) %>%
  print()
```

## 7. Model performance summary

```{r}
#| label: performance

# Cross-validation performance
cv_deviance <- brt_model$cv.statistics$deviance.mean
cv_se <- brt_model$cv.statistics$deviance.se
cv_cor <- brt_model$cv.statistics$correlation.mean
cv_cor_se <- brt_model$cv.statistics$correlation.se

cat("--- BRT Model Performance ---\n")
sprintf("Number of trees: %d", brt_model$n.trees)
sprintf("Tree complexity: %d", brt_model$interaction.depth)
sprintf("Learning rate: %s", brt_model$shrinkage)
sprintf("CV deviance: %.4f (+/- %.4f)", cv_deviance, cv_se)
sprintf("CV correlation: %.4f (+/- %.4f)", cv_cor, cv_cor_se)

# Calculate AUC on training data for comparison with MaxEnt
train_preds <- predict(brt_model, brt_data, n.trees = brt_model$n.trees,
                       type = "response")
pres_preds <- train_preds[brt_data$presence == 1]
bg_preds <- train_preds[brt_data$presence == 0]

# Calculate AUC
auc_val <- 0
for (p in pres_preds) {
  auc_val <- auc_val + sum(p > bg_preds) + 0.5 * sum(p == bg_preds)
}
auc_val <- auc_val / (length(pres_preds) * length(bg_preds))
sprintf("Training AUC: %.3f", auc_val)
```

## 8. Save outputs

```{r}
#| label: save

# Save BRT model
saveRDS(brt_model, file.path(processed_dir, "brt_model.rds"))

# Save variable importance
write_csv(brt_importance, file.path(processed_dir, "brt_variable_importance.csv"))

# Save model metadata
brt_metadata <- list(
  n_trees = brt_model$n.trees,
  tree_complexity = brt_model$interaction.depth,
  learning_rate = brt_model$shrinkage,
  bag_fraction = brt_model$bag.fraction,
  cv_deviance = cv_deviance,
  cv_correlation = cv_cor,
  training_auc = auc_val,
  variables = selected_vars
)
saveRDS(brt_metadata, file.path(processed_dir, "brt_model_metadata.rds"))

sprintf("Saved BRT model and metadata to processed_data/")
```
