---
title: "Step 05d: BIOCLIM Model"
subtitle: "*Dermacentor variabilis* Species Distribution Model"
author: "Alexander W. Gofton"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme:
      light: flatly
      dark: solar
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    results: hold
    cache.lazy: false
---

## Overview

This script fits a **BIOCLIM** model as a third, independent SDM method. BIOCLIM uses
only occurrence data — no background points, no algorithm-specific fitting. It defines
suitability as the minimum percentile position of a location within the species' observed
climate envelope: a location must fall within the occurrence range for ALL variables to
receive a non-zero score.

**Key properties:**
- No extrapolation: novel climate conditions automatically receive score = 0
- No background points: not affected by target-group bias or background choices
- Deterministic: no stochastic fitting, always produces the same result
- Conservative: only predicts suitable where ALL variables are within occurrence range

**Interpretation:** BIOCLIM will give zero suitability at Innisfail (tropical Queensland)
because tropical conditions are outside the training climate envelope. This directly
confirms that the tropical training gap is inherent to the method — not an artifact of
MaxEnt or BRT parameterisation. The BIOCLIM result thus provides a useful floor estimate
(minimum confidence area) and validates the CLIMEX re-implementation as the appropriate
method for tropical extrapolation.

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(dismo)
library(patchwork)
```

## 1. Load data

```{r}
#| label: load

base_dir      <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/SDM/Dvariabilis_SDM"
processed_dir <- file.path(base_dir, "processed_data")
figures_dir   <- file.path(base_dir, "figures")

# Environmental stacks
env_stack_aus <- rast(file.path(processed_dir, "env_stack_aus.tif"))
env_stack_M   <- rast(file.path(processed_dir, "env_stack_M.tif"))

# Occurrence data
occ <- readRDS(file.path(processed_dir, "occurrences_thinned.rds"))
occ_env <- readRDS(file.path(processed_dir, "occurrences_with_env.rds"))

# Variable sets
set_B <- readLines(file.path(processed_dir, "selected_variables_setB.txt"))

# SDM thresholds and predictions for comparison
thresholds    <- readRDS(file.path(processed_dir, "model_thresholds.rds"))
pred_maxent_A <- rast(file.path(processed_dir, "pred_aus_maxent.tif"))
pred_brt_A    <- rast(file.path(processed_dir, "pred_aus_brt.tif"))

# Australia boundary
aus <- ne_countries(country = "Australia", scale = "medium", returnclass = "sf")

cat("Set B variables:", paste(set_B, collapse = ", "), "\n")
cat("Occurrences:", nrow(occ), "\n")
```

## 2. Prepare occurrence climate data

```{r}
#| label: occ-climate

# Extract Set B environmental values at occurrence points
occ_climate_B <- occ_env %>%
  dplyr::select(all_of(set_B)) %>%
  na.omit()

cat("Occurrence climate data (Set B):\n")
summary(occ_climate_B)
cat(sprintf("\nN = %d occurrence records with complete Set B climate data\n",
            nrow(occ_climate_B)))
```

## 3. Fit BIOCLIM model

```{r}
#| label: fit-bioclim

# Convert to raster stack for dismo (requires RasterStack)
env_aus_B_terra <- env_stack_aus[[set_B]]
env_aus_B_raster <- raster::stack(env_aus_B_terra)

# BIOCLIM requires the training data in a 2-column lon/lat format
# Extract coordinates and climate of occurrences
# dismo::bioclim() takes a matrix of training predictors
bc_model <- bioclim(as.matrix(occ_climate_B))

cat("BIOCLIM model fitted.\n")
cat("Model summary:\n")
print(bc_model)
```

## 4. Project BIOCLIM to Australia

```{r}
#| label: project
#| cache: true

pred_bioclim_aus <- predict(env_aus_B_raster, bc_model)
pred_bioclim_terra <- rast(pred_bioclim_aus)
names(pred_bioclim_terra) <- "bioclim_suitability"

cat("BIOCLIM projected to Australia.\n")
cat("Prediction range:", round(range(values(pred_bioclim_terra), na.rm = TRUE), 3), "\n")

# Area at various thresholds
cell_areas <- cellSize(pred_bioclim_terra, unit = "km")
aus_total_area <- sum(values(cell_areas * !is.na(pred_bioclim_terra)), na.rm = TRUE)

for (thr in c(0, 0.1, 0.2, 0.3, 0.5)) {
  area <- sum(values(cell_areas * (pred_bioclim_terra > thr)), na.rm = TRUE)
  cat(sprintf("  BIOCLIM > %.1f: %9.0f km² (%.1f%%)\n",
              thr, area, 100 * area / aus_total_area))
}
```

## 5. Validation at CLIMEX reference locations

```{r}
#| label: validation

ref_locs <- tibble(
  city         = c("Innisfail", "Lismore", "Sydney", "Melbourne",
                   "Brisbane", "Cairns", "Darwin", "Alice Springs", "Hobart"),
  lon          = c(146.03, 153.28, 151.21, 144.96,
                   153.03, 145.77, 130.84, 133.88, 147.33),
  lat          = c(-17.53, -28.81, -33.87, -37.81,
                   -27.47, -16.92, -12.46, -23.70, -42.88),
  climex_EI    = c(63, 43, NA, NA, NA, NA, NA, NA, NA),
  climex_note  = c("Highest EI", "~ Little Rock AR", "", "", "", "Tropical",
                   "Tropical monsoon", "Arid", "Cool temperate")
)

ref_pts <- vect(ref_locs, geom = c("lon", "lat"), crs = "EPSG:4326")

ref_locs$bioclim  <- extract(pred_bioclim_terra, ref_pts)[, 2]
ref_locs$maxent_A <- extract(pred_maxent_A, ref_pts)[, 2]
ref_locs$brt_A    <- extract(pred_brt_A, ref_pts)[, 2]

cat("Reference location predictions — all models:\n")
ref_locs %>%
  mutate(across(c(bioclim, maxent_A, brt_A), ~round(., 3))) %>%
  print(n = Inf)

cat("\nKey observation:\n")
innisfail_bc <- ref_locs %>% filter(city == "Innisfail") %>% pull(bioclim)
cat(sprintf("  Innisfail BIOCLIM = %.3f — ", innisfail_bc))
if (!is.na(innisfail_bc) && innisfail_bc == 0) {
  cat("zero suitability confirms tropical training gap is method-independent\n")
} else if (!is.na(innisfail_bc) && innisfail_bc > 0) {
  cat("some suitability predicted (review)\n")
} else {
  cat("NA (outside model extent)\n")
}
```

## 6. MESS check for BIOCLIM predictions

```{r}
#| label: mess-check

# BIOCLIM by definition gives 0 for cells outside the training envelope.
# Check what fraction of BIOCLIM > 0 cells are within training range for all vars.
# This provides a direct measure of how conservative BIOCLIM is vs MaxEnt.

# The BIOCLIM prediction > 0 region is the "within-envelope" area.
bioclim_positive_area <- sum(values(cell_areas * (pred_bioclim_terra > 0)), na.rm = TRUE)
cat(sprintf("BIOCLIM > 0 area: %.0f km² (%.1f%% of Australia)\n",
            bioclim_positive_area, 100 * bioclim_positive_area / aus_total_area))
cat("This is the area where ALL Set B variables fall within the occurrence range.\n")
cat("It represents a minimum-confidence suitability estimate.\n")
```

## 7. Maps

```{r}
#| label: maps
#| fig-width: 14
#| fig-height: 10

bc_df <- as.data.frame(pred_bioclim_terra, xy = TRUE, na.rm = TRUE)
me_df <- as.data.frame(pred_maxent_A, xy = TRUE, na.rm = TRUE)
br_df <- as.data.frame(pred_brt_A, xy = TRUE, na.rm = TRUE)
names(me_df)[3] <- "suitability"
names(br_df)[3] <- "suitability"

# Combined panel map
p_bc <- ggplot() +
  geom_raster(data = bc_df, aes(x = x, y = y, fill = bioclim_suitability)) +
  geom_sf(data = aus, fill = NA, colour = "black", linewidth = 0.3) +
  scale_fill_viridis_c(
    name = "Suitability",
    option = "viridis",
    limits = c(0, 1),
    na.value = "grey95",
    trans = "sqrt"
  ) +
  labs(
    title = "BIOCLIM (Set B)",
    subtitle = "Zero = outside occurrence climate range"
  ) +
  theme_minimal(base_size = 10) +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10))

p_me <- ggplot() +
  geom_raster(data = me_df, aes(x = x, y = y, fill = suitability)) +
  geom_sf(data = aus, fill = NA, colour = "black", linewidth = 0.3) +
  scale_fill_viridis_c(
    name = "Suitability",
    option = "viridis",
    limits = c(0, 1),
    na.value = "grey95",
    trans = "sqrt"
  ) +
  labs(title = "MaxEnt (Set A)", subtitle = "56.4% suitable") +
  theme_minimal(base_size = 10) +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10))

p_br <- ggplot() +
  geom_raster(data = br_df, aes(x = x, y = y, fill = suitability)) +
  geom_sf(data = aus, fill = NA, colour = "black", linewidth = 0.3) +
  scale_fill_viridis_c(
    name = "Suitability",
    option = "viridis",
    limits = c(0, 1),
    na.value = "grey95",
    trans = "sqrt"
  ) +
  labs(title = "BRT (Set A)", subtitle = "8.4% suitable") +
  theme_minimal(base_size = 10) +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10))

p_combined <- p_bc + p_me + p_br + plot_layout(ncol = 2)
print(p_combined)

ggsave(
  file.path(figures_dir, "05d_bioclim_comparison.png"),
  p_combined,
  width = 14,
  height = 10,
  dpi = 300
)
```

## 8. Save outputs

```{r}
#| label: save

writeRaster(pred_bioclim_terra,
            file.path(processed_dir, "pred_aus_bioclim.tif"),
            overwrite = TRUE)

write_csv(ref_locs, file.path(processed_dir, "bioclim_reference_locations.csv"))

cat("======================================================\n")
cat("  BIOCLIM MODEL SUMMARY\n")
cat("======================================================\n\n")

cat(sprintf("Method: BIOCLIM (dismo::bioclim)\n"))
cat(sprintf("Variables: Set B — %s\n\n", paste(set_B, collapse = ", ")))

cat("Area estimates:\n")
for (thr in c(0, 0.1, 0.2, 0.3)) {
  area <- sum(values(cell_areas * (pred_bioclim_terra > thr)), na.rm = TRUE)
  cat(sprintf("  > %.1f: %9.0f km² (%.1f%%)\n", thr, area,
              100 * area / aus_total_area))
}

cat("\nComparison interpretation:\n")
cat("BIOCLIM gives the most conservative estimate — it predicts where the\n")
cat("Australian climate falls strictly within the North American occurrence envelope.\n")
cat("Zero suitability at Innisfail confirms that tropical/subtropical zones cannot\n")
cat("be reliably predicted by ANY correlative SDM — this is a fundamental\n")
cat("limitation requiring the CLIMEX physiological approach (05e).\n")

sprintf("Saved BIOCLIM outputs")
```
