---
title: "Step 6: Model Evaluation"
subtitle: "*Dermacentor variabilis* Species Distribution Model"
author: "Alexander W. Gofton"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme:
      light: flatly
      dark: solar
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    results: hold
    cache.lazy: false
---

## Overview

This script evaluates and compares the MaxEnt and BRT models from Steps 5 and 5b.
Evaluation includes:

1. Discrimination metrics (AUC, TSS, Boyce index)
2. Response curve comparison
3. Variable importance comparison
4. Threshold selection (MaxSS, 10% training presence)
5. Spatial cross-validation performance

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(dismo)
library(gbm)
library(ENMeval)
library(ecospat)
```

## 1. Load models and data

```{r}
#| label: load
#| cache: false

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/SDM/Dvariabilis_SDM"
processed_dir <- file.path(base_dir, "processed_data")
figures_dir <- file.path(base_dir, "figures")

# Models
maxent_model <- readRDS(file.path(processed_dir, "selected_maxent_model.rds"))
brt_model <- readRDS(file.path(processed_dir, "brt_model.rds"))
enmeval_results <- readRDS(file.path(processed_dir, "enmeval_results.rds"))
maxent_metadata <- readRDS(file.path(processed_dir, "selected_model_metadata.rds"))

# Data
occ_env <- readRDS(file.path(processed_dir, "occurrences_with_env.rds"))
bg_env <- readRDS(file.path(processed_dir, "background_with_env.rds"))
selected_vars <- readLines(file.path(processed_dir, "selected_variables.txt"))
env_M <- rast(file.path(processed_dir, "env_selected_M.tif"))

sprintf("MaxEnt: fc=%s, rm=%s | BRT: %d trees, tc=%d",
        maxent_metadata$feature_class, maxent_metadata$reg_multiplier,
        brt_model$n.trees, brt_model$interaction.depth)
```

## 2. Generate predictions on training data

```{r}
#| label: predictions

# MaxEnt predictions at occurrence and background points
occ_coords <- occ_env %>% dplyr::select(decimalLongitude, decimalLatitude)
bg_coords <- bg_env %>% dplyr::select(lon, lat)

# Extract env values for prediction
occ_env_vals <- occ_env %>% dplyr::select(all_of(selected_vars))
bg_env_vals <- bg_env %>% dplyr::select(all_of(selected_vars))

# MaxEnt predictions
maxent_pres_pred <- predict(maxent_model, occ_env_vals, type = "cloglog")
maxent_bg_pred <- predict(maxent_model, bg_env_vals, type = "cloglog")

# BRT predictions
brt_pres_pred <- predict(brt_model, as.data.frame(occ_env_vals),
                         n.trees = brt_model$n.trees, type = "response")
brt_bg_pred <- predict(brt_model, as.data.frame(bg_env_vals),
                       n.trees = brt_model$n.trees, type = "response")
```

## 3. AUC calculation

```{r}
#| label: auc

# Function to calculate AUC from presence and background predictions
calc_auc <- function(pres_pred, bg_pred) {
  n_pres <- length(pres_pred)
  n_bg <- length(bg_pred)
  auc <- 0
  for (p in pres_pred) {
    auc <- auc + sum(p > bg_pred) + 0.5 * sum(p == bg_pred)
  }
  auc / (n_pres * n_bg)
}

maxent_auc <- calc_auc(maxent_pres_pred, maxent_bg_pred)
brt_auc <- calc_auc(brt_pres_pred, brt_bg_pred)

sprintf("MaxEnt training AUC: %.3f", maxent_auc)
sprintf("BRT training AUC: %.3f", brt_auc)
```

## 4. Boyce index (continuous)

The Boyce index assesses how predictions differ from a random distribution
of observed presences. Values close to 1 indicate good model performance.

```{r}
#| label: boyce

# MaxEnt Boyce index
maxent_boyce <- ecospat.boyce(
  fit = c(maxent_pres_pred, maxent_bg_pred),
  obs = maxent_pres_pred,
  nclass = 0,
  PEplot = FALSE
)

# BRT Boyce index
brt_boyce <- ecospat.boyce(
  fit = c(brt_pres_pred, brt_bg_pred),
  obs = brt_pres_pred,
  nclass = 0,
  PEplot = FALSE
)

sprintf("MaxEnt Boyce index (Spearman cor): %.3f", maxent_boyce$cor)
sprintf("BRT Boyce index (Spearman cor): %.3f", brt_boyce$cor)
```

## 5. Threshold selection

```{r}
#| label: thresholds

# Function to find optimal thresholds
find_thresholds <- function(pres_pred, bg_pred) {
  # Combine predictions
  all_pred <- c(pres_pred, bg_pred)
  all_obs <- c(rep(1, length(pres_pred)), rep(0, length(bg_pred)))

  # Test a range of thresholds
  thresholds <- seq(0, 1, by = 0.001)

  results <- map_dfr(thresholds, function(t) {
    pred_binary <- ifelse(all_pred >= t, 1, 0)
    tp <- sum(pred_binary == 1 & all_obs == 1)
    fp <- sum(pred_binary == 1 & all_obs == 0)
    tn <- sum(pred_binary == 0 & all_obs == 0)
    fn <- sum(pred_binary == 0 & all_obs == 1)

    sens <- tp / (tp + fn)
    spec <- tn / (tn + fp)
    tss <- sens + spec - 1

    tibble(
      threshold = t,
      sensitivity = sens,
      specificity = spec,
      tss = tss,
      sum_ss = sens + spec
    )
  })

  # MaxSS threshold (maximises sensitivity + specificity)
  maxss <- results %>% filter(sum_ss == max(sum_ss)) %>% slice(1)

  # 10% training presence threshold
  p10 <- quantile(pres_pred, probs = 0.10)

  list(
    maxss = maxss,
    p10 = p10,
    all_results = results
  )
}

maxent_thresh <- find_thresholds(maxent_pres_pred, maxent_bg_pred)
brt_thresh <- find_thresholds(brt_pres_pred, brt_bg_pred)

cat("--- MaxEnt Thresholds ---\n")
sprintf("MaxSS threshold: %.3f (TSS = %.3f, Sens = %.3f, Spec = %.3f)",
        maxent_thresh$maxss$threshold, maxent_thresh$maxss$tss,
        maxent_thresh$maxss$sensitivity, maxent_thresh$maxss$specificity)
sprintf("10%% training presence threshold: %.3f", maxent_thresh$p10)

cat("\n--- BRT Thresholds ---\n")
sprintf("MaxSS threshold: %.3f (TSS = %.3f, Sens = %.3f, Spec = %.3f)",
        brt_thresh$maxss$threshold, brt_thresh$maxss$tss,
        brt_thresh$maxss$sensitivity, brt_thresh$maxss$specificity)
sprintf("10%% training presence threshold: %.3f", brt_thresh$p10)
```

## 6. Variable importance comparison

```{r}
#| label: var-importance-comparison
#| fig-width: 10
#| fig-height: 6

# MaxEnt variable importance
maxent_imp <- read_csv(file.path(processed_dir, "variable_importance_final.csv"),
                       show_col_types = FALSE) %>%
  dplyr::select(variable, permutation.importance) %>%
  rename(importance = permutation.importance) %>%
  mutate(model = "MaxEnt")

# BRT variable importance
brt_imp <- read_csv(file.path(processed_dir, "brt_variable_importance.csv"),
                    show_col_types = FALSE) %>%
  rename(importance = relative_influence) %>%
  mutate(model = "BRT")

# Combine
imp_comparison <- bind_rows(maxent_imp, brt_imp)

ggplot(imp_comparison, aes(x = reorder(variable, importance), y = importance,
                           fill = model)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("MaxEnt" = "#0072B2", "BRT" = "#009E73")) +
  labs(
    title = expression(italic("D. variabilis") ~ "— Variable Importance: MaxEnt vs BRT"),
    x = "Variable",
    y = "Importance (%)",
    fill = "Model"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")

ggsave(
  file.path(figures_dir, "06_variable_importance_comparison.png"),
  width = 10, height = 6, dpi = 300
)
```

## 7. Model agreement assessment

Compare spatial predictions between MaxEnt and BRT to quantify divergence.
High AUC for both models does not guarantee they agree spatially — AUC measures
rank discrimination within the training region, not transferability.

```{r}
#| label: model-agreement

# Correlation between MaxEnt and BRT predictions at presence points
pred_cor_pres <- cor(maxent_pres_pred, brt_pres_pred, method = "spearman")
pred_cor_bg <- cor(maxent_bg_pred, brt_bg_pred, method = "spearman")

cat("--- Prediction Correlation (Training Data) ---\n")
sprintf("At presence points (Spearman): %.3f", pred_cor_pres)
sprintf("At background points (Spearman): %.3f", pred_cor_bg)

cat("\nInterpretation:\n")
if (pred_cor_pres > 0.7 & pred_cor_bg > 0.7) {
  cat("  Models show good agreement in prediction space.\n")
} else {
  cat("  Models show divergent predictions. This may reflect:\n")
  cat("  - Different variable importance rankings (multicollinearity artefact)\n")
  cat("  - MaxEnt's multiplicative feature interactions vs BRT's additive trees\n")
  cat("  - Prevalence imbalance effects on BRT threshold calibration\n")
  cat("  Ensemble predictions should be interpreted with caution.\n")
}
```

## 8. Model comparison summary

```{r}
#| label: comparison-table

comparison <- tibble(
  Metric = c("Training AUC", "Validation AUC (CV)", "Boyce Index",
             "MaxSS Threshold", "TSS at MaxSS",
             "Sensitivity at MaxSS", "Specificity at MaxSS",
             "10% Training Presence Threshold"),
  MaxEnt = c(
    round(maxent_auc, 3),
    round(maxent_metadata$auc_val, 3),
    round(maxent_boyce$cor, 3),
    round(maxent_thresh$maxss$threshold, 3),
    round(maxent_thresh$maxss$tss, 3),
    round(maxent_thresh$maxss$sensitivity, 3),
    round(maxent_thresh$maxss$specificity, 3),
    round(maxent_thresh$p10, 3)
  ),
  BRT = c(
    round(brt_auc, 3),
    round(brt_model$cv.statistics$correlation.mean, 3),
    round(brt_boyce$cor, 3),
    round(brt_thresh$maxss$threshold, 3),
    round(brt_thresh$maxss$tss, 3),
    round(brt_thresh$maxss$sensitivity, 3),
    round(brt_thresh$maxss$specificity, 3),
    round(brt_thresh$p10, 3)
  )
)

cat("--- Model Comparison ---\n")
print(comparison, n = Inf)

write_csv(comparison, file.path(processed_dir, "model_comparison.csv"))
```

## 9. Save evaluation outputs

```{r}
#| label: save

# Thresholds
threshold_data <- list(
  maxent = list(
    maxss = maxent_thresh$maxss$threshold,
    p10 = as.numeric(maxent_thresh$p10),
    tss = maxent_thresh$maxss$tss
  ),
  brt = list(
    maxss = brt_thresh$maxss$threshold,
    p10 = as.numeric(brt_thresh$p10),
    tss = brt_thresh$maxss$tss
  )
)
saveRDS(threshold_data, file.path(processed_dir, "model_thresholds.rds"))

# Primary threshold for downstream use (MaxEnt MaxSS)
writeLines(
  as.character(maxent_thresh$maxss$threshold),
  file.path(processed_dir, "primary_threshold.txt")
)

sprintf("Saved model evaluation results and thresholds")
```
