---
title: "Step 6b: Response Curve Diagnosis — MaxEnt vs BRT"
subtitle: "*Dermacentor variabilis* Species Distribution Model"
author: "Alexander W. Gofton"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme:
      light: flatly
      dark: solar
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    results: hold
    cache.lazy: false
---

## Overview

This diagnostic script overlays MaxEnt and BRT response curves (marginal effects)
on the same axes for each variable, with the **Australian environmental range** shaded.
This reveals exactly **where and why** the two models diverge when projected to Australia.

Variables where the models produce different response shapes within the Australian
climate range are the primary drivers of the 7x disagreement in predicted suitable
area (MaxEnt 56.4% vs BRT 8.4%).

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(dismo)
library(gbm)
```

## 1. Load models and data

```{r}
#| label: load

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/SDM/Dvariabilis_SDM"
processed_dir <- file.path(base_dir, "processed_data")
figures_dir <- file.path(base_dir, "figures")

# Models
maxent_model <- readRDS(file.path(processed_dir, "selected_maxent_model.rds"))
brt_model <- readRDS(file.path(processed_dir, "brt_model.rds"))

# Environmental stacks
env_M <- rast(file.path(processed_dir, "env_selected_M.tif"))
env_aus <- rast(file.path(processed_dir, "env_selected_aus.tif"))

# Occurrence data (for computing means)
occ_env <- readRDS(file.path(processed_dir, "occurrences_with_env.rds"))
selected_vars <- readLines(file.path(processed_dir, "selected_variables.txt"))

sprintf("Variables: %s", paste(selected_vars, collapse = ", "))
```

## 2. Extract Australian variable ranges

```{r}
#| label: aus-ranges

# Get Australian environmental ranges for shading
aus_ranges <- list()
for (v in selected_vars) {
  vals <- values(env_aus[[v]], na.rm = TRUE)
  aus_ranges[[v]] <- c(
    min = min(vals),
    q10 = quantile(vals, 0.10),
    q90 = quantile(vals, 0.90),
    max = max(vals)
  )
}

# Training data ranges
train_ranges <- list()
for (v in selected_vars) {
  vals <- values(env_M[[v]], na.rm = TRUE)
  train_ranges[[v]] <- c(min = min(vals), max = max(vals))
}

cat("Variable ranges — Training vs Australia:\n")
for (v in selected_vars) {
  sprintf("  %s: Train [%.1f, %.1f] | Aus [%.1f, %.1f]",
          v, train_ranges[[v]]["min"], train_ranges[[v]]["max"],
          aus_ranges[[v]]["min"], aus_ranges[[v]]["max"])
}
```

## 3. Generate MaxEnt and BRT response curves

For each variable, predict suitability across its full range while holding all
other variables at their mean values (from occurrence points). Overlay both
models on the same axis.

```{r}
#| label: response-curves
#| fig-width: 14
#| fig-height: 10

# Compute mean environmental values at occurrence points
mean_vals <- as.data.frame(t(colMeans(
  occ_env %>% dplyr::select(all_of(selected_vars)),
  na.rm = TRUE
)))

# Generate response curves for both models
response_data <- list()

for (v in selected_vars) {
  # Use the full range spanning both training and Australian environments
  combined_min <- min(train_ranges[[v]]["min"], aus_ranges[[v]]["min"])
  combined_max <- max(train_ranges[[v]]["max"], aus_ranges[[v]]["max"])
  var_seq <- seq(combined_min, combined_max, length.out = 200)

  # Create prediction data frame with other vars at mean
  pred_df <- mean_vals[rep(1, 200), ]
  pred_df[[v]] <- var_seq

  # MaxEnt prediction (cloglog)
  maxent_pred <- predict(maxent_model, pred_df, type = "cloglog")

  # BRT prediction (response)
  brt_pred <- predict(brt_model, as.data.frame(pred_df),
                      n.trees = brt_model$n.trees, type = "response")

  response_data[[v]] <- tibble(
    variable = v,
    value = var_seq,
    MaxEnt = maxent_pred,
    BRT = brt_pred,
    in_training = value >= train_ranges[[v]]["min"] & value <= train_ranges[[v]]["max"],
    in_australia = value >= aus_ranges[[v]]["min"] & value <= aus_ranges[[v]]["max"]
  )
}

response_all <- bind_rows(response_data) %>%
  pivot_longer(cols = c(MaxEnt, BRT), names_to = "Model", values_to = "Suitability")

# Plot with Australian range shaded
p <- ggplot() +
  # Shade Australian range
  geom_rect(
    data = tibble(
      variable = selected_vars,
      xmin = sapply(aus_ranges, `[`, "min"),
      xmax = sapply(aus_ranges, `[`, "max")
    ),
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
    fill = "#009E73", alpha = 0.15
  ) +
  # Mark training range boundaries
  geom_vline(
    data = tibble(
      variable = selected_vars,
      xmin = sapply(train_ranges, `[`, "min"),
      xmax = sapply(train_ranges, `[`, "max")
    ) %>% pivot_longer(c(xmin, xmax), values_to = "x"),
    aes(xintercept = x),
    linetype = "dashed", colour = "grey50", linewidth = 0.4
  ) +
  # Response curves
  geom_line(data = response_all,
            aes(x = value, y = Suitability, colour = Model),
            linewidth = 0.9) +
  facet_wrap(~variable, scales = "free_x", ncol = 3) +
  scale_colour_manual(values = c("MaxEnt" = "#0072B2", "BRT" = "#D55E00")) +
  labs(
    title = expression(italic("D. variabilis") ~ "— Response Curve Comparison: MaxEnt vs BRT"),
    subtitle = "Green shading = Australian environmental range | Dashed lines = training data range boundaries",
    x = "Environmental Value",
    y = "Habitat Suitability",
    colour = "Model"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  ) +
  ylim(0, 1)

print(p)

ggsave(
  file.path(figures_dir, "06b_response_curve_comparison.png"),
  p, width = 14, height = 10, dpi = 300
)
```

## 4. Quantify model divergence within Australian range

For each variable, calculate the mean absolute difference between MaxEnt and BRT
predictions within the Australian environmental range.

```{r}
#| label: divergence

divergence <- bind_rows(response_data) %>%
  filter(in_australia) %>%
  group_by(variable) %>%
  summarise(
    mean_abs_diff = mean(abs(MaxEnt - BRT)),
    max_abs_diff = max(abs(MaxEnt - BRT)),
    mean_maxent = mean(MaxEnt),
    mean_brt = mean(BRT),
    maxent_higher_pct = 100 * mean(MaxEnt > BRT),
    aus_range_in_training_pct = 100 * mean(in_training),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_abs_diff))

cat("Model divergence within Australian environmental range:\n")
print(divergence, n = Inf)

cat("\nVariables driving MaxEnt-BRT disagreement (ranked by mean absolute difference):\n")
for (i in seq_len(nrow(divergence))) {
  row <- divergence[i, ]
  dir <- ifelse(row$maxent_higher_pct > 50, "MaxEnt higher", "BRT higher")
  sprintf("  %d. %s: MAD = %.3f, %s in %.0f%% of Aus range, %.0f%% within training range",
          i, row$variable, row$mean_abs_diff, dir,
          row$maxent_higher_pct, row$aus_range_in_training_pct)
}
```

## 5. Extrapolation zone analysis

Identify which variables have Australian values that fall outside the training range.

```{r}
#| label: extrapolation

cat("Extrapolation analysis — Australian values outside training range:\n\n")

extrap_summary <- tibble()

for (v in selected_vars) {
  aus_vals <- values(env_aus[[v]], na.rm = TRUE)
  tr_min <- train_ranges[[v]]["min"]
  tr_max <- train_ranges[[v]]["max"]

  below <- sum(aus_vals < tr_min)
  above <- sum(aus_vals > tr_max)
  total <- length(aus_vals)

  extrap_summary <- bind_rows(extrap_summary, tibble(
    variable = v,
    pct_below = round(100 * below / total, 1),
    pct_above = round(100 * above / total, 1),
    pct_novel = round(100 * (below + above) / total, 1),
    pct_analogous = round(100 * (total - below - above) / total, 1),
    aus_min = round(min(aus_vals), 1),
    aus_max = round(max(aus_vals), 1),
    train_min = round(tr_min, 1),
    train_max = round(tr_max, 1)
  ))
}

print(extrap_summary, n = Inf)
```

## 6. Summary

```{r}
#| label: summary

cat("======================================================\n")
cat("  RESPONSE CURVE DIAGNOSIS SUMMARY\n")
cat("======================================================\n\n")

cat("Key findings:\n")
cat("1. Variables with highest MaxEnt-BRT divergence within Australian range\n")
cat("   indicate which predictors drive the 7x area disagreement.\n\n")

cat("2. Variables where Australian values fall outside training range (extrapolation)\n")
cat("   cannot be reliably projected by either model.\n\n")

cat("3. Bio1 response curves likely show the largest divergence because:\n")
cat("   - MaxEnt assigns 66.4% contribution to Bio1 (dominant predictor)\n")
cat("   - BRT distributes importance more evenly (Bio1 = 29.9%)\n")
cat("   - Different response shapes under Australian Bio1 values → different projections\n\n")

cat("4. If divergence is concentrated in Bio1 and Bio7, removing these variables\n")
cat("   should substantially improve model agreement (see 05c_sensitivity_models.qmd).\n")

# Save divergence table
write_csv(divergence, file.path(processed_dir, "response_curve_divergence.csv"))
write_csv(extrap_summary, file.path(processed_dir, "extrapolation_summary.csv"))
sprintf("Saved response curve analysis outputs")
```
