---
title: "Step 7: Projection to Australia"
subtitle: "*Dermacentor variabilis* Species Distribution Model"
author: "Alexander W. Gofton"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme:
      light: flatly
      dark: solar
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    results: hold
    cache.lazy: false
---

## Overview

This script projects the fitted MaxEnt and BRT models to Australia to assess
climatic suitability for *D. variabilis*. Critical steps include:

1. Predict suitability within the native range (validation)
2. MESS analysis — identify where Australian climate is novel relative to training data
3. Project both models to Australia
4. Create binary suitability maps using selected thresholds
5. MESS-masked predictions (transparent about extrapolation zones)
6. Calculate area of suitable habitat in Australia

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(sf)
library(dismo)
library(gbm)
library(rnaturalearth)
library(rnaturalearthdata)
library(predicts)
```

## 1. Load models and data

```{r}
#| label: load
#| cache: false

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/SDM/Dvariabilis_SDM"
processed_dir <- file.path(base_dir, "processed_data")
figures_dir <- file.path(base_dir, "figures")

# Models
maxent_model <- readRDS(file.path(processed_dir, "selected_maxent_model.rds"))
brt_model <- readRDS(file.path(processed_dir, "brt_model.rds"))
thresholds <- readRDS(file.path(processed_dir, "model_thresholds.rds"))
selected_vars <- readLines(file.path(processed_dir, "selected_variables.txt"))

# Environmental stacks
env_M <- rast(file.path(processed_dir, "env_selected_M.tif"))
env_aus <- rast(file.path(processed_dir, "env_selected_aus.tif"))

# Occurrence and background data
occ_env <- readRDS(file.path(processed_dir, "occurrences_with_env.rds"))
bg_env <- readRDS(file.path(processed_dir, "background_with_env.rds"))

# Australia boundary
aus <- ne_countries(country = "Australia", scale = "medium", returnclass = "sf")

sprintf("Environmental layers — M: %d vars, %d x %d | Australia: %d vars, %d x %d",
        nlyr(env_M), nrow(env_M), ncol(env_M),
        nlyr(env_aus), nrow(env_aus), ncol(env_aus))
```

## 2. Predict within native range (validation)

```{r}
#| label: predict-native
#| cache: true

cat("Predicting MaxEnt suitability within native range...\n")
pred_M_maxent <- terra::predict(env_M, maxent_model, type = "cloglog")
names(pred_M_maxent) <- "maxent_suitability"

cat("Predicting BRT suitability within native range...\n")
pred_M_brt <- predict(env_M, brt_model, n.trees = brt_model$n.trees,
                      type = "response")
names(pred_M_brt) <- "brt_suitability"

sprintf("Native range prediction complete")
```

## 3. MESS analysis — Multivariate Environmental Similarity Surface

MESS identifies areas in Australia where climate conditions fall outside the
range of the training data. MESS < 0 indicates novel (extrapolated) conditions.

```{r}
#| label: mess
#| cache: true

# Reference data: environmental values at training points (presence + background)
ref_data <- bind_rows(
  occ_env %>% dplyr::select(all_of(selected_vars)),
  bg_env %>% dplyr::select(all_of(selected_vars))
)

# Calculate MESS for Australia
mess_aus <- mess(env_aus, as.data.frame(ref_data))
names(mess_aus) <- "mess"

# Summary statistics
mess_vals <- values(mess_aus, na.rm = TRUE)
n_novel <- sum(mess_vals < 0)
n_total <- length(mess_vals)
pct_novel <- round(100 * n_novel / n_total, 1)

sprintf("MESS Analysis for Australia:")
sprintf("  Total cells: %d", n_total)
sprintf("  Novel climate (MESS < 0): %d (%.1f%%)", n_novel, pct_novel)
sprintf("  Analogous climate (MESS >= 0): %d (%.1f%%)", n_total - n_novel, 100 - pct_novel)
sprintf("  MESS range: %.1f to %.1f", min(mess_vals), max(mess_vals))
```

## 4. Identify most dissimilar variable (MoD)

For cells with MESS < 0, identify which variable is most responsible for the novelty.

```{r}
#| label: mod

# Calculate per-variable similarity
var_similarity <- list()
for (v in selected_vars) {
  ref_range <- range(ref_data[[v]], na.rm = TRUE)
  var_layer <- env_aus[[v]]

  # Similarity: proportion of reference values smaller (0-100 scale)
  # Negative = below reference range, >100 = above reference range
  ref_sorted <- sort(ref_data[[v]])
  n_ref <- length(ref_sorted)

  sim <- app(var_layer, function(x) {
    vapply(x, function(xi) {
      if (is.na(xi)) return(NA_real_)
      if (xi < ref_range[1]) return(100 * (xi - ref_range[1]) / (ref_range[2] - ref_range[1]))
      if (xi > ref_range[2]) return(100 * (ref_range[2] - xi) / (ref_range[2] - ref_range[1]))
      p <- sum(ref_sorted <= xi) / n_ref
      min(p, 1 - p) * 200
    }, numeric(1))
  })
  var_similarity[[v]] <- sim
}

var_sim_stack <- rast(var_similarity)
names(var_sim_stack) <- selected_vars

# Most dissimilar variable (minimum similarity at each cell)
mod_layer <- which.min(var_sim_stack)
names(mod_layer) <- "most_dissimilar_var"

sprintf("Most dissimilar variable analysis complete")
```

## 5. Project to Australia

```{r}
#| label: predict-australia
#| cache: true

cat("Projecting MaxEnt to Australia...\n")
pred_aus_maxent <- terra::predict(env_aus, maxent_model, type = "cloglog", clamp = TRUE)
names(pred_aus_maxent) <- "maxent_suitability"

cat("Projecting BRT to Australia...\n")
pred_aus_brt <- terra::predict(env_aus, brt_model, n.trees = brt_model$n.trees,
                               type = "response", na.rm = TRUE)
names(pred_aus_brt) <- "brt_suitability"

# Ensemble (mean of both models)
pred_aus_ensemble <- mean(pred_aus_maxent, pred_aus_brt)
names(pred_aus_ensemble) <- "ensemble_suitability"

sprintf("Projection complete")
```

## 6. Binary suitability maps

Apply thresholds to create presence/absence maps.

```{r}
#| label: binary-maps

# MaxEnt binary (using MaxSS threshold)
maxent_thresh_val <- thresholds$maxent$maxss
maxent_binary <- pred_aus_maxent >= maxent_thresh_val
names(maxent_binary) <- "maxent_binary"

# BRT binary
brt_thresh_val <- thresholds$brt$maxss
brt_binary <- pred_aus_brt >= brt_thresh_val
names(brt_binary) <- "brt_binary"

# Ensemble binary (using mean of thresholds)
ensemble_thresh <- mean(maxent_thresh_val, brt_thresh_val)
ensemble_binary <- pred_aus_ensemble >= ensemble_thresh
names(ensemble_binary) <- "ensemble_binary"

sprintf("Binary maps created (MaxEnt threshold: %.3f, BRT threshold: %.3f)",
        maxent_thresh_val, brt_thresh_val)
```

## 7. MESS-masked binary maps

Mask areas with novel climate (MESS < 0) in the binary predictions.

```{r}
#| label: mess-mask

mess_mask <- mess_aus >= 0  # TRUE where climate is analogous

maxent_binary_masked <- maxent_binary * mess_mask
brt_binary_masked <- brt_binary * mess_mask
ensemble_binary_masked <- ensemble_binary * mess_mask

# Set masked areas to NA for visualisation
maxent_binary_masked[mess_mask == 0] <- NA
brt_binary_masked[mess_mask == 0] <- NA
ensemble_binary_masked[mess_mask == 0] <- NA

sprintf("MESS-masked binary maps created (%.1f%% of cells masked as novel climate)", pct_novel)
```

## 8. Calculate suitable habitat area

```{r}
#| label: area-calc

# Use terra::cellSize for accurate area calculation
cell_areas <- cellSize(maxent_binary, unit = "km")

# MaxEnt suitable area
maxent_suitable_area <- sum(values(cell_areas * maxent_binary), na.rm = TRUE)
maxent_suitable_area_masked <- sum(values(cell_areas * (maxent_binary_masked == 1)), na.rm = TRUE)

# BRT suitable area
brt_suitable_area <- sum(values(cell_areas * brt_binary), na.rm = TRUE)
brt_suitable_area_masked <- sum(values(cell_areas * (brt_binary_masked == 1)), na.rm = TRUE)

# Ensemble suitable area
ensemble_suitable_area <- sum(values(cell_areas * ensemble_binary), na.rm = TRUE)

# Total Australia land area
aus_total_area <- sum(values(cell_areas * !is.na(pred_aus_maxent)), na.rm = TRUE)

cat("--- Suitable Habitat Area in Australia ---\n")
sprintf("Total land area: %.0f km2", aus_total_area)
sprintf("")
sprintf("MaxEnt — Suitable: %.0f km2 (%.1f%% of Australia)",
        maxent_suitable_area, 100 * maxent_suitable_area / aus_total_area)
sprintf("MaxEnt — Suitable (MESS-masked): %.0f km2 (%.1f%%)",
        maxent_suitable_area_masked, 100 * maxent_suitable_area_masked / aus_total_area)
sprintf("")
sprintf("BRT — Suitable: %.0f km2 (%.1f%%)",
        brt_suitable_area, 100 * brt_suitable_area / aus_total_area)
sprintf("BRT — Suitable (MESS-masked): %.0f km2 (%.1f%%)",
        brt_suitable_area_masked, 100 * brt_suitable_area_masked / aus_total_area)
sprintf("")
sprintf("Ensemble — Suitable: %.0f km2 (%.1f%%)",
        ensemble_suitable_area, 100 * ensemble_suitable_area / aus_total_area)
```

## 9. Visualisation — Australia suitability maps

```{r}
#| label: maps
#| fig-width: 14
#| fig-height: 16

aus_sf <- ne_countries(
  country = "Australia",
  scale = "medium",
  returnclass = "sf"
)
states <- ne_states(country = "Australia", returnclass = "sf")

# Convert rasters to data frames for ggplot
maxent_df <- as.data.frame(pred_aus_maxent, xy = TRUE) %>%
  filter(!is.na(maxent_suitability))
brt_df <- as.data.frame(pred_aus_brt, xy = TRUE) %>%
  filter(!is.na(brt_suitability))
mess_df <- as.data.frame(mess_aus, xy = TRUE) %>% filter(!is.na(mess))
ensemble_df <- as.data.frame(pred_aus_ensemble, xy = TRUE) %>%
  filter(!is.na(ensemble_suitability))

# Common theme
map_theme <- theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(size = 13, face = "bold"),
    legend.position = "bottom",
    legend.key.width = unit(2, "cm"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )

# MaxEnt suitability map
p1 <- ggplot() +
  geom_raster(data = maxent_df, aes(x = x, y = y, fill = maxent_suitability)) +
  geom_sf(data = states, fill = NA, colour = "grey40", linewidth = 0.3) +
  geom_sf(data = aus_sf, fill = NA, colour = "black", linewidth = 0.5) +
  scale_fill_viridis_c(
    option = "inferno",
    limits = c(0, 1),
    name = "Suitability"
  ) +
  labs(title = "A) MaxEnt — Predicted Habitat Suitability") +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10)) +
  map_theme

# BRT suitability map
p2 <- ggplot() +
  geom_raster(data = brt_df, aes(x = x, y = y, fill = brt_suitability)) +
  geom_sf(data = states, fill = NA, colour = "grey40", linewidth = 0.3) +
  geom_sf(data = aus_sf, fill = NA, colour = "black", linewidth = 0.5) +
  scale_fill_viridis_c(
    option = "inferno",
    limits = c(0, 1),
    name = "Suitability",
    oob = scales::squish
  ) +
  labs(title = "B) BRT — Predicted Habitat Suitability") +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10)) +
  map_theme

# MESS map
p3 <- ggplot() +
  geom_raster(data = mess_df, aes(x = x, y = y, fill = mess)) +
  geom_sf(data = states, fill = NA, colour = "grey40", linewidth = 0.3) +
  geom_sf(data = aus_sf, fill = NA, colour = "black", linewidth = 0.5) +
  scale_fill_gradient2(
    low = "#D55E00",
    mid = "white",
    high = "#0072B2",
    midpoint = 0,
    name = "MESS",
    limits = c(min(mess_df$mess), max(mess_df$mess))
  ) +
  labs(
    title = "C) MESS — Multivariate Environmental Similarity",
    subtitle = "Red = novel climate (extrapolation), Blue = analogous climate"
  ) +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10)) +
  map_theme

# Ensemble suitability map
p4 <- ggplot() +
  geom_raster(
    data = ensemble_df,
    aes(x = x, y = y, fill = ensemble_suitability)
  ) +
  geom_sf(data = states, fill = NA, colour = "grey40", linewidth = 0.3) +
  geom_sf(data = aus_sf, fill = NA, colour = "black", linewidth = 0.5) +
  scale_fill_viridis_c(
    option = "inferno",
    limits = c(0, 1),
    name = "Suitability"
  ) +
  labs(
    title = "D) Ensemble (MaxEnt + BRT mean) — Predicted Habitat Suitability"
  ) +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10)) +
  map_theme

library(patchwork)
combined_plot <- (p1 + p2) / (p3 + p4)

combined_plot +
  plot_annotation(
    title = expression(
      italic(
        "Dermacentor variabilis"
      ) ~ "— Projected Climatic Suitability in Australia"
    ),
    theme = theme(plot.title = element_text(size = 16, face = "bold"))
  )

ggsave(
  file.path(figures_dir, "07_australia_projection.png"),
  width = 14,
  height = 16,
  dpi = 300
)
```

## 10. Save outputs

```{r}
#| label: save

# Continuous predictions
writeRaster(pred_aus_maxent, file.path(processed_dir, "pred_aus_maxent.tif"), overwrite = TRUE)
writeRaster(pred_aus_brt, file.path(processed_dir, "pred_aus_brt.tif"), overwrite = TRUE)
writeRaster(pred_aus_ensemble, file.path(processed_dir, "pred_aus_ensemble.tif"), overwrite = TRUE)

# Binary predictions
writeRaster(maxent_binary, file.path(processed_dir, "pred_aus_maxent_binary.tif"), overwrite = TRUE)
writeRaster(brt_binary, file.path(processed_dir, "pred_aus_brt_binary.tif"), overwrite = TRUE)

# MESS
writeRaster(mess_aus, file.path(processed_dir, "mess_australia.tif"), overwrite = TRUE)

# Native range predictions
writeRaster(pred_M_maxent, file.path(processed_dir, "pred_M_maxent.tif"), overwrite = TRUE)
writeRaster(pred_M_brt, file.path(processed_dir, "pred_M_brt.tif"), overwrite = TRUE)

# Area summary
area_summary <- tibble(
  model = c("MaxEnt", "MaxEnt (MESS-masked)", "BRT", "BRT (MESS-masked)", "Ensemble"),
  suitable_area_km2 = c(maxent_suitable_area, maxent_suitable_area_masked,
                        brt_suitable_area, brt_suitable_area_masked,
                        ensemble_suitable_area),
  pct_australia = round(c(maxent_suitable_area, maxent_suitable_area_masked,
                          brt_suitable_area, brt_suitable_area_masked,
                          ensemble_suitable_area) / aus_total_area * 100, 1)
)
write_csv(area_summary, file.path(processed_dir, "suitable_area_summary.csv"))

sprintf("Saved all projection outputs, MESS analysis, and area summaries")
```
