---
title: "Step 7: Projection to Australia"
subtitle: "*Dermacentor variabilis* Species Distribution Model"
author: "Alexander W. Gofton"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme:
      light: flatly
      dark: solar
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    results: hold
    cache.lazy: false
---

## Overview

This script projects the fitted MaxEnt and BRT models to Australia to assess
climatic suitability for *D. variabilis*, and combines them with the CLIMEX
Ecoclimatic Index (EI) into a three-model ensemble. Critical steps include:

1. Predict suitability within the native range (validation)
2. MESS analysis — identify where Australian climate is novel relative to training data
3. Project both correlative models (MaxEnt, BRT) to Australia
4. Load CLIMEX EI predictions (process-based model, not subject to MESS limitations)
5. Build three-model ensemble: continuous weighted mean + committee vote (0–3 agreement)
6. Create binary suitability maps using selected thresholds
7. Calculate area of suitable habitat in Australia

**Why include CLIMEX in the ensemble?**
MaxEnt and BRT are correlative models that interpolate within the climate space
of the training data. Tropical north QLD (and similar regions) may fall outside
that training space (MESS < 0), causing these models to underpredict. CLIMEX is
a mechanistic/process-based model using physiological parameters (stress indices,
growth indices) that can operate beyond the training envelope. Combining all three
improves spatial coverage and reduces model-specific bias.

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(sf)
library(dismo)
library(gbm)
library(rnaturalearth)
library(rnaturalearthdata)
library(predicts)
library(viridis)
```

## 1. Load models and data

```{r}
#| label: load
#| cache: false

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/SDM/Dvariabilis_SDM"
processed_dir <- file.path(base_dir, "processed_data")
figures_dir <- file.path(base_dir, "figures")

# Models
maxent_model <- readRDS(file.path(processed_dir, "selected_maxent_model.rds"))
brt_model <- readRDS(file.path(processed_dir, "brt_model.rds"))
thresholds <- readRDS(file.path(processed_dir, "model_thresholds.rds"))
selected_vars <- readLines(file.path(processed_dir, "selected_variables.txt"))

# Environmental stacks
env_M <- rast(file.path(processed_dir, "env_selected_M.tif"))
env_aus <- rast(file.path(processed_dir, "env_selected_aus.tif"))

# Occurrence and background data
occ_env <- readRDS(file.path(processed_dir, "occurrences_with_env.rds"))
bg_env <- readRDS(file.path(processed_dir, "background_with_env.rds"))

# Australia boundary
aus <- ne_countries(country = "Australia", scale = "medium", returnclass = "sf")

sprintf("Environmental layers — M: %d vars, %d x %d | Australia: %d vars, %d x %d",
        nlyr(env_M), nrow(env_M), ncol(env_M),
        nlyr(env_aus), nrow(env_aus), ncol(env_aus))
```

## 2. Predict within native range (validation)

```{r}
#| label: predict-native
#| cache: true

cat("Predicting MaxEnt suitability within native range...\n")
pred_M_maxent <- terra::predict(env_M, maxent_model, type = "cloglog")
names(pred_M_maxent) <- "maxent_suitability"

cat("Predicting BRT suitability within native range...\n")
pred_M_brt <- predict(env_M, brt_model, n.trees = brt_model$n.trees,
                      type = "response")
names(pred_M_brt) <- "brt_suitability"

sprintf("Native range prediction complete")
```

## 3. MESS analysis — Multivariate Environmental Similarity Surface

MESS identifies areas in Australia where climate conditions fall outside the
range of the training data. MESS < 0 indicates novel (extrapolated) conditions.

```{r}
#| label: mess
#| cache: true

# Reference data: environmental values at training points (presence + background)
ref_data <- bind_rows(
  occ_env %>% dplyr::select(all_of(selected_vars)),
  bg_env %>% dplyr::select(all_of(selected_vars))
)

# Calculate MESS for Australia
mess_aus <- mess(env_aus, as.data.frame(ref_data))
names(mess_aus) <- "mess"

# Summary statistics
mess_vals <- values(mess_aus, na.rm = TRUE)
n_novel <- sum(mess_vals < 0)
n_total <- length(mess_vals)
pct_novel <- round(100 * n_novel / n_total, 1)

sprintf("MESS Analysis for Australia:")
sprintf("  Total cells: %d", n_total)
sprintf("  Novel climate (MESS < 0): %d (%.1f%%)", n_novel, pct_novel)
sprintf("  Analogous climate (MESS >= 0): %d (%.1f%%)", n_total - n_novel, 100 - pct_novel)
sprintf("  MESS range: %.1f to %.1f", min(mess_vals), max(mess_vals))
```

## 4. Identify most dissimilar variable (MoD)

For cells with MESS < 0, identify which variable is most responsible for the novelty.

```{r}
#| label: mod

# Calculate per-variable similarity
var_similarity <- list()
for (v in selected_vars) {
  ref_range <- range(ref_data[[v]], na.rm = TRUE)
  var_layer <- env_aus[[v]]

  # Similarity: proportion of reference values smaller (0-100 scale)
  # Negative = below reference range, >100 = above reference range
  ref_sorted <- sort(ref_data[[v]])
  n_ref <- length(ref_sorted)

  sim <- app(var_layer, function(x) {
    vapply(x, function(xi) {
      if (is.na(xi)) return(NA_real_)
      if (xi < ref_range[1]) return(100 * (xi - ref_range[1]) / (ref_range[2] - ref_range[1]))
      if (xi > ref_range[2]) return(100 * (ref_range[2] - xi) / (ref_range[2] - ref_range[1]))
      p <- sum(ref_sorted <= xi) / n_ref
      min(p, 1 - p) * 200
    }, numeric(1))
  })
  var_similarity[[v]] <- sim
}

var_sim_stack <- rast(var_similarity)
names(var_sim_stack) <- selected_vars

# Most dissimilar variable (minimum similarity at each cell)
mod_layer <- which.min(var_sim_stack)
names(mod_layer) <- "most_dissimilar_var"

sprintf("Most dissimilar variable analysis complete")
```

## 5. Project to Australia

```{r}
#| label: predict-australia
#| cache: true

cat("Projecting MaxEnt to Australia...\n")
pred_aus_maxent <- terra::predict(env_aus, maxent_model, type = "cloglog", clamp = TRUE)
names(pred_aus_maxent) <- "maxent_suitability"

cat("Projecting BRT to Australia...\n")
pred_aus_brt <- terra::predict(env_aus, brt_model, n.trees = brt_model$n.trees,
                               type = "response", na.rm = TRUE)
names(pred_aus_brt) <- "brt_suitability"

sprintf("Correlative model projection complete")
```

## 6. Load CLIMEX predictions and build three-model ensemble

CLIMEX EI (Ecoclimatic Index) is a process-based model output on a 0–100 scale.
We normalise to 0–1 by dividing by 100, then resample to match the correlative
model raster grid before combining.

```{r}
#| label: climex-ensemble

# Load CLIMEX outputs
climex_EI_raw  <- rast(file.path(processed_dir, "climex_EI_aus.tif"))
climex_binary  <- rast(file.path(processed_dir, "climex_suitable_aus.tif"))

# Normalise EI to 0–1 (CLIMEX uses 0–100 scale)
climex_EI_norm <- climex_EI_raw / 100
names(climex_EI_norm) <- "climex_suitability"

# Resample CLIMEX to match MaxEnt/BRT grid (bilinear for continuous, near for binary)
climex_EI_rs  <- resample(climex_EI_norm, pred_aus_maxent, method = "bilinear")
climex_bin_rs <- resample(climex_binary,  pred_aus_maxent, method = "near")
names(climex_EI_rs)  <- "climex_suitability"
names(climex_bin_rs) <- "climex_binary"

cat("CLIMEX EI range (normalised, resampled):\n")
print(global(climex_EI_rs, fun = "range", na.rm = TRUE))

# ---- Continuous three-model ensemble (data-driven skill-above-random AUC weights) ----
# Weights computed in 06_model_evaluation.qmd from spatial block CV AUC skill
# and CLIMEX WorldClim-bridge AUC. All inputs are on the same 0–1 scale.
ensemble_weights <- tryCatch(
  readRDS(file.path(processed_dir, "ensemble_weights.rds")),
  error = function(e) {
    warning("ensemble_weights.rds not found — falling back to equal weights. Run 06_model_evaluation.qmd first.")
    c(MaxEnt = 1/3, BRT = 1/3, CLIMEX = 1/3)
  }
)

cat(sprintf(
  "Ensemble weights: MaxEnt=%.3f, BRT=%.3f, CLIMEX=%.3f (sum=%.3f)\n",
  ensemble_weights["MaxEnt"], ensemble_weights["BRT"],
  ensemble_weights["CLIMEX"], sum(ensemble_weights)
))

pred_aus_ensemble <- (
  ensemble_weights["MaxEnt"] * pred_aus_maxent +
  ensemble_weights["BRT"]    * pred_aus_brt    +
  ensemble_weights["CLIMEX"] * climex_EI_rs
)
names(pred_aus_ensemble) <- "ensemble_suitability"

# ---- Committee vote (0, 1, 2, or 3 models agree) ----
# Each model casts a binary vote using its MaxSS threshold
maxent_thresh_val <- thresholds$maxent$maxss
brt_thresh_val    <- thresholds$brt$maxss
climex_thresh     <- 0.10  # EI > 10 (= 0.1 normalised) = potentially suitable in CLIMEX

maxent_vote <- ifel(pred_aus_maxent >= maxent_thresh_val, 1, 0)
brt_vote    <- ifel(pred_aus_brt    >= brt_thresh_val,    1, 0)
climex_vote <- ifel(climex_EI_rs    >= climex_thresh,     1, 0)

committee_vote <- maxent_vote + brt_vote + climex_vote
names(committee_vote) <- "models_agreeing"

# Summary
cat("--- Three-model ensemble summary ---\n")
cat("Continuous ensemble (equal-weight mean):\n")
print(global(pred_aus_ensemble, fun = c("min", "max", "mean"), na.rm = TRUE))

cat("\nCommittee vote distribution (N cells):\n")
vote_counts <- table(values(committee_vote))
print(vote_counts)
vote_pct <- round(100 * vote_counts / sum(!is.na(values(committee_vote))), 1)
cat("As % of land cells:\n")
print(vote_pct)
```

## 6b. Binary suitability maps

Apply thresholds to create presence/absence maps.

```{r}
#| label: binary-maps

# Individual model binary maps (thresholds already computed in section 6)
maxent_binary <- pred_aus_maxent >= maxent_thresh_val
brt_binary    <- pred_aus_brt    >= brt_thresh_val
climex_binary_map <- climex_bin_rs  # already binary from CLIMEX
names(maxent_binary)     <- "maxent_binary"
names(brt_binary)        <- "brt_binary"
names(climex_binary_map) <- "climex_binary"

# Ensemble binary: suitable where ≥ 2 of 3 models agree
ensemble_binary <- ifel(committee_vote >= 2, 1, 0)
names(ensemble_binary) <- "ensemble_binary"

# Conservative ensemble: all 3 models agree
ensemble_strict <- ifel(committee_vote == 3, 1, 0)
names(ensemble_strict) <- "ensemble_strict"

sprintf("Binary maps created (MaxEnt threshold: %.3f, BRT threshold: %.3f, CLIMEX threshold: %.2f)",
        maxent_thresh_val, brt_thresh_val, climex_thresh)
sprintf("Ensemble: suitable where ≥2/3 models agree. Strict: all 3 agree.")
```

## 7. MESS-masked binary maps

Mask areas with novel climate (MESS < 0) in the binary predictions.

```{r}
#| label: mess-mask

mess_mask <- mess_aus >= 0  # TRUE where climate is analogous

maxent_binary_masked   <- maxent_binary   * mess_mask
brt_binary_masked      <- brt_binary      * mess_mask
ensemble_binary_masked <- ensemble_binary * mess_mask

# Note: CLIMEX is process-based and does NOT require MESS masking —
# it is valid in novel climate zones by design. Only apply MESS to
# the correlative model predictions.
maxent_binary_masked[mess_mask == 0] <- NA
brt_binary_masked[mess_mask == 0]    <- NA
ensemble_binary_masked[mess_mask == 0] <- NA

# Committee vote with MESS annotation
# Cells where CLIMEX predicts suitable but correlative models are in novel space
# represent the highest-value CLIMEX-only predictions
climex_only_novel <- ifel(
  climex_binary_map == 1 & mess_mask == 0 &
    (maxent_vote + brt_vote) == 0,
  1, 0
)
names(climex_only_novel) <- "climex_only_novel_climate"

n_climex_novel <- sum(values(climex_only_novel) == 1, na.rm = TRUE)
sprintf("Cells predicted suitable by CLIMEX only AND in novel climate zone: %d", n_climex_novel)
sprintf("MESS-masked binary maps created (%.1f%% of cells masked as novel climate)", pct_novel)
```

## 8. Calculate suitable habitat area

```{r}
#| label: area-calc

# Use terra::cellSize for accurate area calculation
cell_areas <- cellSize(maxent_binary, unit = "km")

calc_area <- function(binary_r, areas = cell_areas) {
  sum(values(areas * ifel(binary_r == 1, 1, 0)), na.rm = TRUE)
}

aus_total_area              <- sum(values(cell_areas * !is.na(pred_aus_maxent)), na.rm = TRUE)
maxent_suitable_area        <- calc_area(maxent_binary)
maxent_suitable_area_masked <- calc_area(ifel(!is.na(maxent_binary_masked) & maxent_binary_masked == 1, 1, 0))
brt_suitable_area           <- calc_area(brt_binary)
brt_suitable_area_masked    <- calc_area(ifel(!is.na(brt_binary_masked) & brt_binary_masked == 1, 1, 0))
climex_suitable_area        <- calc_area(climex_binary_map)
ensemble_suitable_area      <- calc_area(ensemble_binary)
ensemble_strict_area        <- calc_area(ensemble_strict)

# Area by committee vote level
area_1plus <- calc_area(ifel(committee_vote >= 1, 1, 0))
area_2plus  <- calc_area(ifel(committee_vote >= 2, 1, 0))
area_3      <- calc_area(ifel(committee_vote == 3, 1, 0))

cat("--- Suitable Habitat Area in Australia ---\n")
sprintf("Total land area: %.0f km2", aus_total_area)
sprintf("")
sprintf("MaxEnt — Suitable: %.0f km2 (%.1f%%)",
        maxent_suitable_area, 100 * maxent_suitable_area / aus_total_area)
sprintf("MaxEnt — Suitable (MESS-masked, correlative only): %.0f km2 (%.1f%%)",
        maxent_suitable_area_masked, 100 * maxent_suitable_area_masked / aus_total_area)
sprintf("")
sprintf("BRT — Suitable: %.0f km2 (%.1f%%)",
        brt_suitable_area, 100 * brt_suitable_area / aus_total_area)
sprintf("BRT — Suitable (MESS-masked): %.0f km2 (%.1f%%)",
        brt_suitable_area_masked, 100 * brt_suitable_area_masked / aus_total_area)
sprintf("")
sprintf("CLIMEX — Suitable (EI > 10): %.0f km2 (%.1f%%)",
        climex_suitable_area, 100 * climex_suitable_area / aus_total_area)
sprintf("")
sprintf("--- Three-model Ensemble ---")
sprintf("≥1 model agrees suitable: %.0f km2 (%.1f%%)",
        area_1plus, 100 * area_1plus / aus_total_area)
sprintf("≥2 models agree suitable: %.0f km2 (%.1f%%) [recommended threshold]",
        area_2plus, 100 * area_2plus / aus_total_area)
sprintf("All 3 models agree suitable: %.0f km2 (%.1f%%) [conservative]",
        area_3, 100 * area_3 / aus_total_area)
```

## 9. Visualisation — Australia suitability maps

```{r}
#| label: maps
#| fig-width: 14
#| fig-height: 21

aus_sf <- ne_countries(
  country = "Australia",
  scale = "medium",
  returnclass = "sf"
)
states <- ne_states(country = "Australia", returnclass = "sf")

# Convert rasters to data frames for ggplot
maxent_df <- as.data.frame(pred_aus_maxent, xy = TRUE) %>%
  filter(!is.na(maxent_suitability))
brt_df <- as.data.frame(pred_aus_brt, xy = TRUE) %>%
  filter(!is.na(brt_suitability))
climex_df <- as.data.frame(climex_EI_rs, xy = TRUE) %>%
  filter(!is.na(climex_suitability))
ensemble_df <- as.data.frame(pred_aus_ensemble, xy = TRUE) %>%
  filter(!is.na(ensemble_suitability))
mess_df <- as.data.frame(mess_aus, xy = TRUE) %>% filter(!is.na(mess))
vote_df <- as.data.frame(committee_vote, xy = TRUE) %>%
  filter(!is.na(models_agreeing)) %>%
  mutate(
    models_agreeing = factor(
      models_agreeing,
      levels = 0:3,
      labels = c("0 — None", "1 — One model", "2 — Two models", "3 — All three")
    )
  )

# Common theme
map_theme <- theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(size = 12, face = "bold"),
    legend.position = "bottom",
    legend.key.width = unit(1.8, "cm"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )

# A) MaxEnt
p1 <- ggplot() +
  geom_raster(data = maxent_df, aes(x = x, y = y, fill = maxent_suitability)) +
  geom_sf(data = states, fill = NA, colour = "grey40", linewidth = 0.3) +
  geom_sf(data = aus_sf, fill = NA, colour = "black", linewidth = 0.5) +
  scale_fill_gradientn(
    colours = c("white", magma(256, direction = -1)),
    limits = c(0, 1),
    name = "Suitability",
    trans = "sqrt"
  ) +
  labs(title = "A) MaxEnt (correlative)") +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10)) +
  map_theme

# B) BRT
p2 <- ggplot() +
  geom_raster(data = brt_df, aes(x = x, y = y, fill = brt_suitability)) +
  geom_sf(data = states, fill = NA, colour = "grey40", linewidth = 0.3) +
  geom_sf(data = aus_sf, fill = NA, colour = "black", linewidth = 0.5) +
  scale_fill_gradientn(
    colours = c("white", magma(256, direction = -1)),
    limits = c(0, 1),
    name = "Suitability",
    trans = "sqrt"
  ) +
  labs(title = "B) BRT (correlative)") +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10)) +
  map_theme

# C) CLIMEX EI
p3 <- ggplot() +
  geom_raster(data = climex_df, aes(x = x, y = y, fill = climex_suitability)) +
  geom_sf(data = states, fill = NA, colour = "grey40", linewidth = 0.3) +
  geom_sf(data = aus_sf, fill = NA, colour = "black", linewidth = 0.5) +
  scale_fill_gradientn(
    colours = c("white", magma(256, direction = -1)),
    limits = c(0, 1),
    name = "Suitability",
    trans = "sqrt"
  ) +
  labs(
    title = "C) CLIMEX Ecoclimatic Index (process-based)",
    subtitle = "Not subject to MESS extrapolation limitation"
  ) +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10)) +
  map_theme

# D) Continuous ensemble (equal-weight mean of all three)
p4 <- ggplot() +
  geom_raster(
    data = ensemble_df,
    aes(x = x, y = y, fill = ensemble_suitability)
  ) +
  geom_sf(data = states, fill = NA, colour = "grey40", linewidth = 0.3) +
  geom_sf(data = aus_sf, fill = NA, colour = "black", linewidth = 0.5) +
  scale_fill_gradientn(
    colours = c("white", magma(256, direction = -1)),
    limits = c(0, 1),
    name = "Suitability",
    trans = "sqrt"
  ) +
  labs(title = "D) Ensemble mean (MaxEnt + BRT + CLIMEX)") +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10)) +
  map_theme

# E) Committee vote
p5 <- ggplot() +
  geom_raster(data = vote_df, aes(x = x, y = y, fill = models_agreeing)) +
  geom_sf(data = states, fill = NA, colour = "grey40", linewidth = 0.3) +
  geom_sf(data = aus_sf, fill = NA, colour = "black", linewidth = 0.5) +
  scale_fill_manual(
    values = c(
      "0 — None" = "grey92",
      "1 — One model" = "#FEE391",
      "2 — Two models" = "#EC7014",
      "3 — All three" = "#7F0000"
    ),
    name = "Models agreeing"
  ) +
  labs(
    title = "E) Committee vote — model agreement",
    subtitle = "Darker = more models predict suitable"
  ) +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10)) +
  map_theme

# F) MESS map
p6 <- ggplot() +
  geom_raster(data = mess_df, aes(x = x, y = y, fill = mess)) +
  geom_sf(data = states, fill = NA, colour = "grey40", linewidth = 0.3) +
  geom_sf(data = aus_sf, fill = NA, colour = "black", linewidth = 0.5) +
  scale_fill_gradient2(
    low = "#D55E00",
    mid = "white",
    high = "#0072B2",
    midpoint = 0,
    name = "MESS"
  ) +
  labs(
    title = "F) MESS — Environmental Similarity",
    subtitle = "Red = novel climate | Blue = analogous (correlative models only)"
  ) +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10)) +
  map_theme

library(patchwork)
combined_plot <- (p1 + p2) / (p3 + p4) / (p5 + p6)

combined_plot +
  plot_annotation(
    title = expression(
      italic("Dermacentor variabilis") ~
        "— Three-Model Projected Climatic Suitability in Australia"
    ),
    theme = theme(plot.title = element_text(size = 15, face = "bold"))
  )

ggsave(
  file.path(figures_dir, "07_australia_projection.png"),
  width = 14,
  height = 21,
  dpi = 300
)
```

## 10. Save outputs

```{r}
#| label: save

# Continuous predictions (individual models)
writeRaster(pred_aus_maxent, file.path(processed_dir, "pred_aus_maxent.tif"), overwrite = TRUE)
writeRaster(pred_aus_brt,    file.path(processed_dir, "pred_aus_brt.tif"),    overwrite = TRUE)
writeRaster(climex_EI_rs,    file.path(processed_dir, "climex_EI_norm_aus.tif"), overwrite = TRUE)

# Three-model ensemble
writeRaster(pred_aus_ensemble, file.path(processed_dir, "pred_aus_ensemble.tif"), overwrite = TRUE)
writeRaster(committee_vote,    file.path(processed_dir, "committee_vote_aus.tif"), overwrite = TRUE)

# Binary predictions
writeRaster(maxent_binary,      file.path(processed_dir, "pred_aus_maxent_binary.tif"), overwrite = TRUE)
writeRaster(brt_binary,         file.path(processed_dir, "pred_aus_brt_binary.tif"),    overwrite = TRUE)
writeRaster(ensemble_binary,    file.path(processed_dir, "pred_aus_ensemble_binary.tif"), overwrite = TRUE)
writeRaster(ensemble_strict,    file.path(processed_dir, "pred_aus_ensemble_strict.tif"), overwrite = TRUE)
writeRaster(climex_only_novel,  file.path(processed_dir, "climex_only_novel_aus.tif"), overwrite = TRUE)

# MESS
writeRaster(mess_aus, file.path(processed_dir, "mess_australia.tif"), overwrite = TRUE)

# Native range predictions
writeRaster(pred_M_maxent, file.path(processed_dir, "pred_M_maxent.tif"), overwrite = TRUE)
writeRaster(pred_M_brt,    file.path(processed_dir, "pred_M_brt.tif"),    overwrite = TRUE)

# Area summary (comprehensive)
area_summary <- tibble(
  model = c(
    "MaxEnt (correlative)",
    "MaxEnt (MESS-masked)",
    "BRT (correlative)",
    "BRT (MESS-masked)",
    "CLIMEX (EI > 10, process-based)",
    "Ensemble ≥1 model",
    "Ensemble ≥2 models (recommended)",
    "Ensemble = 3 models (conservative)"
  ),
  suitable_area_km2 = round(c(
    maxent_suitable_area, maxent_suitable_area_masked,
    brt_suitable_area,    brt_suitable_area_masked,
    climex_suitable_area,
    area_1plus, area_2plus, area_3
  ), 0),
  pct_australia = round(c(
    maxent_suitable_area, maxent_suitable_area_masked,
    brt_suitable_area,    brt_suitable_area_masked,
    climex_suitable_area,
    area_1plus, area_2plus, area_3
  ) / aus_total_area * 100, 1)
)
write_csv(area_summary, file.path(processed_dir, "suitable_area_summary.csv"))

cat("Area summary:\n")
print(area_summary, n = Inf)

sprintf("Saved all projection outputs, three-model ensemble, and area summaries")
```
