---
title: "Step 7c: Improved Ensemble and Model Agreement"
subtitle: "*Dermacentor variabilis* Species Distribution Model"
author: "Alexander W. Gofton"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme:
      light: flatly
      dark: solar
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    results: hold
    cache.lazy: false
---

## Overview

The original ensemble (simple mean of MaxEnt and BRT) is not meaningful when
models disagree by 7x (56.4% vs 8.4%). This script produces improved ensemble
products using the **sensitivity model results** (Set B and C variable sets):

1. **Conservative (intersection) ensemble** — suitable only where BOTH MaxEnt and BRT agree
2. **Model disagreement map** — spatial uncertainty indicator
3. **MESS-masked conservative ensemble** — intersection further masked to analogous climate
4. **Multi-evidence synthesis** — areas where SDMs, CLIMEX constraints, and climate
   analogues converge

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
```

## 1. Load predictions and thresholds

```{r}
#| label: load

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/SDM/Dvariabilis_SDM"
processed_dir <- file.path(base_dir, "processed_data")
figures_dir <- file.path(base_dir, "figures")

aus <- ne_countries(country = "Australia", scale = "medium", returnclass = "sf")

# Original (Set A) predictions
pred_A_maxent <- rast(file.path(processed_dir, "pred_aus_maxent.tif"))
pred_A_brt <- rast(file.path(processed_dir, "pred_aus_brt.tif"))
thresh_A <- readRDS(file.path(processed_dir, "model_thresholds.rds"))

# Sensitivity predictions (Set B and C)
# Check which sensitivity outputs exist
has_setB <- file.exists(file.path(processed_dir, "pred_aus_maxent_setB.tif"))
has_setC <- file.exists(file.path(processed_dir, "pred_aus_maxent_setC.tif"))

if (has_setB) {
  pred_B_maxent <- rast(file.path(processed_dir, "pred_aus_maxent_setB.tif"))
  pred_B_brt <- rast(file.path(processed_dir, "pred_aus_brt_setB.tif"))
  thresh_B <- readRDS(file.path(processed_dir, "sensitivity_thresholds.rds"))$setB
}

if (has_setC) {
  pred_C_maxent <- rast(file.path(processed_dir, "pred_aus_maxent_setC.tif"))
  pred_C_brt <- rast(file.path(processed_dir, "pred_aus_brt_setC.tif"))
  thresh_C <- readRDS(file.path(processed_dir, "sensitivity_thresholds.rds"))$setC
}

# CLIMEX constraints (if available)
has_climex <- file.exists(file.path(processed_dir, "climex_suitable_aus.tif"))
if (has_climex) {
  climex_suitable <- rast(file.path(processed_dir, "climex_suitable_aus.tif"))
}

# Mahalanobis distance (if available)
has_mahal <- file.exists(file.path(processed_dir, "mahalanobis_distance_aus.tif"))
if (has_mahal) {
  mahal_dist <- rast(file.path(processed_dir, "mahalanobis_distance_aus.tif"))
  chi_sq_95 <- qchisq(0.95, df = length(readLines(
    file.path(processed_dir, "selected_variables_setC.txt"))))
  climate_analogue <- mahal_dist <= chi_sq_95
}

cat("Available layers:\n")
cat(sprintf("  Set A (original): %s\n", "YES"))
cat(sprintf("  Set B (no Bio1): %s\n", ifelse(has_setB, "YES", "NO — run 05c first")))
cat(sprintf("  Set C (moisture): %s\n", ifelse(has_setC, "YES", "NO — run 05c first")))
cat(sprintf("  CLIMEX constraints: %s\n", ifelse(has_climex, "YES", "NO — run 10c first")))
cat(sprintf("  Climate analogue: %s\n", ifelse(has_mahal, "YES", "NO — run 10 first")))
```

## 2. Conservative (intersection) ensembles

```{r}
#| label: intersection

cell_areas <- cellSize(pred_A_maxent, unit = "km")
aus_total <- sum(values(cell_areas * !is.na(pred_A_maxent)), na.rm = TRUE)

calc_area_pct <- function(binary_raster) {
  area <- sum(values(cell_areas * binary_raster), na.rm = TRUE)
  c(area_km2 = area, pct = 100 * area / aus_total)
}

# Set A intersection
A_maxent_bin <- pred_A_maxent >= thresh_A$maxent$maxss
A_brt_bin <- pred_A_brt >= thresh_A$brt$maxss
A_intersection <- A_maxent_bin & A_brt_bin
A_union <- A_maxent_bin | A_brt_bin

# Set B intersection (if available)
if (has_setB) {
  B_maxent_bin <- pred_B_maxent >= thresh_B$maxent$threshold
  B_brt_bin <- pred_B_brt >= thresh_B$brt$threshold
  B_intersection <- B_maxent_bin & B_brt_bin
  B_union <- B_maxent_bin | B_brt_bin
}

# Set C intersection (if available)
if (has_setC) {
  C_maxent_bin <- pred_C_maxent >= thresh_C$maxent$threshold
  C_brt_bin <- pred_C_brt >= thresh_C$brt$threshold
  C_intersection <- C_maxent_bin & C_brt_bin
  C_union <- C_maxent_bin | C_brt_bin
}

# Report areas
cat("--- Intersection (Conservative) Ensemble Areas ---\n\n")
a <- calc_area_pct(A_intersection)
sprintf("Set A intersection: %.0f km2 (%.1f%%)", a[1], a[2])
a <- calc_area_pct(A_union)
sprintf("Set A union: %.0f km2 (%.1f%%)", a[1], a[2])

if (has_setB) {
  a <- calc_area_pct(B_intersection)
  sprintf("Set B intersection: %.0f km2 (%.1f%%)", a[1], a[2])
}
if (has_setC) {
  a <- calc_area_pct(C_intersection)
  sprintf("Set C intersection: %.0f km2 (%.1f%%)", a[1], a[2])
}
```

## 3. Model disagreement maps

```{r}
#| label: disagreement
#| fig-width: 14
#| fig-height: 10

# Continuous disagreement: |MaxEnt - BRT| for each set
disagree_A <- abs(pred_A_maxent - pred_A_brt)
names(disagree_A) <- "disagreement"

panels <- list(
  list(rast = disagree_A, label = "Set A (with Bio1)")
)

if (has_setB) {
  disagree_B <- abs(pred_B_maxent - pred_B_brt)
  panels <- c(panels, list(list(rast = disagree_B, label = "Set B (no Bio1)")))
}
if (has_setC) {
  disagree_C <- abs(pred_C_maxent - pred_C_brt)
  panels <- c(panels, list(list(rast = disagree_C, label = "Set C (moisture-focused)")))
}

# Build plot data
disagree_plot <- list()
for (p in panels) {
  df <- as.data.frame(p$rast, xy = TRUE, na.rm = TRUE)
  names(df)[3] <- "disagreement"
  df$set <- p$label
  disagree_plot <- c(disagree_plot, list(df))
}
disagree_df <- bind_rows(disagree_plot)

p <- ggplot() +
  geom_raster(data = disagree_df, aes(x = x, y = y, fill = disagreement)) +
  geom_sf(data = aus, fill = NA, colour = "black", linewidth = 0.3) +
  facet_wrap(~set, ncol = length(panels)) +
  scale_fill_viridis_c(
    name = "Model\nDisagreement\n|MaxEnt - BRT|",
    option = "plasma",
    limits = c(0, 0.8)
  ) +
  labs(
    title = expression(italic("D. variabilis") ~ "— MaxEnt vs BRT Prediction Disagreement"),
    subtitle = "Lower values = better model agreement; yellow/bright = high disagreement",
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold"), legend.position = "bottom") +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10))

print(p)

ggsave(file.path(figures_dir, "07c_model_disagreement.png"),
       p, width = 14, height = 6, dpi = 300)

# Report mean disagreement
cat("Mean absolute disagreement across Australia:\n")
sprintf("  Set A: %.3f", mean(values(disagree_A, na.rm = TRUE)))
if (has_setB) sprintf("  Set B: %.3f", mean(values(disagree_B, na.rm = TRUE)))
if (has_setC) sprintf("  Set C: %.3f", mean(values(disagree_C, na.rm = TRUE)))
```

## 4. Multi-evidence synthesis

Combine all available evidence sources to identify areas of highest confidence.

```{r}
#| label: synthesis
#| fig-width: 10
#| fig-height: 8

# Use the best available sensitivity set for the synthesis
if (has_setC) {
  best_intersection <- C_intersection
  best_label <- "Set C"
} else if (has_setB) {
  best_intersection <- B_intersection
  best_label <- "Set B"
} else {
  best_intersection <- A_intersection
  best_label <- "Set A"
}

# Count evidence layers (how many independent methods agree on suitability)
evidence_count <- best_intersection * 1  # SDM intersection = 1

if (has_climex) {
  # Resample if needed
  if (!compareGeom(climex_suitable, best_intersection, stopOnError = FALSE)) {
    climex_r <- resample(climex_suitable, best_intersection, method = "near")
  } else {
    climex_r <- climex_suitable
  }
  evidence_count <- evidence_count + climex_r
}

if (has_mahal) {
  if (!compareGeom(climate_analogue, best_intersection, stopOnError = FALSE)) {
    analogue_r <- resample(climate_analogue, best_intersection, method = "near")
  } else {
    analogue_r <- climate_analogue
  }
  evidence_count <- evidence_count + analogue_r
}

names(evidence_count) <- "n_evidence"

# Map evidence count
ev_df <- as.data.frame(evidence_count, xy = TRUE, na.rm = TRUE)

max_evidence <- max(values(evidence_count, na.rm = TRUE))
evidence_labels <- c("0" = "No evidence", "1" = "1 method",
                     "2" = "2 methods", "3" = "All 3 methods")

p_evidence <- ggplot() +
  geom_raster(data = ev_df, aes(x = x, y = y, fill = factor(n_evidence))) +
  geom_sf(data = aus, fill = NA, colour = "black", linewidth = 0.3) +
  scale_fill_manual(
    values = c("0" = "grey90", "1" = "#56B4E9", "2" = "#E69F00", "3" = "#D55E00"),
    labels = evidence_labels[as.character(0:max_evidence)],
    name = "Lines of\nEvidence"
  ) +
  labs(
    title = expression(italic("D. variabilis") ~ "— Multi-Evidence Suitability Synthesis"),
    subtitle = sprintf("SDM intersection (%s) + CLIMEX constraints + Climate analogue",
                       best_label),
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right") +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10))

print(p_evidence)

ggsave(file.path(figures_dir, "07c_multi_evidence_synthesis.png"),
       p_evidence, width = 10, height = 8, dpi = 300)

# Report
cat("\n--- Multi-Evidence Summary ---\n")
for (i in 0:max_evidence) {
  n <- sum(values(evidence_count, na.rm = TRUE) == i)
  a <- sum(values(cell_areas * (evidence_count == i), na.rm = TRUE))
  sprintf("  %d methods agree: %.0f km2 (%.1f%%)", i, a, 100 * a / aus_total)
}
```

## 5. Summary area comparison table

```{r}
#| label: final-summary

area_table <- tibble(
  Approach = character(),
  Area_km2 = numeric(),
  Pct_Australia = numeric()
)

add_row <- function(tbl, label, rast) {
  a <- calc_area_pct(rast)
  bind_rows(tbl, tibble(Approach = label, Area_km2 = round(a[1]),
                        Pct_Australia = round(a[2], 1)))
}

area_table <- add_row(area_table, "Set A: MaxEnt only", A_maxent_bin)
area_table <- add_row(area_table, "Set A: BRT only", A_brt_bin)
area_table <- add_row(area_table, "Set A: Intersection", A_intersection)

if (has_setB) {
  area_table <- add_row(area_table, "Set B: MaxEnt only", B_maxent_bin)
  area_table <- add_row(area_table, "Set B: BRT only", B_brt_bin)
  area_table <- add_row(area_table, "Set B: Intersection", B_intersection)
}
if (has_setC) {
  area_table <- add_row(area_table, "Set C: MaxEnt only", C_maxent_bin)
  area_table <- add_row(area_table, "Set C: BRT only", C_brt_bin)
  area_table <- add_row(area_table, "Set C: Intersection", C_intersection)
}
if (has_climex) {
  area_table <- add_row(area_table, "CLIMEX constraints", climex_r)
}
if (has_mahal) {
  area_table <- add_row(area_table, "Climate analogue (95% CI)", analogue_r)
}

# Multi-evidence highest confidence
if (max_evidence >= 2) {
  high_conf <- evidence_count >= 2
  area_table <- add_row(area_table, "Multi-evidence (>=2 methods)", high_conf)
}
if (max_evidence >= 3) {
  highest_conf <- evidence_count >= 3
  area_table <- add_row(area_table, "Multi-evidence (all 3 methods)", highest_conf)
}

cat("======================================================\n")
cat("  COMPREHENSIVE AREA COMPARISON\n")
cat("======================================================\n\n")
print(area_table, n = Inf)

write_csv(area_table, file.path(processed_dir, "comprehensive_area_comparison.csv"))

# Save synthesis raster
writeRaster(evidence_count,
            file.path(processed_dir, "multi_evidence_count_aus.tif"),
            overwrite = TRUE)

sprintf("Saved all ensemble and synthesis outputs")
```
