---
title: "Step 8b: Niche Analysis — Variable Set Sensitivity"
subtitle: "*Dermacentor variabilis* Species Distribution Model"
author: "Alexander W. Gofton"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme:
      light: flatly
      dark: solar
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    results: hold
    cache.lazy: false
---

## Overview

The original niche analysis (Step 8) used all 8 variables including Bio1 and Bio7,
yielding low niche overlap (Schoener's D = 0.233) and borderline similarity tests
(p = 0.078, 0.064). This script repeats the analysis using the moisture-focused
variable set (Set C) to test whether cold-tolerance variables diluted the overlap signal.

**Hypothesis:** Removing cold-tolerance variables (Bio1, Bio7) increases Schoener's D
and renders the similarity test significant, because the remaining variables
(moisture, warm-season temperature) better capture the ecological gradients
relevant to both continents.

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(ecospat)
library(ade4)
library(sf)
library(rnaturalearth)
```

## 1. Load data

```{r}
#| label: load

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/SDM/Dvariabilis_SDM"
processed_dir <- file.path(base_dir, "processed_data")
figures_dir <- file.path(base_dir, "figures")

# Full environmental stacks (all candidates)
env_stack_M <- rast(file.path(processed_dir, "env_stack_M.tif"))
env_stack_aus <- rast(file.path(processed_dir, "env_stack_aus.tif"))

# Occurrence data
occ_env <- readRDS(file.path(processed_dir, "occurrences_with_env.rds"))

# Variable sets
set_A <- readLines(file.path(processed_dir, "selected_variables.txt"))

# Set C: moisture-focused (from 04b)
set_C_file <- file.path(processed_dir, "selected_variables_setC.txt")
if (file.exists(set_C_file)) {
  set_C <- readLines(set_C_file)
} else {
  set_C <- c("bio5", "bio10", "bio12", "bio15",
             "aridity_idx", "PET_seasonality", "vpd_annual")
  set_C <- intersect(set_C, names(env_stack_M))
}

cat("Set A (original):", paste(set_A, collapse = ", "), "\n")
cat("Set C (moisture):", paste(set_C, collapse = ", "), "\n")
```

## 2. Function: complete niche analysis for a variable set

```{r}
#| label: niche-function

run_niche_analysis <- function(env_M, env_aus, occ_data, vars, set_name, seed = 42) {
  cat(sprintf("\n=== Niche Analysis for %s ===\n", set_name))

  # Subset to selected variables
  env_M_sub <- env_M[[vars]]
  env_aus_sub <- env_aus[[vars]]

  # Sample backgrounds
  set.seed(seed)
  bg_M <- spatSample(env_M_sub, size = 10000, method = "random",
                     na.rm = TRUE, as.df = TRUE, xy = TRUE)
  bg_aus <- spatSample(env_aus_sub, size = 10000, method = "random",
                       na.rm = TRUE, as.df = TRUE, xy = TRUE)

  # PCA on combined environmental space
  all_env <- bind_rows(
    bg_M %>% dplyr::select(all_of(vars)) %>% mutate(region = "Native"),
    bg_aus %>% dplyr::select(all_of(vars)) %>% mutate(region = "Australia")
  )

  pca_input <- all_env %>% dplyr::select(all_of(vars))
  pca_result <- dudi.pca(pca_input, scannf = FALSE, nf = 2)

  var_explained <- round(100 * pca_result$eig / sum(pca_result$eig), 1)
  cat(sprintf("  PC1: %.1f%% | PC2: %.1f%% | Cumulative: %.1f%%\n",
              var_explained[1], var_explained[2], sum(var_explained[1:2])))

  # Project occurrences
  occ_pca <- occ_data %>% dplyr::select(all_of(vars))
  occ_scores <- suprow(pca_result, occ_pca)$li

  n_M <- nrow(bg_M)
  scores_M <- pca_result$li[1:n_M, ]
  scores_aus <- pca_result$li[(n_M + 1):nrow(pca_result$li), ]

  # Ecospat grids
  grid_native <- ecospat.grid.clim.dyn(
    glob = pca_result$li, glob1 = scores_M, sp = occ_scores, R = 100
  )
  grid_aus <- ecospat.grid.clim.dyn(
    glob = pca_result$li, glob1 = scores_aus, sp = scores_aus, R = 100
  )

  # Niche overlap
  overlap <- ecospat.niche.overlap(grid_native, grid_aus, cor = TRUE)
  cat(sprintf("  Schoener's D: %.3f | Warren's I: %.3f\n", overlap$D, overlap$I))

  # Equivalency test
  set.seed(seed)
  equiv <- ecospat.niche.equivalency.test(grid_native, grid_aus, rep = 1000)
  cat(sprintf("  Equivalency p-value: %.4f\n", equiv$p.D))

  # Similarity tests
  set.seed(seed)
  sim_1 <- ecospat.niche.similarity.test(grid_native, grid_aus,
                                          rep = 1000, rand.type = 2)
  set.seed(seed)
  sim_2 <- ecospat.niche.similarity.test(grid_aus, grid_native,
                                          rep = 1000, rand.type = 2)
  cat(sprintf("  Similarity (Native -> Aus) p-value: %.4f\n", sim_1$p.D))
  cat(sprintf("  Similarity (Aus -> Native) p-value: %.4f\n", sim_2$p.D))

  list(
    overlap = overlap,
    equiv = equiv,
    sim_native_to_aus = sim_1,
    sim_aus_to_native = sim_2,
    pca = pca_result,
    var_explained = var_explained,
    grid_native = grid_native,
    grid_aus = grid_aus,
    occ_scores = occ_scores,
    scores_M = scores_M,
    scores_aus = scores_aus
  )
}
```

## 3. Run niche analysis for both variable sets

```{r}
#| label: run-analyses
#| cache: true

niche_A <- run_niche_analysis(env_stack_M, env_stack_aus, occ_env, set_A, "Set A (original)")
niche_C <- run_niche_analysis(env_stack_M, env_stack_aus, occ_env, set_C, "Set C (moisture)")
```

## 4. Comparison table

```{r}
#| label: comparison

comparison <- tibble(
  Metric = c("Schoener's D", "Warren's I",
             "Equivalency p-value",
             "Similarity (Native -> Aus) p",
             "Similarity (Aus -> Native) p",
             "PC1 variance (%)", "PC2 variance (%)"),
  Set_A = c(
    round(niche_A$overlap$D, 3),
    round(niche_A$overlap$I, 3),
    round(niche_A$equiv$p.D, 4),
    round(niche_A$sim_native_to_aus$p.D, 4),
    round(niche_A$sim_aus_to_native$p.D, 4),
    niche_A$var_explained[1],
    niche_A$var_explained[2]
  ),
  Set_C = c(
    round(niche_C$overlap$D, 3),
    round(niche_C$overlap$I, 3),
    round(niche_C$equiv$p.D, 4),
    round(niche_C$sim_native_to_aus$p.D, 4),
    round(niche_C$sim_aus_to_native$p.D, 4),
    niche_C$var_explained[1],
    niche_C$var_explained[2]
  )
)

cat("======================================================\n")
cat("  NICHE ANALYSIS COMPARISON\n")
cat("======================================================\n\n")
print(comparison, n = Inf)

D_change <- niche_C$overlap$D - niche_A$overlap$D
cat(sprintf("\nSchoener's D change: %.3f -> %.3f (%+.3f)\n",
            niche_A$overlap$D, niche_C$overlap$D, D_change))

if (D_change > 0) {
  cat("FINDING: Removing cold-tolerance variables INCREASES niche overlap.\n")
  cat("This supports the hypothesis that Bio1/Bio7 diluted the similarity signal.\n")
} else {
  cat("NOTE: Removing cold-tolerance variables did not increase overlap.\n")
  cat("The low overlap may reflect genuine climatic differences in moisture/temperature.\n")
}

if (niche_C$sim_native_to_aus$p.D < 0.05 & niche_A$sim_native_to_aus$p.D >= 0.05) {
  cat("\nCRITICAL: Similarity test becomes significant with Set C variables.\n")
  cat("Australia's climate IS more similar to the native range than expected by chance\n")
  cat("when assessed using ecologically appropriate variables.\n")
}
```

## 5. PCA niche comparison plots

```{r}
#| label: niche-plots
#| fig-width: 14
#| fig-height: 12

png(file.path(figures_dir, "08b_niche_comparison.png"),
    width = 14, height = 12, units = "in", res = 300)
par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))

# Set A plots
ecospat.plot.niche(niche_A$grid_native, title = "Set A: Native Niche",
                   name.axis1 = "PC1", name.axis2 = "PC2")
ecospat.plot.niche(niche_A$grid_aus, title = "Set A: Australia",
                   name.axis1 = "PC1", name.axis2 = "PC2")
ecospat.plot.niche.dyn(niche_A$grid_native, niche_A$grid_aus,
                       quant = 0.25, interest = 2,
                       title = sprintf("Set A: Overlap (D=%.3f)", niche_A$overlap$D),
                       name.axis1 = "PC1", name.axis2 = "PC2")

# Set C plots
ecospat.plot.niche(niche_C$grid_native, title = "Set C: Native Niche",
                   name.axis1 = "PC1", name.axis2 = "PC2")
ecospat.plot.niche(niche_C$grid_aus, title = "Set C: Australia",
                   name.axis1 = "PC1", name.axis2 = "PC2")
ecospat.plot.niche.dyn(niche_C$grid_native, niche_C$grid_aus,
                       quant = 0.25, interest = 2,
                       title = sprintf("Set C: Overlap (D=%.3f)", niche_C$overlap$D),
                       name.axis1 = "PC1", name.axis2 = "PC2")
dev.off()

# Display inline
par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))
ecospat.plot.niche(niche_A$grid_native, title = "Set A: Native Niche",
                   name.axis1 = "PC1", name.axis2 = "PC2")
ecospat.plot.niche(niche_A$grid_aus, title = "Set A: Australia",
                   name.axis1 = "PC1", name.axis2 = "PC2")
ecospat.plot.niche.dyn(niche_A$grid_native, niche_A$grid_aus,
                       quant = 0.25, interest = 2,
                       title = sprintf("Set A: Overlap (D=%.3f)", niche_A$overlap$D),
                       name.axis1 = "PC1", name.axis2 = "PC2")
ecospat.plot.niche(niche_C$grid_native, title = "Set C: Native Niche",
                   name.axis1 = "PC1", name.axis2 = "PC2")
ecospat.plot.niche(niche_C$grid_aus, title = "Set C: Australia",
                   name.axis1 = "PC1", name.axis2 = "PC2")
ecospat.plot.niche.dyn(niche_C$grid_native, niche_C$grid_aus,
                       quant = 0.25, interest = 2,
                       title = sprintf("Set C: Overlap (D=%.3f)", niche_C$overlap$D),
                       name.axis1 = "PC1", name.axis2 = "PC2")
```

## 6. Save outputs

```{r}
#| label: save

write_csv(comparison, file.path(processed_dir, "niche_sensitivity_comparison.csv"))
saveRDS(niche_C, file.path(processed_dir, "niche_analysis_setC.rds"))

sprintf("Saved niche sensitivity analysis outputs")
```
