---
title: "Step 8c: Niche Truncation Assessment"
subtitle: "*Dermacentor variabilis* Species Distribution Model"
author: "Alexander W. Gofton"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme:
      light: flatly
      dark: solar
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    results: hold
    cache.lazy: false
---

## Overview

Niche truncation occurs when a species' response curve is still rising or
remains at maximum suitability at the boundary of available environmental
conditions in the training region. When truncated, the correlative model cannot
estimate the true niche boundary — it has never "seen" where the species stops
occurring along that gradient.

For cross-continental projection, niche truncation means the model's response
may be completely wrong beyond the training range. This is particularly relevant
for *D. variabilis*:

- **Warm-season variables (Bio5, Bio10):** If the species occurs at the warmest
  available conditions in eastern N. America, we cannot determine its upper
  thermal limit from correlative data alone.
- **Moisture variables (aridity_idx, PET_seasonality):** If the species avoids
  the driest conditions available, this validates humidity limitation.
- **Cold variables (Bio1, Bio7):** If the response is highest at intermediate
  values, the cold-end decline encodes winter severity — irrelevant to Australia.

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(dismo)
library(gbm)
```

## 1. Load data

```{r}
#| label: load

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/SDM/Dvariabilis_SDM"
processed_dir <- file.path(base_dir, "processed_data")
figures_dir <- file.path(base_dir, "figures")

maxent_model <- readRDS(file.path(processed_dir, "selected_maxent_model.rds"))
brt_model <- readRDS(file.path(processed_dir, "brt_model.rds"))
env_M <- rast(file.path(processed_dir, "env_selected_M.tif"))
occ_env <- readRDS(file.path(processed_dir, "occurrences_with_env.rds"))
selected_vars <- readLines(file.path(processed_dir, "selected_variables.txt"))
```

## 2. Generate response curves and available environment distributions

For each variable:
- Histogram of available environmental conditions in the training region (M)
- Histogram of conditions at occurrence points
- MaxEnt and BRT response curves overlaid

```{r}
#| label: truncation-analysis
#| fig-width: 14
#| fig-height: 16

# Extract environmental values
occ_vals <- occ_env %>% dplyr::select(all_of(selected_vars))
mean_occ_vals <- as.data.frame(t(colMeans(occ_vals, na.rm = TRUE)))

# Available environment in M (sample for plotting)
set.seed(42)
bg_sample <- spatSample(env_M, size = 50000, method = "random",
                        na.rm = TRUE, as.df = TRUE)

# Generate response curves and truncation metrics
trunc_results <- list()
plot_data <- list()

for (v in selected_vars) {
  # Available range in training region
  avail_range <- range(bg_sample[[v]], na.rm = TRUE)
  occ_range <- range(occ_vals[[v]], na.rm = TRUE)

  # Response curve across training range
  var_seq <- seq(avail_range[1], avail_range[2], length.out = 200)
  pred_df <- mean_occ_vals[rep(1, 200), ]
  pred_df[[v]] <- var_seq

  maxent_pred <- predict(maxent_model, pred_df, type = "cloglog")
  brt_pred <- predict(brt_model, as.data.frame(pred_df),
                      n.trees = brt_model$n.trees, type = "response")

  # Truncation assessment: check if response is still elevated at range edges
  # Check the last 10% of the range at each end
  n_edge <- 20  # points in the edge zone
  low_edge_maxent <- mean(maxent_pred[1:n_edge])
  high_edge_maxent <- mean(maxent_pred[(200 - n_edge + 1):200])
  low_edge_brt <- mean(brt_pred[1:n_edge])
  high_edge_brt <- mean(brt_pred[(200 - n_edge + 1):200])
  peak_maxent <- max(maxent_pred)
  peak_brt <- max(brt_pred)

  # Truncation = edge suitability > 50% of peak suitability
  trunc_low_maxent <- low_edge_maxent > 0.5 * peak_maxent
  trunc_high_maxent <- high_edge_maxent > 0.5 * peak_maxent
  trunc_low_brt <- low_edge_brt > 0.5 * peak_brt
  trunc_high_brt <- high_edge_brt > 0.5 * peak_brt

  # Species occupancy at edges of available environment
  q05 <- quantile(occ_vals[[v]], 0.05, na.rm = TRUE)
  q95 <- quantile(occ_vals[[v]], 0.95, na.rm = TRUE)
  avail_q05 <- quantile(bg_sample[[v]], 0.05, na.rm = TRUE)
  avail_q95 <- quantile(bg_sample[[v]], 0.95, na.rm = TRUE)

  # Does the species occur at the extremes of available environment?
  occ_at_low_edge <- q05 <= avail_q05 + 0.05 * diff(avail_range)
  occ_at_high_edge <- q95 >= avail_q95 - 0.05 * diff(avail_range)

  trunc_results[[v]] <- tibble(
    variable = v,
    avail_min = avail_range[1],
    avail_max = avail_range[2],
    occ_q05 = as.numeric(q05),
    occ_q95 = as.numeric(q95),
    peak_maxent = peak_maxent,
    peak_brt = peak_brt,
    low_edge_maxent = round(low_edge_maxent, 3),
    high_edge_maxent = round(high_edge_maxent, 3),
    low_edge_brt = round(low_edge_brt, 3),
    high_edge_brt = round(high_edge_brt, 3),
    truncated_low = trunc_low_maxent | trunc_low_brt,
    truncated_high = trunc_high_maxent | trunc_high_brt,
    occ_at_low_extreme = occ_at_low_edge,
    occ_at_high_extreme = occ_at_high_edge
  )

  # Store plot data
  plot_data[[v]] <- list(
    response = tibble(
      value = var_seq,
      MaxEnt = maxent_pred,
      BRT = brt_pred
    ),
    available = bg_sample[[v]],
    occurrences = occ_vals[[v]]
  )
}

trunc_summary <- bind_rows(trunc_results)

# Plot: response curves overlaid on environmental availability histograms
n_vars <- length(selected_vars)
ncol_plot <- 2
nrow_plot <- ceiling(n_vars / ncol_plot)

par(mfrow = c(nrow_plot, ncol_plot), mar = c(4, 4, 3, 4), oma = c(0, 0, 2, 0))

for (v in selected_vars) {
  pd <- plot_data[[v]]
  tr <- trunc_summary %>% filter(variable == v)

  # Histogram of available environment (background)
  h <- hist(pd$available, breaks = 50, plot = FALSE)
  # Scale histogram to fit with response curves
  h_scale <- h$density / max(h$density) * 0.9

  # Plot histogram
  plot(h$mids, h_scale, type = "n",
       xlim = range(pd$response$value),
       ylim = c(0, 1.1),
       xlab = v, ylab = "Suitability / Scaled Density",
       main = v, cex.main = 1.2)

  # Available environment as bars
  rect(h$breaks[-length(h$breaks)], 0, h$breaks[-1], h_scale,
       col = adjustcolor("grey80", alpha = 0.5), border = NA)

  # Occurrence histogram overlay
  h_occ <- hist(pd$occurrences, breaks = h$breaks, plot = FALSE)
  h_occ_scale <- h_occ$density / max(h$density) * 0.9
  rect(h_occ$breaks[-length(h_occ$breaks)], 0,
       h_occ$breaks[-1], h_occ_scale,
       col = adjustcolor("#D55E00", alpha = 0.3), border = NA)

  # Response curves
  lines(pd$response$value, pd$response$MaxEnt, col = "#0072B2", lwd = 2)
  lines(pd$response$value, pd$response$BRT, col = "#D55E00", lwd = 2)

  # Mark truncation
  if (tr$truncated_low) {
    abline(v = tr$avail_min, col = "red", lty = 2, lwd = 1.5)
    text(tr$avail_min, 1.05, "TRUNCATED", col = "red", cex = 0.7, pos = 4)
  }
  if (tr$truncated_high) {
    abline(v = tr$avail_max, col = "red", lty = 2, lwd = 1.5)
    text(tr$avail_max, 1.05, "TRUNCATED", col = "red", cex = 0.7, pos = 2)
  }

  legend("topright", legend = c("MaxEnt", "BRT", "Available", "Occurrences"),
         col = c("#0072B2", "#D55E00", "grey60", adjustcolor("#D55E00", 0.5)),
         lty = c(1, 1, NA, NA), lwd = c(2, 2, NA, NA),
         pch = c(NA, NA, 15, 15), pt.cex = 1.5,
         cex = 0.7, bg = "white")
}

mtext(expression(italic("D. variabilis") ~ "— Niche Truncation Assessment"),
      outer = TRUE, cex = 1.2)
```

```{r}
#| label: save-truncation-plot

png(file.path(figures_dir, "08c_niche_truncation.png"),
    width = 14, height = 16, units = "in", res = 300)

par(mfrow = c(nrow_plot, ncol_plot), mar = c(4, 4, 3, 4), oma = c(0, 0, 2, 0))

for (v in selected_vars) {
  pd <- plot_data[[v]]
  tr <- trunc_summary %>% filter(variable == v)

  h <- hist(pd$available, breaks = 50, plot = FALSE)
  h_scale <- h$density / max(h$density) * 0.9

  plot(h$mids, h_scale, type = "n",
       xlim = range(pd$response$value),
       ylim = c(0, 1.1),
       xlab = v, ylab = "Suitability / Scaled Density",
       main = v, cex.main = 1.2)

  rect(h$breaks[-length(h$breaks)], 0, h$breaks[-1], h_scale,
       col = adjustcolor("grey80", alpha = 0.5), border = NA)

  h_occ <- hist(pd$occurrences, breaks = h$breaks, plot = FALSE)
  h_occ_scale <- h_occ$density / max(h$density) * 0.9
  rect(h_occ$breaks[-length(h_occ$breaks)], 0,
       h_occ$breaks[-1], h_occ_scale,
       col = adjustcolor("#D55E00", alpha = 0.3), border = NA)

  lines(pd$response$value, pd$response$MaxEnt, col = "#0072B2", lwd = 2)
  lines(pd$response$value, pd$response$BRT, col = "#D55E00", lwd = 2)

  if (tr$truncated_low) {
    abline(v = tr$avail_min, col = "red", lty = 2, lwd = 1.5)
    text(tr$avail_min, 1.05, "TRUNCATED", col = "red", cex = 0.7, pos = 4)
  }
  if (tr$truncated_high) {
    abline(v = tr$avail_max, col = "red", lty = 2, lwd = 1.5)
    text(tr$avail_max, 1.05, "TRUNCATED", col = "red", cex = 0.7, pos = 2)
  }

  legend("topright", legend = c("MaxEnt", "BRT", "Available", "Occurrences"),
         col = c("#0072B2", "#D55E00", "grey60", adjustcolor("#D55E00", 0.5)),
         lty = c(1, 1, NA, NA), lwd = c(2, 2, NA, NA),
         pch = c(NA, NA, 15, 15), pt.cex = 1.5,
         cex = 0.7, bg = "white")
}

mtext(expression(italic("D. variabilis") ~ "— Niche Truncation Assessment"),
      outer = TRUE, cex = 1.2)
dev.off()
```

## 3. Truncation summary

```{r}
#| label: summary

cat("======================================================\n")
cat("  NICHE TRUNCATION SUMMARY\n")
cat("======================================================\n\n")

cat("Truncation detected at variable range boundaries:\n")
trunc_summary %>%
  dplyr::select(variable, truncated_low, truncated_high,
         occ_at_low_extreme, occ_at_high_extreme,
         low_edge_maxent, high_edge_maxent,
         low_edge_brt, high_edge_brt) %>%
  print(n = Inf)

cat("\n--- Interpretation Guide ---\n")
cat("truncated_low/high: Response curve > 50% of peak at range boundary\n")
cat("occ_at_low/high_extreme: Species occurs within 5% of available range edge\n\n")

# Flag key findings
cat("--- Key Findings ---\n\n")

for (v in selected_vars) {
  tr <- trunc_summary %>% filter(variable == v)
  if (tr$truncated_high & tr$occ_at_high_extreme) {
    cat(sprintf("  %s: UPPER NICHE TRUNCATION — species occurs at warmest/highest\n", v))
    cat("    available conditions. True upper limit cannot be estimated.\n\n")
  }
  if (tr$truncated_low & tr$occ_at_low_extreme) {
    cat(sprintf("  %s: LOWER NICHE TRUNCATION — species occurs at coolest/lowest\n", v))
    cat("    available conditions. True lower limit cannot be estimated.\n\n")
  }
  if (!tr$truncated_low & !tr$truncated_high) {
    cat(sprintf("  %s: No truncation — both niche limits captured within training range.\n\n", v))
  }
}

cat("Variables with upper truncation are problematic for Australian projection:\n")
cat("- If the species tolerates warmer/drier conditions than observed in N. America,\n")
cat("  the model underestimates Australian suitability.\n")
cat("- CLIMEX physiological thresholds (Halliday & Sutherst 1990) can fill this gap:\n")
cat("  optimal 28C, heat stress 35C, lethal 40C.\n")

# Save
write_csv(trunc_summary, file.path(processed_dir, "niche_truncation_summary.csv"))
sprintf("Saved niche truncation assessment")
```
