---
title: "Step 9: Final Visualization and Results"
subtitle: "*Dermacentor variabilis* Species Distribution Model"
author: "Alexander W. Gofton"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme:
      light: flatly
      dark: solar
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    results: hold
    cache.lazy: false
---

## Overview

This script generates publication-quality figures and a comprehensive results
summary for the *D. variabilis* SDM. All figures use a consistent style and
are saved at 300 DPI.

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(patchwork)
library(viridis)
```

## 1. Load all results

```{r}
#| label: load
#| cache: false

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/SDM/Dvariabilis_SDM"
processed_dir <- file.path(base_dir, "processed_data")
figures_dir <- file.path(base_dir, "figures")
outputs_dir <- file.path(base_dir, "outputs")
dir.create(outputs_dir, showWarnings = FALSE)

# Occurrence data
occ <- readRDS(file.path(processed_dir, "occurrences_thinned.rds"))
bg <- readRDS(file.path(processed_dir, "background_coords.rds"))

# Models and metadata
maxent_metadata <- readRDS(file.path(processed_dir, "selected_model_metadata.rds"))
brt_metadata <- readRDS(file.path(processed_dir, "brt_model_metadata.rds"))
thresholds <- readRDS(file.path(processed_dir, "model_thresholds.rds"))

# Predictions
pred_aus_maxent    <- rast(file.path(processed_dir, "pred_aus_maxent.tif"))
pred_aus_brt       <- rast(file.path(processed_dir, "pred_aus_brt.tif"))
pred_aus_climex    <- rast(file.path(processed_dir, "climex_EI_norm_aus.tif"))
pred_aus_ensemble  <- rast(file.path(processed_dir, "pred_aus_ensemble.tif"))
committee_vote_aus <- rast(file.path(processed_dir, "committee_vote_aus.tif"))
pred_M_maxent      <- rast(file.path(processed_dir, "pred_M_maxent.tif"))
mess_aus           <- rast(file.path(processed_dir, "mess_australia.tif"))

# Niche results
niche_results <- readRDS(file.path(processed_dir, "niche_analysis_results.rds"))

# Area summary
area_summary <- read_csv(file.path(processed_dir, "suitable_area_summary.csv"),
                         show_col_types = FALSE)

# Variable importance
maxent_imp <- read_csv(file.path(processed_dir, "variable_importance_final.csv"),
                       show_col_types = FALSE)
brt_imp <- read_csv(file.path(processed_dir, "brt_variable_importance.csv"),
                    show_col_types = FALSE)

# Model comparison
model_comparison <- read_csv(file.path(processed_dir, "model_comparison.csv"),
                             show_col_types = FALSE)

# Spatial data
aus <- ne_countries(country = "Australia", scale = "medium", returnclass = "sf")
states <- ne_states(country = "Australia", returnclass = "sf")
world <- ne_countries(scale = "medium", returnclass = "sf")
M <- readRDS(file.path(processed_dir, "accessible_area_M.rds"))
```

## 2. Figure 1 — Study design overview (occurrences + M + Australia)

```{r}
#| label: fig1
#| fig-width: 14
#| fig-height: 6

occ_sf <- st_as_sf(occ, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)
M_sf <- st_as_sf(M)

# Panel A: Native range with occurrences and M
p_native <- ggplot() +
  geom_sf(data = world, fill = "grey92", colour = "grey70", linewidth = 0.15) +
  geom_sf(data = M_sf, fill = NA, colour = "#0072B2", linewidth = 0.8, linetype = "dashed") +
  geom_sf(data = occ_sf, colour = "#D55E00", size = 0.6, alpha = 0.6) +
  coord_sf(xlim = c(-107, -58), ylim = c(22, 55)) +
  labs(title = "A) Native Range — Occurrences & Accessible Area (M)") +
  theme_minimal(base_size = 10) +
  theme(panel.grid = element_blank(), plot.title = element_text(face = "bold", size = 11))

# Panel B: Australia projection extent
aus_maxent_df <- as.data.frame(pred_aus_ensemble, xy = TRUE) %>%
  filter(!is.na(ensemble_suitability))

p_aus <- ggplot() +
  geom_raster(data = aus_maxent_df, aes(x = x, y = y, fill = ensemble_suitability)) +
  geom_sf(data = states, fill = NA, colour = "grey50", linewidth = 0.3) +
  geom_sf(data = aus, fill = NA, colour = "black", linewidth = 0.5) +
  scale_fill_viridis_c(option = "inferno", limits = c(0, 1), name = "Suitability", trans = "sqrt") +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10)) +
  labs(title = "B) Australia — Ensemble Predicted Suitability") +
  theme_minimal(base_size = 10) +
  theme(panel.grid = element_blank(), plot.title = element_text(face = "bold", size = 11),
        legend.position = "bottom", legend.key.width = unit(2, "cm"))

fig1 <- p_native + p_aus +
  plot_annotation(
    title = expression(italic("Dermacentor variabilis") ~
                       "— Species Distribution Model Study Design"),
    theme = theme(plot.title = element_text(size = 14, face = "bold"))
  )

ggsave(file.path(figures_dir, "fig1_study_design.png"),
       fig1, width = 14, height = 6, dpi = 300)
fig1
```

## 3. Figure 2 — Three-model comparison (MaxEnt, BRT, CLIMEX, ensemble)

```{r}
#| label: fig2
#| fig-width: 14
#| fig-height: 18

maxent_df   <- as.data.frame(pred_aus_maxent,   xy = TRUE) %>% filter(!is.na(maxent_suitability))
brt_df      <- as.data.frame(pred_aus_brt,      xy = TRUE) %>% filter(!is.na(brt_suitability))
climex_df   <- as.data.frame(pred_aus_climex,   xy = TRUE) %>% filter(!is.na(climex_suitability))
ensemble_df <- as.data.frame(pred_aus_ensemble, xy = TRUE) %>% filter(!is.na(ensemble_suitability))
mess_df     <- as.data.frame(mess_aus,          xy = TRUE) %>% filter(!is.na(mess))
vote_df     <- as.data.frame(committee_vote_aus, xy = TRUE) %>%
  filter(!is.na(models_agreeing)) %>%
  mutate(models_agreeing = factor(models_agreeing, levels = 0:3,
                                  labels = c("0 — None", "1 — One model",
                                             "2 — Two models", "3 — All three")))

map_theme <- theme_minimal(base_size = 10) +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 11),
        legend.position = "bottom", legend.key.width = unit(1.8, "cm"))

p_maxent <- ggplot() +
  geom_raster(data = maxent_df, aes(x = x, y = y, fill = maxent_suitability)) +
  geom_sf(data = states, fill = NA, colour = "grey50", linewidth = 0.2) +
  scale_fill_viridis_c(option = "inferno", limits = c(0, 1), name = "Suitability", trans = "sqrt") +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10)) +
  labs(title = "A) MaxEnt (correlative)") + map_theme

p_brt <- ggplot() +
  geom_raster(data = brt_df, aes(x = x, y = y, fill = brt_suitability)) +
  geom_sf(data = states, fill = NA, colour = "grey50", linewidth = 0.2) +
  scale_fill_viridis_c(option = "inferno", limits = c(0, 1), name = "Suitability", trans = "sqrt") +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10)) +
  labs(title = "B) BRT (correlative)") + map_theme

p_climex <- ggplot() +
  geom_raster(data = climex_df, aes(x = x, y = y, fill = climex_suitability)) +
  geom_sf(data = states, fill = NA, colour = "grey50", linewidth = 0.2) +
  scale_fill_viridis_c(option = "inferno", limits = c(0, 1), name = "EI / 100", trans = "sqrt") +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10)) +
  labs(title = "C) CLIMEX EI (process-based, no MESS limitation)") + map_theme

p_ensemble <- ggplot() +
  geom_raster(data = ensemble_df, aes(x = x, y = y, fill = ensemble_suitability)) +
  geom_sf(data = states, fill = NA, colour = "grey50", linewidth = 0.2) +
  scale_fill_viridis_c(option = "inferno", limits = c(0, 1), name = "Suitability", trans = "sqrt") +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10)) +
  labs(title = "D) Ensemble mean (MaxEnt + BRT + CLIMEX)") + map_theme

p_vote <- ggplot() +
  geom_raster(data = vote_df, aes(x = x, y = y, fill = models_agreeing)) +
  geom_sf(data = states, fill = NA, colour = "grey50", linewidth = 0.2) +
  scale_fill_manual(
    values = c("0 — None"       = "grey92",
               "1 — One model"  = "#FEE391",
               "2 — Two models" = "#EC7014",
               "3 — All three"  = "#7F0000"),
    name = "Models agreeing"
  ) +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10)) +
  labs(title = "E) Committee vote — model agreement") + map_theme

p_mess <- ggplot() +
  geom_raster(data = mess_df, aes(x = x, y = y, fill = mess)) +
  geom_sf(data = states, fill = NA, colour = "grey50", linewidth = 0.2) +
  scale_fill_gradient2(low = "#D55E00", mid = "white", high = "#0072B2",
                       midpoint = 0, name = "MESS") +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10)) +
  labs(title = "F) MESS — Environmental Similarity",
       subtitle = "Red = novel climate | Blue = analogous") + map_theme

fig2 <- (p_maxent + p_brt) / (p_climex + p_ensemble) / (p_vote + p_mess) +
  plot_annotation(
    title = expression(italic("Dermacentor variabilis") ~
                       "— Three-Model Projected Climatic Suitability in Australia"),
    theme = theme(plot.title = element_text(size = 14, face = "bold"))
  )

ggsave(file.path(figures_dir, "fig2_australia_suitability.png"),
       fig2, width = 14, height = 18, dpi = 300)
fig2
```

## 4. Figure 3 — Three-model consensus binary map

This map combines information from all three models into a consensus classification,
distinguishing between regions predicted suitable by: all three models (highest
confidence), correlative models only (within analogue climate), CLIMEX only (in
novel climate — where correlative models cannot reliably predict), and no models.

```{r}
#| label: fig3
#| fig-width: 10
#| fig-height: 8

mess_mask <- mess_aus >= 0  # 1 = analogous climate, 0 = novel
thresholds <- readRDS(file.path(processed_dir, "model_thresholds.rds"))

# Reconstruct binary votes (already saved but recompute for clarity)
maxent_vote_09 <- ifel(pred_aus_maxent >= thresholds$maxent$maxss, 1, 0)
brt_vote_09    <- ifel(pred_aus_brt    >= thresholds$brt$maxss,    1, 0)
climex_vote_09 <- ifel(pred_aus_climex >= 0.10,                    1, 0)
vote_sum_09    <- maxent_vote_09 + brt_vote_09 + climex_vote_09

# Build a categorical map with 6 meaningful classes
# Priority: all-three > two-with-analogous > CLIMEX-in-novel > one-correlative > unsuitable
consensus <- vote_sum_09  # temp; overwrite with classification
consensus[] <- 0  # default: unsuitable

# Class 1: All three agree suitable (high confidence)
consensus[vote_sum_09 == 3] <- 5

# Class 2: ≥2 models agree, analogous climate (good confidence)
consensus[vote_sum_09 == 2 & mess_mask == 1] <- 4

# Class 3: ≥2 models agree, novel climate (moderate — includes CLIMEX support)
consensus[vote_sum_09 == 2 & mess_mask == 0] <- 3

# Class 4: CLIMEX only, novel climate (CLIMEX-only extrapolation zone)
consensus[climex_vote_09 == 1 & (maxent_vote_09 + brt_vote_09) == 0 & mess_mask == 0] <- 2

# Class 5: Correlative models only (1 model, analogous climate)
consensus[(maxent_vote_09 == 1 | brt_vote_09 == 1) & climex_vote_09 == 0 &
            vote_sum_09 == 1 & mess_mask == 1] <- 1

names(consensus) <- "lyr.1"
cat_df <- as.data.frame(consensus, xy = TRUE) %>%
  filter(!is.na(lyr.1)) %>%
  mutate(category = factor(lyr.1,
    levels = 0:5,
    labels = c(
      "0 — Not suitable (any model)",
      "1 — Correlative only, analogous climate",
      "2 — CLIMEX only, novel climate",
      "3 — ≥2 models agree, novel climate",
      "4 — ≥2 models agree, analogous climate",
      "5 — All 3 models agree"
    )
  ))

fig3 <- ggplot() +
  geom_raster(data = cat_df, aes(x = x, y = y, fill = category)) +
  geom_sf(data = states, fill = NA, colour = "grey40", linewidth = 0.3) +
  geom_sf(data = aus, fill = NA, colour = "black", linewidth = 0.5) +
  scale_fill_manual(
    values = c(
      "0 — Not suitable (any model)"          = "grey90",
      "1 — Correlative only, analogous climate" = "#FEE391",
      "2 — CLIMEX only, novel climate"         = "#74C476",
      "3 — ≥2 models agree, novel climate"     = "#E69F00",
      "4 — ≥2 models agree, analogous climate" = "#D55E00",
      "5 — All 3 models agree"                 = "#7F0000"
    ),
    name = NULL,
    drop = FALSE
  ) +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10)) +
  labs(
    title = expression(italic("D. variabilis") ~
                       "— Three-Model Consensus Suitability"),
    subtitle = paste(
      "MaxEnt + BRT + CLIMEX. Green = CLIMEX-only in novel climate (e.g. north QLD).",
      "Correlative models cannot predict here due to climate novelty."
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", size = 13),
    legend.position = "bottom",
    legend.text = element_text(size = 9)
  ) +
  guides(fill = guide_legend(nrow = 3))

ggsave(file.path(figures_dir, "fig3_binary_suitability_mess.png"),
       fig3, width = 10, height = 8, dpi = 300)
fig3
```

## 5. Results summary tables

```{r}
#| label: results-tables

# Table 1: Model comparison
cat("=== TABLE 1: Model Performance Comparison ===\n")
print(model_comparison, n = Inf)

# Table 2: Variable importance
cat("\n=== TABLE 2: Variable Importance ===\n")
imp_combined <- maxent_imp %>%
  dplyr::select(variable, percent.contribution, permutation.importance) %>%
  rename(maxent_contribution = percent.contribution,
         maxent_permutation = permutation.importance) %>%
  left_join(
    brt_imp %>% rename(brt_relative_influence = relative_influence),
    by = "variable"
  ) %>%
  arrange(desc(maxent_permutation))
print(imp_combined, n = Inf)

# Table 3: Area of suitable habitat
cat("\n=== TABLE 3: Suitable Habitat Area in Australia ===\n")
print(area_summary, n = Inf)

# Table 4: Niche analysis results
cat("\n=== TABLE 4: Niche Analysis ===\n")
niche_summary <- read_csv(file.path(processed_dir, "niche_overlap_summary.csv"),
                          show_col_types = FALSE)
print(niche_summary, n = Inf)

# Save all tables
write_csv(model_comparison, file.path(outputs_dir, "table1_model_comparison.csv"))
write_csv(imp_combined, file.path(outputs_dir, "table2_variable_importance.csv"))
write_csv(area_summary, file.path(outputs_dir, "table3_suitable_area.csv"))
write_csv(niche_summary, file.path(outputs_dir, "table4_niche_analysis.csv"))
```

## 6. Comprehensive results summary

```{r}
#| label: final-summary

cat("================================================================\n")
cat("  SPECIES DISTRIBUTION MODEL RESULTS SUMMARY\n")
cat("  Dermacentor variabilis — Climatic Suitability for Australia\n")
cat("================================================================\n\n")

cat("RESEARCH QUESTION:\n")
cat("If D. variabilis were introduced to Australia, does Australia have\n")
cat("the climatic conditions required for the species to survive?\n\n")

cat("DATA:\n")
cat(sprintf("  Occurrence records: %d (GBIF, thinned to 10 km, eastern N. America)\n", nrow(occ)))
cat(sprintf("  Background points: %d (Ixodidae target-group from GBIF)\n", nrow(bg)))
cat(sprintf("  Environmental variables: %d selected from %d candidates\n",
            length(maxent_metadata$variables),
            length(readLines(file.path(processed_dir, "candidate_variables.txt")))))
cat("  Variables:", paste(maxent_metadata$variables, collapse = ", "), "\n\n")

cat("MODEL PERFORMANCE:\n")
cat(sprintf("  MaxEnt: fc=%s, rm=%s, AUC_train=%.3f, AUC_val=%.3f\n",
            maxent_metadata$feature_class, maxent_metadata$reg_multiplier,
            maxent_metadata$auc_train, maxent_metadata$auc_val))
cat(sprintf("  BRT: %d trees, tc=%d, CV_cor=%.3f\n",
            brt_metadata$n_trees, brt_metadata$tree_complexity,
            brt_metadata$cv_correlation))

cat("\nPROJECTION TO AUSTRALIA (Three-model ensemble):\n")
print(area_summary, n = Inf)
cat("\nNote: Ensemble uses MaxEnt + BRT (correlative) + CLIMEX (process-based).\n")
cat("CLIMEX covers novel-climate zones (e.g. north QLD) missed by correlative models.\n")

mess_vals <- values(mess_aus, na.rm = TRUE)
pct_novel <- round(100 * sum(mess_vals < 0) / length(mess_vals), 1)
cat(sprintf("\n  MESS: %.1f%% of Australia has novel climate (MESS < 0)\n", pct_novel))

cat("\nNICHE ANALYSIS:\n")
cat(sprintf("  Schoener's D: %.3f\n", niche_results$overlap$D))
cat(sprintf("  Warren's I: %.3f\n", niche_results$overlap$I))
cat(sprintf("  Equivalency test p-value: %.4f\n", niche_results$equivalency$p_D))
cat(sprintf("  Similarity test p-value: %.4f\n", niche_results$similarity_native_to_aus$p_D))

cat("\n================================================================\n")
cat("  END OF RESULTS SUMMARY\n")
cat("================================================================\n")
```

## 7. Save final results object

```{r}
#| label: save-final

# Comprehensive results object
final_results <- list(
  species = "Dermacentor variabilis",
  n_occurrences = nrow(occ),
  n_background = nrow(bg),
  selected_variables = maxent_metadata$variables,
  maxent = maxent_metadata,
  brt = brt_metadata,
  thresholds = thresholds,
  area_summary = area_summary,
  niche_analysis = niche_results,
  mess_pct_novel = pct_novel
)

saveRDS(final_results, file.path(outputs_dir, "final_results.rds"))

sprintf("All results and figures saved to outputs/ and figures/")
```
