---
title: "Step 10: Climate Analogue Mapping"
subtitle: "*Dermacentor variabilis* Species Distribution Model"
author: "Alexander W. Gofton"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme:
      light: flatly
      dark: solar
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    results: hold
    cache.lazy: false
---

## Overview

This script provides a **model-free** assessment of climatic similarity between
the *D. variabilis* native range and Australia. Instead of fitting a correlative
model, we directly compute the **Mahalanobis distance** from each Australian cell
to the climatic centroid of the native range, using only ecologically appropriate
variables (moisture and warm-season temperature — Set C).

This approach:

- Avoids extrapolation artefacts inherent in correlative SDMs
- Is not sensitive to model algorithm choice (MaxEnt vs BRT)
- Provides an independent validation for SDM predictions
- Can be directly compared with CLIMEX Ecoclimatic Index

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
```

## 1. Load data

```{r}
#| label: load

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/SDM/Dvariabilis_SDM"
processed_dir <- file.path(base_dir, "processed_data")
figures_dir <- file.path(base_dir, "figures")

# Full environmental stacks
env_stack_aus <- rast(file.path(processed_dir, "env_stack_aus.tif"))

# Occurrence data (contains all 21 variables)
occ_env <- readRDS(file.path(processed_dir, "occurrences_with_env.rds"))

# Moisture-focused variables (Set C)
set_C_file <- file.path(processed_dir, "selected_variables_setC.txt")
if (file.exists(set_C_file)) {
  set_C <- readLines(set_C_file)
} else {
  set_C <- c("bio5", "bio10", "bio12", "bio15",
             "aridity_idx", "PET_seasonality", "vpd_annual")
  set_C <- intersect(set_C, names(env_stack_aus))
}

aus <- ne_countries(country = "Australia", scale = "medium", returnclass = "sf")

cat("Climate analogue variables:", paste(set_C, collapse = ", "), "\n")
```

## 2. Characterise native range climate

```{r}
#| label: native-climate

# Extract environmental values at occurrence points
occ_climate <- occ_env %>%
  dplyr::select(all_of(set_C)) %>%
  na.omit()

# Core climate envelope (10th-90th percentile)
core_envelope <- tibble()
for (v in set_C) {
  core_envelope <- bind_rows(core_envelope, tibble(
    variable = v,
    q10 = quantile(occ_climate[[v]], 0.10),
    q50 = quantile(occ_climate[[v]], 0.50),
    q90 = quantile(occ_climate[[v]], 0.90),
    mean = mean(occ_climate[[v]]),
    sd = sd(occ_climate[[v]])
  ))
}

cat("Native range core climate envelope (10th-90th percentile):\n")
print(core_envelope, n = Inf)

# Scale statistics derived from occurrence data
# Used to normalise both occurrence data and Australian cells to unit variance
# before computing Mahalanobis distance.
# Rationale: variables in this set span very different scales (e.g. bio12 mean
# diurnal range SD ≈ 1.3°C vs PET_seasonality SD ≈ 479). Without scaling, low-SD
# variables dominate the Mahalanobis metric regardless of ecological relevance.
# Scaling to unit variance (using occurrence SD) is equivalent to using the
# correlation matrix, giving equal weight per standardised unit of deviation.
occ_scale_center <- colMeans(occ_climate)
occ_scale_sd     <- apply(occ_climate, 2, sd)

cat("\nNative range centroid (unscaled):\n")
print(round(occ_scale_center, 2))
cat("\nPer-variable SD (used for scaling):\n")
print(round(occ_scale_sd, 2))

# Scale occurrence data to zero mean, unit variance
occ_scaled <- scale(occ_climate, center = occ_scale_center, scale = occ_scale_sd)

# Covariance matrix and centroid on SCALED data
cov_matrix <- cov(occ_scaled)
centroid    <- colMeans(occ_scaled)
```

## 3. Compute Mahalanobis distance across Australia

```{r}
#| label: mahalanobis
#| cache: true

# Subset Australian stack to Set C variables
env_aus_C <- env_stack_aus[[set_C]]

# Extract all Australian cell values
aus_vals <- as.data.frame(env_aus_C, xy = TRUE, na.rm = TRUE)
aus_env <- aus_vals %>% dplyr::select(all_of(set_C))

# Scale Australian cells using the SAME center and SD as the occurrence data
# This ensures distances are expressed in units of the native range SD
aus_scaled <- scale(aus_env, center = occ_scale_center, scale = occ_scale_sd)

# Compute Mahalanobis distance from each scaled cell to the scaled centroid
# Both aus_scaled and centroid are now on the same standardised scale
mah_dist <- mahalanobis(aus_scaled, center = centroid, cov = cov_matrix)

# Add to data frame
aus_vals$mahal_dist <- mah_dist

# Convert to raster
mah_rast <- rast(aus_vals %>% dplyr::select(x, y, mahal_dist),
                 type = "xyz", crs = "EPSG:4326")

# Chi-squared threshold for outlier detection (df = number of variables)
chi_sq_95 <- qchisq(0.95, df = length(set_C))
chi_sq_99 <- qchisq(0.99, df = length(set_C))

# Climate analogue classification
aus_vals$analogue_class <- case_when(
  aus_vals$mahal_dist <= chi_sq_95 ~ "High similarity (p > 0.05)",
  aus_vals$mahal_dist <= chi_sq_99 ~ "Moderate similarity (0.01 < p < 0.05)",
  TRUE ~ "Low similarity (p < 0.01)"
)

cat("Mahalanobis distance summary:\n")
sprintf("  Range: %.1f to %.1f", min(mah_dist), max(mah_dist))
sprintf("  Chi-sq 95th percentile (df=%d): %.1f", length(set_C), chi_sq_95)
sprintf("  Chi-sq 99th percentile (df=%d): %.1f", length(set_C), chi_sq_99)

n_total <- nrow(aus_vals)
sprintf("  High similarity (Mah <= %.1f): %.1f%%",
        chi_sq_95, 100 * sum(mah_dist <= chi_sq_95) / n_total)
sprintf("  Moderate similarity: %.1f%%",
        100 * sum(mah_dist > chi_sq_95 & mah_dist <= chi_sq_99) / n_total)
sprintf("  Low similarity (Mah > %.1f): %.1f%%",
        chi_sq_99, 100 * sum(mah_dist > chi_sq_99) / n_total)
```

## 4. Climate analogue maps

```{r}
#| label: maps
#| fig-width: 14
#| fig-height: 10

# Continuous Mahalanobis distance map
p1 <- ggplot() +
  geom_raster(data = aus_vals, aes(x = x, y = y, fill = mahal_dist)) +
  geom_sf(data = aus, fill = NA, colour = "black", linewidth = 0.3) +
  scale_fill_viridis_c(
    name = "Mahalanobis\nDistance",
    option = "magma",
    direction = -1,
    trans = "sqrt",
    limits = c(0, quantile(mah_dist, 0.99))
  ) +
  labs(
    title = expression(italic("D. variabilis") ~ "— Climate Analogue: Mahalanobis Distance"),
    subtitle = "Lower values = more similar to native range climate (Set C variables)",
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal(base_size = 11) +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10))

# Categorical analogue map
aus_vals$analogue_class <- factor(
  aus_vals$analogue_class,
  levels = c("High similarity (p > 0.05)",
             "Moderate similarity (0.01 < p < 0.05)",
             "Low similarity (p < 0.01)")
)

p2 <- ggplot() +
  geom_raster(data = aus_vals, aes(x = x, y = y, fill = analogue_class)) +
  geom_sf(data = aus, fill = NA, colour = "black", linewidth = 0.3) +
  scale_fill_manual(
    values = c("High similarity (p > 0.05)" = "#D55E00",
               "Moderate similarity (0.01 < p < 0.05)" = "#E69F00",
               "Low similarity (p < 0.01)" = "#56B4E9"),
    name = "Climate\nSimilarity"
  ) +
  labs(
    title = expression(italic("D. variabilis") ~ "— Climate Analogue Classification"),
    subtitle = sprintf("Based on Mahalanobis distance (chi-sq df=%d)", length(set_C)),
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal(base_size = 11) +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10))

library(patchwork)
p_combined <- p1 + p2 + plot_layout(ncol = 2)
print(p_combined)

ggsave(file.path(figures_dir, "10_climate_analogue.png"),
       p_combined, width = 14, height = 6, dpi = 300)
```

## 5. CLIMEX reference location comparison

```{r}
#| label: reference-locations

# Key locations from Halliday & Sutherst (1990)
ref_locs <- tibble(
  city = c("Innisfail", "Lismore", "Sydney", "Melbourne", "Perth",
           "Darwin", "Alice Springs", "Hobart", "Brisbane", "Cairns"),
  lon = c(146.03, 153.28, 151.21, 144.96, 115.86,
          130.84, 133.88, 147.33, 153.03, 145.77),
  lat = c(-17.53, -28.81, -33.87, -37.81, -31.95,
          -12.46, -23.70, -42.88, -27.47, -16.92),
  climex_EI = c(63, 43, NA, NA, NA, NA, NA, NA, NA, NA),
  climex_note = c("Highest EI", "~ Little Rock AR (EI=44)",
                  "", "", "Dry", "Tropical", "Arid",
                  "Cool temperate", "", "Tropical")
)

# Extract Mahalanobis distance at reference locations
ref_pts <- vect(ref_locs, geom = c("lon", "lat"), crs = "EPSG:4326")
ref_locs$mahal_dist <- extract(mah_rast, ref_pts)[, 2]
ref_locs$p_value <- 1 - pchisq(ref_locs$mahal_dist, df = length(set_C))

cat("CLIMEX reference locations — Mahalanobis distance:\n")
ref_locs %>%
  mutate(mahal_dist = round(mahal_dist, 1),
         p_value = round(p_value, 4)) %>%
  print(n = Inf)

cat("\nInterpretation:\n")
cat("Locations with p > 0.05 have climate within the 95% confidence envelope\n")
cat("of the native range. Compare with CLIMEX EI: higher EI = more suitable.\n")
```

## 6. Calculate analogue area

```{r}
#| label: area

# Convert to raster for area calculation
analogue_binary <- mah_rast <= chi_sq_95
cell_areas <- cellSize(analogue_binary, unit = "km")

analogue_area <- sum(values(cell_areas * analogue_binary), na.rm = TRUE)
total_area <- sum(values(cell_areas * !is.na(analogue_binary)), na.rm = TRUE)

sprintf("Climate analogue area (Mahalanobis, 95%% CI): %.0f km2 (%.1f%% of Australia)",
        analogue_area, 100 * analogue_area / total_area)
```

## 7. Summary

```{r}
#| label: summary

cat("======================================================\n")
cat("  CLIMATE ANALOGUE ANALYSIS SUMMARY\n")
cat("======================================================\n\n")

cat("Method: Mahalanobis distance from Australian cells to native-range centroid\n")
cat(sprintf("Variables (Set C): %s\n", paste(set_C, collapse = ", ")))
cat(sprintf("Number of occurrence points: %d\n\n", nrow(occ_climate)))

cat("Results:\n")
sprintf("  High similarity area: %.0f km2 (%.1f%%)",
        analogue_area, 100 * analogue_area / total_area)
sprintf("  This represents the area where Australian climate falls within")
sprintf("  the 95%% confidence envelope of the native range climate.\n")

cat("\nKey advantage over correlative SDMs:\n")
cat("  - No extrapolation artefacts\n")
cat("  - Not sensitive to model algorithm choice\n")
cat("  - Uses ecologically appropriate variables only\n")
cat("  - Accounts for covariance structure among variables\n")

# Save outputs
writeRaster(mah_rast, file.path(processed_dir, "mahalanobis_distance_aus.tif"),
            overwrite = TRUE)
write_csv(ref_locs, file.path(processed_dir, "climate_analogue_reference_locations.csv"))

sprintf("Saved climate analogue outputs")
```
