---
title: "Step 10b: Limiting Factor Analysis"
subtitle: "*Dermacentor variabilis* Species Distribution Model"
author: "Alexander W. Gofton"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme:
      light: flatly
      dark: solar
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    results: hold
    cache.lazy: false
---

## Overview

For each cell in Australia, this script identifies the **most limiting environmental
variable** — the variable whose value departs most from the species' optimal range.
This provides mechanistic insight consistent with the CLIMEX approach.

The 1990 Halliday & Sutherst CLIMEX study found that **low summer humidity** is the
primary factor preventing *D. variabilis* establishment in inland/western Australia.
If aridity_idx or PET_seasonality emerges as the dominant limiting factor in those
regions, this independently validates the CLIMEX finding within our correlative
SDM framework.

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
```

## 1. Load data

```{r}
#| label: load

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/SDM/Dvariabilis_SDM"
processed_dir <- file.path(base_dir, "processed_data")
figures_dir <- file.path(base_dir, "figures")

# Use Set C variables (moisture-focused) for ecologically meaningful assessment
set_C_file <- file.path(processed_dir, "selected_variables_setC.txt")
if (file.exists(set_C_file)) {
  analysis_vars <- readLines(set_C_file)
} else {
  analysis_vars <- c("bio5", "bio10", "bio12", "bio15",
                     "aridity_idx", "PET_seasonality", "vpd_annual")
}

# Also run with original Set A for comparison
set_A <- readLines(file.path(processed_dir, "selected_variables.txt"))

# Environmental stack
env_stack_aus <- rast(file.path(processed_dir, "env_stack_aus.tif"))

# Occurrence data
occ_env <- readRDS(file.path(processed_dir, "occurrences_with_env.rds"))

aus <- ne_countries(country = "Australia", scale = "medium", returnclass = "sf")

cat("Set A variables:", paste(set_A, collapse = ", "), "\n")
cat("Set C variables:", paste(analysis_vars, collapse = ", "), "\n")
```

## 2. Define optimal ranges from occurrence data

```{r}
#| label: optimal-ranges

compute_optimal <- function(occ_data, vars) {
  optimal <- tibble()
  for (v in vars) {
    vals <- occ_data[[v]]
    optimal <- bind_rows(optimal, tibble(
      variable = v,
      q10 = quantile(vals, 0.10, na.rm = TRUE),
      q50 = quantile(vals, 0.50, na.rm = TRUE),
      q90 = quantile(vals, 0.90, na.rm = TRUE),
      range = as.numeric(quantile(vals, 0.90, na.rm = TRUE) - quantile(vals, 0.10, na.rm = TRUE))
    ))
  }
  optimal
}

optimal_A <- compute_optimal(occ_env, set_A)
optimal_C <- compute_optimal(occ_env, analysis_vars)

cat("Optimal ranges (10th-90th percentile of occurrences) — Set C:\n")
print(optimal_C, n = Inf)
```

## 3. Compute limiting factor maps

For each cell, compute the **standardised departure** from the optimal range for
each variable. The variable with the greatest departure is the most limiting factor.

Departure is standardised by the species' tolerance breadth (range of 10th-90th
percentile) so that variables measured in different units are comparable.

```{r}
#| label: limiting-factors
#| cache: true

compute_limiting <- function(env_stack, vars, optimal, set_name) {
  cat(sprintf("Computing limiting factors for %s...\n", set_name))

  # Compute standardised departure for each variable
  departure_layers <- list()
  for (v in vars) {
    opt <- optimal %>% filter(variable == v)
    layer <- env_stack[[v]]

    # Departure: 0 if within optimal range, positive if outside
    # Standardised by tolerance breadth
    dep <- app(layer, function(x) {
      vapply(x, function(xi) {
        if (is.na(xi)) return(NA_real_)
        if (xi < opt$q10) return((opt$q10 - xi) / opt$range)
        if (xi > opt$q90) return((xi - opt$q90) / opt$range)
        return(0)
      }, numeric(1))
    })
    departure_layers[[v]] <- dep
  }

  dep_stack <- rast(departure_layers)
  names(dep_stack) <- vars

  # Most limiting = variable with maximum departure
  max_dep <- max(dep_stack)
  limiting_var <- which.max(dep_stack)

  # Total departure (sum of all variable departures)
  total_dep <- sum(dep_stack)

  # Count of variables outside optimal range
  n_outside <- sum(dep_stack > 0)

  list(
    departure_stack = dep_stack,
    max_departure = max_dep,
    limiting_var = limiting_var,
    total_departure = total_dep,
    n_outside = n_outside
  )
}

lim_A <- compute_limiting(env_stack_aus, set_A, optimal_A, "Set A")
lim_C <- compute_limiting(env_stack_aus, analysis_vars, optimal_C, "Set C")
```

## 4. Limiting factor maps

```{r}
#| label: maps
#| fig-width: 14
#| fig-height: 12

# Set C limiting factor map
lim_df_C <- as.data.frame(lim_C$limiting_var, xy = TRUE, na.rm = TRUE)
names(lim_df_C)[3] <- "lim_idx"
lim_df_C$variable <- analysis_vars[lim_df_C$lim_idx]
lim_df_C$variable <- factor(lim_df_C$variable, levels = analysis_vars)

dep_df_C <- as.data.frame(lim_C$max_departure, xy = TRUE, na.rm = TRUE)
names(dep_df_C)[3] <- "max_dep"

# Limiting factor dominance counts
lim_counts_C <- lim_df_C %>%
  count(variable) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  arrange(desc(pct))

cat("Set C — Most limiting variable dominance:\n")
print(as_tibble(lim_counts_C), n = Inf)

# Map 1: Most limiting variable
p1 <- ggplot() +
  geom_raster(data = lim_df_C, aes(x = x, y = y, fill = variable)) +
  geom_sf(data = aus, fill = NA, colour = "black", linewidth = 0.3) +
  scale_fill_brewer(palette = "Set2", name = "Most Limiting\nVariable") +
  labs(
    title = expression(italic("D. variabilis") ~ "— Most Limiting Factor (Set C)"),
    subtitle = "Variable with greatest standardised departure from optimal range",
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal(base_size = 11) +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10))

# Map 2: Maximum departure magnitude
p2 <- ggplot() +
  geom_raster(data = dep_df_C, aes(x = x, y = y, fill = max_dep)) +
  geom_sf(data = aus, fill = NA, colour = "black", linewidth = 0.3) +
  scale_fill_viridis_c(
    name = "Max Departure\n(standardised)",
    option = "inferno",
    trans = "sqrt",
    limits = c(0, quantile(dep_df_C$max_dep, 0.99, na.rm = TRUE))
  ) +
  labs(
    title = expression(italic("D. variabilis") ~ "— Maximum Environmental Departure"),
    subtitle = "0 = within optimal range | Higher = further from optimal",
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal(base_size = 11) +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10))

library(patchwork)
p_combined <- p1 / p2
print(p_combined)

ggsave(file.path(figures_dir, "10b_limiting_factors.png"),
       p_combined, width = 10, height = 12, dpi = 300)
```

## 5. Set A comparison

```{r}
#| label: setA-comparison

lim_df_A <- as.data.frame(lim_A$limiting_var, xy = TRUE, na.rm = TRUE)
names(lim_df_A)[3] <- "lim_idx"
lim_df_A$variable <- set_A[lim_df_A$lim_idx]

lim_counts_A <- lim_df_A %>%
  count(variable) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  arrange(desc(pct))

cat("\nSet A — Most limiting variable dominance:\n")
print(as_tibble(lim_counts_A), n = Inf)

cat("\nComparison:\n")
cat("If Bio1 or Bio7 dominates Set A but moisture variables dominate Set C,\n")
cat("this confirms that cold-tolerance variables artificially constrain the\n")
cat("SDM projections in regions where they are ecologically irrelevant.\n")
```

## 6. Cells within optimal range for all variables

```{r}
#| label: fully-suitable

# Cells where ALL variables are within optimal range (departure = 0 for all)
fully_suitable_C <- lim_C$n_outside == 0
cell_areas <- cellSize(fully_suitable_C, unit = "km")
fully_suitable_area <- sum(values(cell_areas * fully_suitable_C), na.rm = TRUE)
total_area <- sum(values(cell_areas * !is.na(fully_suitable_C)), na.rm = TRUE)

sprintf("Cells within optimal range for ALL Set C variables: %.0f km2 (%.1f%%)",
        fully_suitable_area, 100 * fully_suitable_area / total_area)

# Cells within optimal range for most variables (n_outside <= 1)
mostly_suitable <- lim_C$n_outside <= 1
mostly_suitable_area <- sum(values(cell_areas * mostly_suitable), na.rm = TRUE)
sprintf("Cells with at most 1 variable outside optimal: %.0f km2 (%.1f%%)",
        mostly_suitable_area, 100 * mostly_suitable_area / total_area)
```

## 7. Summary

```{r}
#| label: summary

cat("======================================================\n")
cat("  LIMITING FACTOR ANALYSIS SUMMARY\n")
cat("======================================================\n\n")

cat("Method: Standardised departure from species' optimal range (10-90th pctl)\n")
cat(sprintf("Variables: %s\n\n", paste(analysis_vars, collapse = ", ")))

cat("Most limiting variables (Set C, % of Australia):\n")
print(as_tibble(lim_counts_C), n = Inf)

moisture_vars <- c("aridity_idx", "PET_seasonality", "vpd_annual", "bio12", "bio15")
moisture_pct <- lim_counts_C %>%
  filter(variable %in% moisture_vars) %>%
  summarise(total_pct = sum(pct)) %>%
  pull(total_pct)

temp_vars <- c("bio5", "bio10")
temp_pct <- lim_counts_C %>%
  filter(variable %in% temp_vars) %>%
  summarise(total_pct = sum(pct)) %>%
  pull(total_pct)

cat(sprintf("\nMoisture-related variables dominate: %.1f%% of cells\n", moisture_pct))
cat(sprintf("Temperature-related variables dominate: %.1f%% of cells\n", temp_pct))

if (moisture_pct > temp_pct) {
  cat("\nFINDING: Moisture is the PRIMARY limiting factor across Australia.\n")
  cat("This is CONSISTENT with Halliday & Sutherst (1990) CLIMEX findings.\n")
}

# Save
write_csv(lim_counts_C, file.path(processed_dir, "limiting_factor_counts.csv"))
writeRaster(lim_C$limiting_var,
            file.path(processed_dir, "limiting_factor_map_setC.tif"),
            overwrite = TRUE)

sprintf("Saved limiting factor analysis outputs")
```
