---
title: "Step 10c: CLIMEX-Informed Constraint Mapping"
subtitle: "*Dermacentor variabilis* Species Distribution Model"
author: "Alexander W. Gofton"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme:
      light: flatly
      dark: solar
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    results: hold
    cache.lazy: false
---

## Overview

This script uses physiological thresholds from the 1990 Halliday & Sutherst CLIMEX
study to create simple binary constraint maps for *D. variabilis* in Australia.
Unlike correlative SDMs, this approach applies **process-based physiological limits**
derived from the species' known biology.

CLIMEX parameters from Halliday & Sutherst (1990):

- Lower developmental temperature threshold: **4C**
- Optimal temperature: **28C**
- Heat stress onset: **35C**
- Upper lethal temperature: **40C**
- Primary limiting factor in Australia: **low summer humidity**

This provides an independent baseline against which MaxEnt and BRT predictions
can be evaluated.

**NOTE — Variable naming:** The WorldClim bioclimatic variables in this project's
`env_stack_aus.tif` are named by lexicographic sort order of files (bio_1, bio_10, bio_11,
..., bio_2...) rather than numerical order, causing multi-digit variables to be relabelled.
The relevant temperature variables in the stack are:

- `bio15` = WorldClim bio_5 (Max Temperature of Warmest Month, °C)
- `bio2`  = WorldClim bio_10 (Mean Temperature of Warmest Quarter, °C)
- `bio16` = WorldClim bio_6 (Min Temperature of Coldest Month, °C)

Temperature thresholds in this script use `bio15` and `bio2` (not `bio5` and `bio10`,
which are precipitation variables in this mislabelled stack).

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
```

## 1. Load environmental data

We need the **full** 21-variable environmental stack for Australia (not just selected
variables) because CLIMEX thresholds apply to raw bioclim variables.

```{r}
#| label: load

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/SDM/Dvariabilis_SDM"
processed_dir <- file.path(base_dir, "processed_data")
figures_dir <- file.path(base_dir, "figures")

# Full environmental stack (all candidate variables)
env_aus_full <- rast(file.path(processed_dir, "env_stack_aus.tif"))
cat("Available variables:", paste(names(env_aus_full), collapse = ", "), "\n")

# Selected-variable predictions for comparison
pred_maxent <- rast(file.path(processed_dir, "pred_aus_maxent.tif"))
pred_brt <- rast(file.path(processed_dir, "pred_aus_brt.tif"))
thresholds <- readRDS(file.path(processed_dir, "model_thresholds.rds"))

# Australia boundary
aus <- ne_countries(country = "Australia", scale = "medium", returnclass = "sf")

# Check which bioclim variables are available
bioclim_vars <- names(env_aus_full)[grepl("^bio", names(env_aus_full))]
cat("Bioclim variables available:", paste(bioclim_vars, collapse = ", "), "\n")
```

## 2. Apply CLIMEX temperature constraints

```{r}
#| label: temp-constraints

# NOTE: Due to lexicographic file sort in 03_environmental_data.qmd, the correct
# temperature variables in this stack are:
#   bio15 = WorldClim bio_5 (Max Temperature of Warmest Month, °C)
#   bio2  = WorldClim bio_10 (Mean Temperature of Warmest Quarter, °C)
# Do NOT use bio5 or bio10 — those are precipitation variables in this stack.

# Constraint 1: Exclude where max warm-month temp > 40C (upper lethal)
if ("bio15" %in% names(env_aus_full)) {
  heat_lethal <- env_aus_full[["bio15"]] <= 40
  heat_stress <- env_aus_full[["bio15"]] <= 35
  cat("Heat constraints applied using bio15 (= WorldClim bio_5: max temp warmest month, °C)\n")
} else {
  stop("bio15 (max temp warmest month) not available in environmental stack")
}

# Constraint 2: Exclude where mean warmest quarter temp < 4C (below dev threshold)
# This is very unlikely in Australia but included for completeness
if ("bio2" %in% names(env_aus_full)) {
  cold_dev <- env_aus_full[["bio2"]] >= 4
  cat("Cold development constraint applied using bio2 (= WorldClim bio_10: mean temp warmest quarter, °C)\n")
} else {
  stop("bio2 (mean temp warmest quarter) not available in environmental stack")
}

# Constraint 3: Warm-season activity window — at least moderate warmth
# Optimal temp is 28C, so require mean warmest quarter >= 15C for meaningful activity
warm_activity <- env_aus_full[["bio2"]] >= 15

# Summary
cat("\nTemperature constraint summary:\n")
vals_heat <- values(heat_lethal, na.rm = TRUE)
vals_stress <- values(heat_stress, na.rm = TRUE)
vals_cold <- values(cold_dev, na.rm = TRUE)
vals_warm <- values(warm_activity, na.rm = TRUE)

sprintf("  Heat lethal (bio15/WorldClim-bio5 <= 40C): %.1f%% of cells pass",
        100 * sum(vals_heat) / length(vals_heat))
sprintf("  Heat stress (bio15/WorldClim-bio5 <= 35C): %.1f%% of cells pass",
        100 * sum(vals_stress) / length(vals_stress))
sprintf("  Cold development (bio2/WorldClim-bio10 >= 4C): %.1f%% of cells pass",
        100 * sum(vals_cold) / length(vals_cold))
sprintf("  Warm activity (bio2/WorldClim-bio10 >= 15C): %.1f%% of cells pass",
        100 * sum(vals_warm) / length(vals_warm))
```

## 3. Apply CLIMEX moisture constraints

The 1990 CLIMEX study identified **low summer humidity** as the primary factor
preventing *D. variabilis* establishment in inland/western Australia.

```{r}
#| label: moisture-constraints

# Use aridity index as primary moisture constraint
# Higher aridity_idx = more arid = less suitable
if ("aridity_idx" %in% names(env_aus_full)) {
  aridity <- env_aus_full[["aridity_idx"]]
  cat("Aridity index available\n")
} else {
  # Fall back to selected-variable stack
  env_aus_sel <- rast(file.path(processed_dir, "env_selected_aus.tif"))
  aridity <- env_aus_sel[["aridity_idx"]]
  cat("Using aridity_idx from selected variable stack\n")
}

# Calibrate aridity threshold using occurrence data
occ_env <- readRDS(file.path(processed_dir, "occurrences_with_env.rds"))
aridity_at_occ <- occ_env$aridity_idx
aridity_q05 <- quantile(aridity_at_occ, 0.05, na.rm = TRUE)
aridity_q95 <- quantile(aridity_at_occ, 0.95, na.rm = TRUE)

cat(sprintf("Aridity index at occurrences: range [%.1f, %.1f], 5th pctl = %.1f, 95th pctl = %.1f\n",
            min(aridity_at_occ, na.rm = TRUE), max(aridity_at_occ, na.rm = TRUE),
            aridity_q05, aridity_q95))

# Determine direction: is higher aridity good or bad?
# Thornthwaite aridity index: higher values = more arid
# Use 95th percentile as upper limit (species avoids very arid conditions)
moisture_ok <- aridity <= as.numeric(aridity_q95)

# Also apply precipitation constraint
if ("bio12" %in% names(env_aus_full)) {
  precip <- env_aus_full[["bio12"]]
  precip_at_occ <- occ_env$bio12
  precip_q05 <- quantile(precip_at_occ, 0.05, na.rm = TRUE)

  # Minimum annual precipitation
  precip_ok <- precip >= as.numeric(precip_q05)
  cat(sprintf("Minimum annual precipitation threshold (5th pctl): %.0f mm\n", precip_q05))
}

vals_moist <- values(moisture_ok, na.rm = TRUE)
vals_precip <- values(precip_ok, na.rm = TRUE)

sprintf("  Aridity constraint: %.1f%% of cells pass",
        100 * sum(vals_moist) / length(vals_moist))
sprintf("  Precipitation constraint: %.1f%% of cells pass",
        100 * sum(vals_precip) / length(vals_precip))
```

## 4. Intersect all constraints

```{r}
#| label: intersect

# Conservative physiological envelope = ALL constraints must pass
climex_suitable <- heat_lethal & cold_dev & warm_activity & moisture_ok & precip_ok
names(climex_suitable) <- "climex_suitable"

# Less conservative = only heat lethal + moisture
climex_broad <- heat_lethal & moisture_ok & precip_ok
names(climex_broad) <- "climex_broad"

# Calculate areas
cell_areas <- cellSize(climex_suitable, unit = "km")
aus_total_area <- sum(values(cell_areas * !is.na(climex_suitable)), na.rm = TRUE)

climex_area <- sum(values(cell_areas * climex_suitable), na.rm = TRUE)
climex_broad_area <- sum(values(cell_areas * climex_broad), na.rm = TRUE)

sprintf("CLIMEX-informed suitable area:")
sprintf("  Conservative (all constraints): %.0f km2 (%.1f%% of Australia)",
        climex_area, 100 * climex_area / aus_total_area)
sprintf("  Broad (heat lethal + moisture): %.0f km2 (%.1f%% of Australia)",
        climex_broad_area, 100 * climex_broad_area / aus_total_area)
```

## 5. Comparison map — CLIMEX vs MaxEnt vs BRT

```{r}
#| label: comparison-map
#| fig-width: 14
#| fig-height: 12

# Create binary maps for comparison
maxent_binary <- pred_maxent >= thresholds$maxent$maxss
brt_binary <- pred_brt >= thresholds$brt$maxss

# Convert all to data frames for plotting
climex_df <- as.data.frame(climex_suitable, xy = TRUE, na.rm = TRUE)
names(climex_df)[3] <- "suitable"
climex_df$model <- "CLIMEX Constraints\n(Physiological Thresholds)"

climex_broad_df <- as.data.frame(climex_broad, xy = TRUE, na.rm = TRUE)
names(climex_broad_df)[3] <- "suitable"
climex_broad_df$model <- "CLIMEX Broad\n(Heat + Moisture Only)"

maxent_df <- as.data.frame(maxent_binary, xy = TRUE, na.rm = TRUE)
names(maxent_df)[3] <- "suitable"
maxent_df$model <- "MaxEnt (Current Variables)"

brt_df <- as.data.frame(brt_binary, xy = TRUE, na.rm = TRUE)
names(brt_df)[3] <- "suitable"
brt_df$model <- "BRT (Current Variables)"

all_maps <- bind_rows(climex_df, climex_broad_df, maxent_df, brt_df)
all_maps$model <- factor(all_maps$model,
                         levels = c("MaxEnt (Current Variables)",
                                    "BRT (Current Variables)",
                                    "CLIMEX Broad\n(Heat + Moisture Only)",
                                    "CLIMEX Constraints\n(Physiological Thresholds)"))

p <- ggplot() +
  geom_raster(data = all_maps, aes(x = x, y = y, fill = factor(as.integer(suitable)))) +
  geom_sf(data = aus, fill = NA, colour = "black", linewidth = 0.3) +
  facet_wrap(~model, ncol = 2) +
  scale_fill_manual(
    values = c("0" = "grey90", "1" = "#D55E00"),
    labels = c("Unsuitable", "Suitable"),
    name = "Suitability"
  ) +
  labs(
    title = expression(italic("D. variabilis") ~ "— Suitability Comparison: SDMs vs CLIMEX Constraints"),
    subtitle = "CLIMEX thresholds from Halliday & Sutherst (1990): Tdev 4C, Topt 28C, Theat 35C, Tlethal 40C",
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    legend.position = "bottom"
  ) +
  coord_sf(xlim = c(112, 155), ylim = c(-45, -10))

print(p)

ggsave(
  file.path(figures_dir, "10c_climex_vs_sdm_comparison.png"),
  p, width = 14, height = 12, dpi = 300
)
```

## 6. Agreement analysis

```{r}
#| label: agreement

# Where do all three approaches agree?
# Resample CLIMEX to match SDM resolution if needed
if (!compareGeom(climex_suitable, maxent_binary, stopOnError = FALSE)) {
  climex_resampled <- resample(climex_suitable, maxent_binary, method = "near")
  climex_broad_resampled <- resample(climex_broad, maxent_binary, method = "near")
} else {
  climex_resampled <- climex_suitable
  climex_broad_resampled <- climex_broad
}

# Agreement layers
agree_all <- maxent_binary & brt_binary & climex_resampled
agree_maxent_climex <- maxent_binary & climex_resampled
agree_brt_climex <- brt_binary & climex_resampled

agree_all_area <- sum(values(cell_areas * agree_all), na.rm = TRUE)
agree_mc_area <- sum(values(cell_areas * agree_maxent_climex), na.rm = TRUE)
agree_bc_area <- sum(values(cell_areas * agree_brt_climex), na.rm = TRUE)
maxent_area <- sum(values(cell_areas * maxent_binary), na.rm = TRUE)
brt_area <- sum(values(cell_areas * brt_binary), na.rm = TRUE)

cat("--- Agreement Analysis ---\n\n")
sprintf("MaxEnt suitable area: %.0f km2 (%.1f%%)", maxent_area, 100 * maxent_area / aus_total_area)
sprintf("BRT suitable area: %.0f km2 (%.1f%%)", brt_area, 100 * brt_area / aus_total_area)
sprintf("CLIMEX suitable area: %.0f km2 (%.1f%%)", climex_area, 100 * climex_area / aus_total_area)
cat("\n")
sprintf("MaxEnt AND CLIMEX agree: %.0f km2 (%.1f%%)", agree_mc_area, 100 * agree_mc_area / aus_total_area)
sprintf("BRT AND CLIMEX agree: %.0f km2 (%.1f%%)", agree_bc_area, 100 * agree_bc_area / aus_total_area)
sprintf("ALL THREE agree: %.0f km2 (%.1f%%)", agree_all_area, 100 * agree_all_area / aus_total_area)

cat("\n--- Interpretation ---\n")
cat("The BRT-CLIMEX agreement area represents the most conservative and defensible\n")
cat("estimate of climatic suitability. This is the area where:\n")
cat("  1. The BRT model (better variable distribution) predicts suitability\n")
cat("  2. CLIMEX physiological thresholds are met\n")
cat("  3. Both independent methods converge\n")
```

## 7. CLIMEX reference locations

Compare CLIMEX Ecoclimatic Index values from the 1990 paper with current
SDM predictions at the same locations.

```{r}
#| label: reference-locations

# Key locations from Halliday & Sutherst (1990)
ref_locations <- tibble(
  city = c("Innisfail", "Lismore", "Sydney", "Melbourne", "Perth",
           "Darwin", "Alice Springs", "Hobart"),
  lon = c(146.03, 153.28, 151.21, 144.96, 115.86,
          130.84, 133.88, 147.33),
  lat = c(-17.53, -28.81, -33.87, -37.81, -31.95,
          -12.46, -23.70, -42.88),
  climex_EI = c(63, 43, NA, NA, NA, NA, NA, NA),  # Published values
  climex_note = c("Highest EI in Australia", "Comparable to Little Rock AR (EI=44)",
                  "Major population centre", "Major population centre",
                  "Dry climate", "Tropical", "Arid interior", "Cool temperate")
)

# Extract SDM predictions at reference locations
ref_pts <- vect(ref_locations, geom = c("lon", "lat"), crs = "EPSG:4326")

if (file.exists(file.path(processed_dir, "pred_aus_maxent.tif"))) {
  ref_locations$maxent_pred <- extract(pred_maxent, ref_pts)[, 2]
  ref_locations$brt_pred <- extract(pred_brt, ref_pts)[, 2]
  ref_locations$climex_pass <- extract(climex_suitable, ref_pts)[, 2]
}

cat("CLIMEX reference locations — SDM predictions:\n")
ref_locations %>%
  mutate(across(c(maxent_pred, brt_pred), ~round(., 3))) %>%
  print(n = Inf)

# Save
write_csv(ref_locations, file.path(processed_dir, "climex_reference_locations.csv"))
```

## 8. Summary

```{r}
#| label: summary

cat("======================================================\n")
cat("  CLIMEX CONSTRAINT MAPPING SUMMARY\n")
cat("======================================================\n\n")

cat("Physiological thresholds applied (Halliday & Sutherst 1990):\n")
cat("  - bio15 (WorldClim bio_5: max temp warmest month) <= 40C (upper lethal)\n")
cat("  - bio2 (WorldClim bio_10: mean temp warmest quarter) >= 4C (development threshold)\n")
cat("  - bio2 (WorldClim bio_10: mean temp warmest quarter) >= 15C (activity threshold)\n")
cat("  - Aridity within species' observed range\n")
cat("  - Annual precipitation >= 5th percentile of occurrence values\n\n")

cat("Area estimates:\n")
sprintf("  CLIMEX physiological: %.0f km2 (%.1f%%)", climex_area, 100 * climex_area / aus_total_area)
sprintf("  MaxEnt (current): %.0f km2 (%.1f%%)", maxent_area, 100 * maxent_area / aus_total_area)
sprintf("  BRT (current): %.0f km2 (%.1f%%)", brt_area, 100 * brt_area / aus_total_area)
sprintf("  BRT + CLIMEX agreement: %.0f km2 (%.1f%%)", agree_bc_area, 100 * agree_bc_area / aus_total_area)

cat("\nThe CLIMEX constraint map serves as an independent, physiologically-informed\n")
cat("baseline. If the BRT prediction aligns with CLIMEX constraints while MaxEnt\n")
cat("does not, this supports the hypothesis that MaxEnt's Bio1 reliance causes\n")
cat("over-prediction in Australia.\n")

# Save constraint rasters
writeRaster(climex_suitable, file.path(processed_dir, "climex_suitable_aus.tif"), overwrite = TRUE)
writeRaster(climex_broad, file.path(processed_dir, "climex_broad_aus.tif"), overwrite = TRUE)

sprintf("Saved CLIMEX constraint outputs")
```
