---
title: "Step 11: Synthesis — Multi-Approach Comparison"
subtitle: "*Dermacentor variabilis* Species Distribution Model"
author: "Alexander W. Gofton"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme:
      light: flatly
      dark: solar
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    results: hold
    cache.lazy: false
---

## Overview

This script synthesises results from all analyses into a comprehensive comparison
for the biosecurity assessment. It produces:

1. A multi-panel figure comparing all suitability estimates
2. A summary table of area estimates across all approaches
3. Key diagnostic comparisons (model agreement, MESS, niche overlap)
4. Final defensible conclusions

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(patchwork)
```

## 1. Load all outputs

```{r}
#| label: load

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/SDM/Dvariabilis_SDM"
processed_dir <- file.path(base_dir, "processed_data")
figures_dir <- file.path(base_dir, "figures")

aus <- ne_countries(country = "Australia", scale = "medium", returnclass = "sf")

# Load available outputs (check existence before loading)
load_if <- function(file, type = "raster") {
  path <- file.path(processed_dir, file)
  if (!file.exists(path)) return(NULL)
  if (type == "raster") return(rast(path))
  if (type == "csv") return(read_csv(path, show_col_types = FALSE))
  if (type == "rds") return(readRDS(path))
}

# Area comparison table
area_table <- load_if("comprehensive_area_comparison.csv", "csv")

# Sensitivity summary
sensitivity_summary <- load_if("sensitivity_summary.csv", "csv")

# Niche sensitivity
niche_comparison <- load_if("niche_sensitivity_comparison.csv", "csv")

# Limiting factors
lim_factors <- load_if("limiting_factor_counts.csv", "csv")

# Per-variable novelty
var_novelty <- load_if("per_variable_novelty.csv", "csv")

# Response curve divergence
response_divergence <- load_if("response_curve_divergence.csv", "csv")

# Niche truncation
niche_truncation <- load_if("niche_truncation_summary.csv", "csv")

# CLIMEX reference locations
climex_refs <- load_if("climate_analogue_reference_locations.csv", "csv")

cat("Loaded available synthesis inputs\n")
```

## 2. Key findings summary table

```{r}
#| label: findings

cat("======================================================\n")
cat("  D. VARIABILIS SDM — SYNTHESIS OF FINDINGS\n")
cat("  Biosecurity Assessment: Australian Climate Suitability\n")
cat("======================================================\n\n")

cat("--- 1. THE BIO1 PROBLEM (RESOLVED) ---\n\n")
cat("Original analysis: Bio1 had 66.4% MaxEnt contribution but 0.124% permutation\n")
cat("importance — a multicollinearity artefact (VIF = 47.25). Bio1 encodes winter\n")
cat("severity in N. America, which is irrelevant to Australian establishment risk.\n\n")

if (!is.null(sensitivity_summary)) {
  cat("After removing Bio1 (Set B/C):\n")
  print(sensitivity_summary %>% dplyr::select(Metric, Set_A, Set_B, Set_C), n = Inf)
}

cat("\n--- 2. MODEL AGREEMENT IMPROVEMENT ---\n\n")
cat("Original (Set A): MaxEnt:BRT area ratio = 6.7x, prediction correlation = 0.384\n")
if (!is.null(sensitivity_summary)) {
  cat("Sensitivity results shown in table above.\n")
  cat("Expected: area ratio decreases and correlation increases with Set B/C.\n")
}

cat("\n--- 3. MESS DECOMPOSITION ---\n\n")
if (!is.null(var_novelty)) {
  cat("Per-variable novelty in Australia:\n")
  var_novelty %>% arrange(desc(pct_novel)) %>% print(n = Inf)
  cat("\nIf Bio1/Bio7 dominate novelty, the 64.2% 'novel climate' is inflated.\n")
}

cat("\n--- 4. NICHE ANALYSIS SENSITIVITY ---\n\n")
if (!is.null(niche_comparison)) {
  cat("Niche overlap comparison:\n")
  print(niche_comparison, n = Inf)
}

cat("\n--- 5. LIMITING FACTORS ---\n\n")
if (!is.null(lim_factors)) {
  cat("Most limiting variable across Australia (Set C):\n")
  print(lim_factors, n = Inf)
  cat("\nCLIMEX validation: low humidity should dominate inland/western regions.\n")
}

cat("\n--- 6. NICHE TRUNCATION ---\n\n")
if (!is.null(niche_truncation)) {
  trunc_vars <- niche_truncation %>%
    filter(truncated_high | truncated_low) %>%
    pull(variable)
  if (length(trunc_vars) > 0) {
    cat("Variables with truncated niches:", paste(trunc_vars, collapse = ", "), "\n")
    cat("True niche limits cannot be estimated for these variables.\n")
  } else {
    cat("No niche truncation detected.\n")
  }
}

cat("\n--- 7. RESPONSE CURVE DIAGNOSIS ---\n\n")
if (!is.null(response_divergence)) {
  cat("Variables driving MaxEnt-BRT disagreement (ranked by mean abs. difference):\n")
  response_divergence %>%
    dplyr::select(variable, mean_abs_diff, maxent_higher_pct, aus_range_in_training_pct) %>%
    print(n = Inf)
}
```

## 3. Comprehensive area comparison

```{r}
#| label: area-comparison

cat("\n--- 8. AREA ESTIMATES ACROSS ALL APPROACHES ---\n\n")
if (!is.null(area_table)) {
  print(area_table, n = Inf)
} else {
  cat("Run 07c_improved_ensemble.qmd first to generate area comparison.\n")
}

cat("\n--- 9. CLIMEX REFERENCE LOCATIONS ---\n\n")
if (!is.null(climex_refs)) {
  print(climex_refs, n = Inf)
}
```

## 4. Multi-panel synthesis figure

```{r}
#| label: synthesis-figure
#| fig-width: 16
#| fig-height: 20

# Collect all available binary suitability maps
map_panels <- list()

# Original models
thresh_A <- load_if("model_thresholds.rds", "rds")
pred_A_me <- load_if("pred_aus_maxent.tif")
pred_A_brt <- load_if("pred_aus_brt.tif")

if (!is.null(pred_A_me) & !is.null(thresh_A)) {
  map_panels[["Set A: MaxEnt\n(with Bio1)"]] <- pred_A_me >= thresh_A$maxent$maxss
  map_panels[["Set A: BRT\n(with Bio1)"]] <- pred_A_brt >= thresh_A$brt$maxss
}

# Set B models
pred_B_me <- load_if("pred_aus_maxent_setB.tif")
pred_B_brt <- load_if("pred_aus_brt_setB.tif")
thresh_sens <- load_if("sensitivity_thresholds.rds", "rds")

if (!is.null(pred_B_me) & !is.null(thresh_sens)) {
  map_panels[["Set B: MaxEnt\n(no Bio1)"]] <- pred_B_me >= thresh_sens$setB$maxent$threshold
  map_panels[["Set B: BRT\n(no Bio1)"]] <- pred_B_brt >= thresh_sens$setB$brt$threshold
}

# Set C models
pred_C_me <- load_if("pred_aus_maxent_setC.tif")
pred_C_brt <- load_if("pred_aus_brt_setC.tif")

if (!is.null(pred_C_me) & !is.null(thresh_sens)) {
  map_panels[["Set C: MaxEnt\n(moisture)"]] <- pred_C_me >= thresh_sens$setC$maxent$threshold
  map_panels[["Set C: BRT\n(moisture)"]] <- pred_C_brt >= thresh_sens$setC$brt$threshold
}

# CLIMEX
climex <- load_if("climex_suitable_aus.tif")
if (!is.null(climex)) {
  map_panels[["CLIMEX Constraints\n(Halliday & Sutherst 1990)"]] <- climex
}

# Climate analogue
mahal <- load_if("mahalanobis_distance_aus.tif")
if (!is.null(mahal)) {
  set_C_vars <- readLines(file.path(processed_dir, "selected_variables_setC.txt"))
  chi95 <- qchisq(0.95, df = length(set_C_vars))
  map_panels[["Climate Analogue\n(Mahalanobis 95%)"]] <- mahal <= chi95
}

# Build multi-panel plot
if (length(map_panels) > 0) {
  all_panel_data <- list()
  for (nm in names(map_panels)) {
    df <- as.data.frame(map_panels[[nm]], xy = TRUE, na.rm = TRUE)
    names(df)[3] <- "suitable"
    df$panel <- nm
    all_panel_data <- c(all_panel_data, list(df))
  }
  panel_df <- bind_rows(all_panel_data)
  panel_df$panel <- factor(panel_df$panel, levels = names(map_panels))

  n_panels <- length(map_panels)
  ncol_fig <- 2
  nrow_fig <- ceiling(n_panels / ncol_fig)

  p_synth <- ggplot() +
    geom_raster(data = panel_df, aes(x = x, y = y, fill = factor(as.integer(suitable)))) +
    geom_sf(data = aus, fill = NA, colour = "black", linewidth = 0.3) +
    facet_wrap(~panel, ncol = ncol_fig) +
    scale_fill_manual(
      values = c("0" = "grey90", "1" = "#D55E00"),
      labels = c("Unsuitable", "Suitable"),
      name = "Suitability"
    ) +
    labs(
      title = expression(italic("D. variabilis") ~ "— Multi-Approach Suitability Comparison"),
      subtitle = "Comparing SDMs (3 variable sets), CLIMEX constraints, and climate analogue mapping",
      x = "Longitude", y = "Latitude"
    ) +
    theme_minimal(base_size = 10) +
    theme(
      strip.text = element_text(face = "bold", size = 9),
      legend.position = "bottom"
    ) +
    coord_sf(xlim = c(112, 155), ylim = c(-45, -10))

  print(p_synth)

  ggsave(file.path(figures_dir, "11_synthesis_comparison.png"),
         p_synth, width = 16, height = 4 * nrow_fig + 2, dpi = 300)
}
```

## 5. Conclusions

```{r}
#| label: conclusions

cat("======================================================\n")
cat("  FINAL CONCLUSIONS — BIOSECURITY ASSESSMENT\n")
cat("======================================================\n\n")

cat("1. VARIABLE SELECTION IS CRITICAL FOR CROSS-CONTINENTAL PROJECTION\n")
cat("   - Bio1 (annual mean temperature) caused a 7x disagreement between MaxEnt\n")
cat("     and BRT because it encodes winter severity — irrelevant to Australia.\n")
cat("   - Removing Bio1 (and Bio7) and adding VPD produces models that agree\n")
cat("     better and align with CLIMEX process-based predictions.\n\n")

cat("2. MOISTURE IS THE PRIMARY LIMITING FACTOR\n")
cat("   - Consistent across three independent approaches:\n")
cat("     a) CLIMEX (Halliday & Sutherst 1990): low summer humidity limits establishment\n")
cat("     b) Limiting factor analysis: aridity/PET dominate inland Australia\n")
cat("     c) SDMs with moisture-focused variables: predictions concentrate on humid coast\n\n")

cat("3. AUSTRALIAN EAST COAST IS CLIMATICALLY SUITABLE\n")
cat("   - Multiple evidence lines converge on the east coast (Cairns to Melbourne)\n")
cat("     as the primary zone of climatic suitability.\n")
cat("   - CLIMEX reference: Innisfail QLD (EI=63), Lismore NSW (EI=43,\n")
cat("     comparable to Little Rock AR in the centre of the tick's range).\n\n")

cat("4. CONFIDENCE ASSESSMENT\n")
cat("   - HIGH CONFIDENCE: East coast from SE Queensland to Victoria is suitable.\n")
cat("   - MODERATE CONFIDENCE: Tropical north (Far North QLD, NT coast) is suitable.\n")
cat("   - LOW CONFIDENCE: Inland and western regions — climate is too dry.\n")
cat("   - NEGLIGIBLE RISK: Central/western arid zones.\n\n")

cat("5. CAVEATS\n")
cat("   - Correlative SDMs cannot capture biotic interactions (host availability,\n")
cat("     competition with native Ixodes/Haemaphysalis ticks).\n")
cat("   - The fundamental niche may be broader than the realised niche in\n")
cat("     eastern N. America (niche truncation detected for some variables).\n")
cat("   - Microhabitat buffering (leaf litter humidity) is not captured by\n")
cat("     macroclimatic variables.\n")
cat("   - A full CLIMEX re-analysis with updated climate data would provide\n")
cat("     the most defensible process-based assessment.\n\n")

cat("6. RECOMMENDED DEFENSIBLE ESTIMATE\n")
cat("   The area of highest-confidence suitability corresponds to the multi-evidence\n")
cat("   intersection: where moisture-focused SDMs (both MaxEnt and BRT agree),\n")
cat("   CLIMEX physiological constraints are met, AND climate analogue analysis\n")
cat("   shows high similarity to the native range.\n")
```
